---
title: "The Natural History and Dynamics of Clonal Haematopoiesis"
subtitle: "Phylogeny and phylodynamics in Mitchell et al. (2021)"
author: 
    - José Guilherme de Almeida
    - Moritz Gerstung
output: 
  html_document:
    theme: lumen
    highlight: tango
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# setup

```{r,echo=F,collapse=T}
tm <- 0.352777778
pop_size <- 1e5
source("Scripts/vaf_dynamics_functions.R")
library(phylodyn)
library(parallel)
INLA:::inla.dynload.workaround()
set.seed(42)
source("Scripts/prepare_data.R")

change_point <- function(x, b0, m1, m2, delta) { 
  b0 + (x*m1) + (sapply(x-delta, function (t) max(0, t)) * m2)
}

expand_edge <- function(tree,edge_id) {
  # takes an edge_id and retrieves the tips which are common to this edge
  n_tips <- length(tree$tip.label)
  edge_ids <- c(edge_id)
  done <- F
  while (done == F) {
    new_edge <- tree$edge[,2][tree$edge[,1] %in% edge_id]
    edge_ids <- c(edge_ids,new_edge)
    edge_id <- new_edge
    if (length(edge_id) == 0) {
      done <- T
    }
  }
  edge_ids <- unique(edge_ids)
  edge_ids <- edge_ids[edge_ids <= n_tips]
  return(edge_ids)
}

trim_tree <- function(tree,l) {
  n_tips <- length(tree$tip.label)
  tree$edge.length <- ifelse(
    tree$edge[,2] <= n_tips,
    tree$edge.length - l,
    tree$edge.length
  )
  tree <- drop.tip(
    tree,
    tree$edge[tree$edge.length < 0] %>% sapply(
      function(x) expand_edge(tree,x)) %>% unlist())
  return(tree)
}

expanded_clade_reader <- function(path) {
  # convenience function to read expanded clade csv
  read.csv(path) %>%
    mutate(HasDriver = !grepl("Clade",variant_ID)) %>% 
    mutate(CladeNo = as.numeric(ifelse(!HasDriver,str_match(variant_ID,"[0-9]+"),NA))) %>% 
    arrange(HasDriver,CladeNo) %>%
    mutate(variant_ID = gsub(" p.","-",variant_ID)) %>% 
    mutate(variant_ID = gsub("_p.","-",variant_ID)) %>% 
    mutate(variant_ID = gsub("_","-",variant_ID)) %>% 
    mutate(variant_ID = gsub(" ","",variant_ID)) %>% 
    return
}

bnpr_at_coalescence <- function(bnpr_estimate) {
  time_at_coal <- bnpr_estimate$coal_times
  X <- bnpr_estimate$summary$time
  Y <- bnpr_estimate$summary$quant0.5
  Y025 <- bnpr_estimate$summary$quant0.025
  Y975 <- bnpr_estimate$summary$quant0.975
  Y_log <- log(na.omit(approx(X,Y,xout = time_at_coal,rule = 2)$y))
  values_at_coalescence <- data.frame(
    X = time_at_coal,
    Y = Y_log,
    Y_exp = approx(X,Y,xout = time_at_coal)$y,
    Y025 = approx(X,Y025,xout = time_at_coal)$y,
    Y975 = approx(X,Y975,xout = time_at_coal)$y
  ) %>%
    na.omit()
  return(values_at_coalescence)
}

bnpr_at_all <- function(bnpr_estimate) {
  X <- bnpr_estimate$summary$time
  Y <- bnpr_estimate$summary$quant0.5
  Y025 <- bnpr_estimate$summary$quant0.025
  Y975 <- bnpr_estimate$summary$quant0.975
  Y_log <- log(Y)
  values_at_coalescence <- data.frame(
    X = X,
    Y = Y_log,
    Y_exp = Y,
    Y025 = Y025,
    Y975 = Y975
  ) %>%
    na.omit()
  return(values_at_coalescence)
}

colours <- c(
  RColorBrewer::brewer.pal(9, "Set1"),
  RColorBrewer::brewer.pal(8, "Set2"),
  RColorBrewer::brewer.pal(8, "Set3"))

ages <- list(KX003 = 81,KX004 = 77,
             KX007 = 75,KX008 = 76)

trees <- list(
  KX003 = read.tree("data/Trees_emily/KX003_5_ultra.tree"),
  KX004 = read.tree("data/Trees_emily/KX004_4_ultra.tree"),
  KX007 = read.tree("data/Trees_emily/KX007_2_ultra.tree"),
  KX008 = read.tree("data/Trees_emily/KX008_2_ultra.tree")
)

trees$KX003$edge.length <- trees$KX003$edge.length * ages$KX003
trees$KX004$edge.length <- trees$KX004$edge.length * ages$KX004
trees$KX007$edge.length <- trees$KX007$edge.length * ages$KX007
trees$KX008$edge.length <- trees$KX008$edge.length * ages$KX008

clade_identification <- list(
  KX003 = expanded_clade_reader("data/Trees_emily/KX003_5_expanded_clades.csv"),
  KX004 = expanded_clade_reader("data/Trees_emily/KX004_4_expanded_clades.csv"),
  KX007 = expanded_clade_reader("data/Trees_emily/KX007_2_expanded_clades.csv"),
  KX008 = expanded_clade_reader("data/Trees_emily/KX008_2_expanded_clades.csv")
)
```

# bnpr estimates

```{r,echo=F,collapse=T}
file_name <- "data_output/michell_bnpr_estimates.R"
if (file.exists(file_name) == T) {
  tmp_obj <- readRDS(file_name)
  clades <- tmp_obj[[1]]
  clade_ids <- tmp_obj[[2]]
  bnpr_estimates <- tmp_obj[[3]]
  bnpr_estimates_trimmed <- tmp_obj[[4]]
  bnpr_estimates_extended <- tmp_obj[[5]]
  individual_bnpr_estimates <- tmp_obj[[6]]
  individual_bnpr_estimates_trimmed <- tmp_obj[[7]]
} else {
  clades <- list()
  clade_ids <- list()
  bnpr_estimates <- list()
  bnpr_estimates_trimmed <- list()
  bnpr_estimates_extended <- list()
  individual_bnpr_estimates <- list()
  individual_bnpr_estimates_trimmed <- list()
  
  for (x in c("KX003","KX004","KX007","KX008")) {
    tr <- trees[[x]]
    tr$edge.length <- tr$edge.length
    bnpr_estimates[[x]] <- list()
    clades[[x]] <- clade_identification[[x]]$node %>%
      lapply(expand_edge,tree = tr)
    clade_ids[[x]] <- clade_identification[[x]]$variant_ID
    color_vector <- rep("black",length(tr$tip.label)) 
    for (i in 1:length(clades[[x]])) {
      clade <- clades[[x]][[i]]
      color_vector[clade] <- colours[i]
    }
    tree_groups <- groupOTU(tr,clades[[x]])
    ggtree(tree_groups,
           aes(colour = group)) +
      scale_color_manual(breaks = c(0:length(clades[[x]])),
                         values = c("black",colours[1:length(clades[[x]])]))
    X <- mclapply(list(tr,trim_tree(tr,5)),BNPR,mc.cores = 2)
    individual_bnpr_estimates[[x]] <- X[[1]]
    # bnpr for tree trimmed by 5 years
    individual_bnpr_estimates_trimmed[[x]] <- X[[2]]
    for (i in 1:length(clades[[x]])) {
      print(c(x,i))
      clade <- clades[[x]][[i]]
      tree_subset <- keep.tip(tr,clade)
      tree_subset <- multi2di(tree_subset)
      X <- mclapply(list(tree_subset,
                         trim_tree(tree_subset,5),
                         trim_tree(tree_subset,-10)),
                    BNPR,mc.cores = 2)
      bnpr_estimates[[x]][[i]] <- X[[1]]
      # bnpr for tree subsets trimmed by 5 years
      bnpr_estimates_trimmed[[x]][[i]] <- X[[2]]
      bnpr_estimates_extended[[x]][[i]] <- X[[3]]
    }
  }
  saveRDS(
    list(clades,clade_ids,bnpr_estimates,bnpr_estimates_trimmed,bnpr_estimates_extended,
         individual_bnpr_estimates,individual_bnpr_estimates_trimmed),
    file_name)
}
```

# plotting estimates

## fitting models and creating plots

```{r,echo=F,collapse=T}
all_plots <- list()
all_growth_estimates <- list()
par(mfrow = c(4,5),mar = c(1,1,1,1))
for (x in names(clades)) {
  all_estimates <- list()
  all_growth_estimates[[x]] <- list()
  max_pop <- max(individual_bnpr_estimates[[x]]$summary$quant0.975) 
  tr <- trees[[x]]
  total_size <- length(tr$tip.label)
  for (i in 1:length(clades[[x]])) {
    clade_size <- clade_identification[[x]]$number_samples[i]
    vaf_from_clade <- clade_size / total_size
    vaf_from_clade_005 <- qbeta(0.05,clade_size,total_size - clade_size)
    vaf_from_clade_095 <- qbeta(0.95,clade_size,total_size - clade_size)
    dff <- bnpr_estimates[[x]][[i]]$summary
    dff$clade <- i
    all_estimates[[i]] <- dff
    X <- (ages[[x]] - dff$time)
    Y <- log(dff$quant0.5)
    not_valid <- is.infinite(Y) | is.na(Y)
    Y <- Y[!not_valid]
    X <- X[!not_valid]
    Y_exp <- exp(Y)
    Y975 <- dff$quant0.975[!not_valid]
    Y025 <- dff$quant0.025[!not_valid]
    W <- (log(Y975) - log(Y025))^2/16
    # M fixes some problems regarding initialization for the nonlinear fit
    M <- 0
    M <- ifelse(x == "KX007" & i == 16,10,M)
    Y_exp_vaf <- Y_exp / Y_exp[1] * vaf_from_clade
    nonlinear <- nls(Y_exp ~ SSlogis(X, Asym, b2, b3),
                     start = list(Asym = max(Y_exp) + 100,b2 = max(X),b3 = 10),
                     control = nls.control(maxiter = 1000,warnOnly = T),
                     data = data.frame(X, Y_exp),
                     upper = list(Asym = max_pop,b2 = max(X),b3 = 1000),
                     algorithm = "port",
                     weights = 1/W)
    nonlinear_vaf <- nls(Y_exp_vaf ~ SSlogis(X, 1, b2, b3),
                         start = list(b2 = max(X),b3 = 10),
                         control = nls.control(maxiter = 1000,warnOnly = T),
                         data = data.frame(X, Y_exp_vaf),
                         upper = list(Asym = max_pop,b2 = max(X),b3 = 1000),
                         algorithm = "port",
                         weights = 1/W)

    linear <- lm(Y ~ X,weights = 1/W)
    opt_fn <- function(par) {
      se <- (Y - change_point(X,par[1],par[2],par[3],par[4]))^2
      return(mean(se / W))
    }
    change_point_regression <- optim(
      par = c(linear$coefficients[1],
              linear$coefficients[2],
              0,
              mean(X)),
      method = "L-BFGS-B",
      lower = c(NA,0,NA,min(X) + diff(range(X)) * 0.25),
      upper = c(NA,NA,NA,max(X) - diff(range(X)) * 0.25),
      fn = opt_fn)
    cprp <- change_point_regression$par
    cp_y <- change_point(X,cprp[1],cprp[2],cprp[3],cprp[4])
    cp_early <- change_point(X,cprp[1],cprp[2],0,cprp[4])

    b <- -log(200e3 - 1)
    expected_vaf <- inv.logit(change_point_regression$par[2] * diff(range(X)) + b)
    
    if (length(X) > 50) {
      linear_early <- lm(Y ~ X,data = data.frame(X, Y)[81:100,],weights = 1/W[81:100])
      linear_late <- lm(Y ~ X,data = data.frame(X, Y)[1:20,],weights = 1/W[1:20])
    } else {
      linear_early <- list(coefficients=NA)
      linear_late <- list(coefficients=NA)
    }
    
    pred_nl <- predict(nonlinear)
    pred_nl_vaf <- predict(nonlinear_vaf)/vaf_from_clade * Y_exp[1]
    pred_l <- predict(linear)
    n_samples <- clade_identification[[x]]$number_samples[i]
    clade_id <- clade_identification[[x]]$variant_ID[i]
    all_estimates[[i]]$n_samples <- n_samples
    all_estimates[[i]]$clade_id <- clade_id
    all_growth_estimates[[x]][[i]] <- list(
      linear = linear,
      nonlinear = nonlinear,
      linear_early = linear_early,
      linear_late = linear_late,
      change_point_regression = change_point_regression,
      nonlinear_vaf = nonlinear_vaf,
      vaf_from_clade = vaf_from_clade,
      vaf_from_clade_005 = vaf_from_clade_005,
      vaf_from_clade_095 = vaf_from_clade_095,
      expected_vaf = expected_vaf,
      clade_id = clade_id,
      data = data.frame(
        X,Y,Y025,Y975,Y_exp,pred_l,pred_nl,pred_nl_vaf,W,pred_cp = cp_y,
        pred_cp_early = cp_early,
        clade = i,clade_id = clade_id,id = x,n_samples = n_samples))
  }
  all_estimates <- do.call(rbind,all_estimates)
  # plotting the tree
  tree_groups <- groupOTU(trees[[x]],clades[[x]])
  mrca <- c()
  mrca_names <- c()
  for (C in 1:length(clades[[x]])) {
    clade <- clades[[x]][[C]]
    mrca[C] <- MRCA(trees[[x]],clade)
    mrca_names[C] <- C
  }
  edge_labels <- data.frame(trees[[x]]$edge)
  edge_labels$LL <- trees[[x]]$edge %>%
    apply(1,function(x) {
      ix <- which(mrca %in% x[2])
      if (length(ix) > 0) {
        return(mrca_names[ix])
      } else {
        return(NA)
      }
    }) 
  colnames(edge_labels) <- c("parent", "node", "edge_num")
  edge_labels$edge_num <- clade_identification[[x]]$variant_ID[edge_labels$edge_num]
  tree_plot <- ggtree(tree_groups,aes(colour = group,size = group == 0)) +
    scale_color_manual(breaks = c(0:length(clades[[x]])),
                       values = c("grey80",colours[1:length(clades[[x]])]),
                       guide = F) + 
    scale_size_discrete(breaks = c(F,T),range = c(0.25,0.1),guide = F) +
    ggtitle(x) + 
    geom_text(aes(label = 1)) + 
    theme_gerstung(base_size = 6) + 
    theme(axis.line.y = element_blank(),axis.ticks.y = element_blank(),axis.text.y = element_blank()) + 
    scale_x_continuous(breaks = c(0,1),labels = c(0,ages[[x]]),expand = c(0,0))
  tree_plot <- tree_plot %<+% edge_labels + 
    geom_label_repel(aes(x=branch, label=edge_num),
                     alpha = 0.8,
                     size = 2.2,
                     vjust = 0.6,
                     min.segment.length = 0.01,
                     segment.size = 0.25,
                     label.padding = unit(0,"cm"),
                     label.r = unit(0,"cm"),
                     label.size = unit(0,"cm"),
                     colour = "black")
  
  # plotting the trajectories
  bnpr_plot <- all_estimates %>%
    subset(n_samples >= 5) %>% 
    mutate(clade_id = factor(clade_id,levels = clade_identification[[x]]$variant_ID)) %>%
    ggplot(aes(x = (1-time),y = quant0.5,
               color = as.factor(clade),
               fill = as.factor(clade))) + 
    geom_ribbon(aes(ymin = quant0.025,ymax = quant0.975),color = NA,alpha = 0.1) +
    geom_line(size = 0.25) + 
    scale_y_continuous(trans = 'log10') + 
    theme_gerstung(base_size = 6) + 
    ylab("Effective population size") + 
    xlab("Age") +
    facet_wrap(~ clade_id) +
    scale_color_manual(breaks = c(0:length(colours)),
                       values = c("black",colours),
                       guide = F) +
    scale_fill_manual(breaks = c(0:length(colours)),
                      values = c("black",colours),
                      guide = F) + 
    theme(strip.text = element_text(margin = margin(0)),
          strip.placement = "inside") + 
    coord_cartesian(xlim = c(0,ages[[x]]),
                    ylim = c(NA,1e6)) + 
    scale_x_continuous(breaks = c(0,30,60,90))
  all_plots[[x]] <- list(
    tree = tree_plot,
    bnpr = bnpr_plot
  )
}
```

```{r}
all_growth_estimates_df <- do.call(
  rbind,
  lapply(
    c("KX003","KX004","KX007","KX008"),
    function(x) do.call(rbind,lapply(all_growth_estimates[[x]],function(x) x$data))
  )
)

final_pop_size <- all_growth_estimates_df %>%
  group_by(clade,id,clade_id) %>%
  summarise(N = Y[which.max(X)],N025 = log(Y025[which.max(X)]),N975 = log(Y975[which.max(X)]),
            n_samples = n_samples[1]) %>%
  as.data.frame %>% 
  rbind(
    data.frame(
      clade = rep(NA,4),
      id = c("KX003","KX004","KX007","KX008"),
      clade_id = rep("Total",4),
      N = sapply(individual_bnpr_estimates,function(x) log(x$summary$quant0.5[1])),
      N025 = sapply(individual_bnpr_estimates,function(x) log(x$summary$quant0.025[1])),
      N975 = sapply(individual_bnpr_estimates,function(x) log(x$summary$quant0.975[1])),
      n_samples = sapply(trees,function(x) length(x$tip.label))
    )
  )

all_growth_estimates_values_df <- lapply(
  c("KX003","KX004","KX007","KX008"),
  function(x) data.frame(
    linear = unlist(lapply(all_growth_estimates[[x]],
                           function(x) x$linear$coefficients[2])),
    linear_se = unlist(lapply(all_growth_estimates[[x]],
                              function(x) summary(x$linear)$coefficients[2,2])),
    nonlinear = unlist(lapply(all_growth_estimates[[x]],
                              function(x) 1/x$nonlinear$m$getPars()[3])),
    nonlinear_vaf = unlist(lapply(all_growth_estimates[[x]],
                                  function(x) 1/x$nonlinear_vaf$m$getPars()[2])),
    linear_early = unlist(lapply(all_growth_estimates[[x]],
                                 function(x) x$linear_early$coefficients[2])),
    linear_late = unlist(lapply(all_growth_estimates[[x]],
                                function(x) x$linear_late$coefficients[2])),
    offset = unlist(lapply(all_growth_estimates[[x]],
                           function(x) x$change_point_regression$par[1])),
    cp_1 = unlist(lapply(all_growth_estimates[[x]],
                         function(x) x$change_point_regression$par[2])),
    cp_2 = unlist(
      lapply(
        all_growth_estimates[[x]],
        function(x) sum(x$change_point_regression$par[c(3,2)]))),
    midpoint = unlist(lapply(all_growth_estimates[[x]],
                             function(x) x$change_point_regression$par[4])),
    id = x,
    vaf_from_clade = unlist(lapply(all_growth_estimates[[x]],
                                   function(x) x$vaf_from_clade)),
    vaf_from_clade_005 = unlist(lapply(all_growth_estimates[[x]],
                                       function(x) x$vaf_from_clade_005)),
    vaf_from_clade_095 = unlist(lapply(all_growth_estimates[[x]],
                                       function(x) x$vaf_from_clade_095)),
    expected_vaf = unlist(lapply(all_growth_estimates[[x]],
                                 function(x) x$expected_vaf)),
    min_age = unlist(lapply(all_growth_estimates[[x]],
                            function(x) min(x$data$X))),
    max_age = unlist(lapply(all_growth_estimates[[x]],
                            function(x) max(x$data$X))),
    w_sum = unlist(lapply(all_growth_estimates[[x]],
                          function(x) mean(x$data$W))),
    clade = unlist(lapply(all_growth_estimates[[x]],
                          function(x) x$data$clade_id[1])),
    clade_no = unlist(lapply(all_growth_estimates[[x]],
                             function(x) x$data$clade[1])),
    N_samples = clade_identification[[x]]$number_samples[
      c(1:length(all_growth_estimates[[x]]))]
  )
) %>% 
  do.call(what = rbind)

size_estimates_df <- lapply(
  c("KX003","KX004","KX007","KX008"),
  function(x) data.frame(
    no_cap = unlist(lapply(all_growth_estimates[[x]],function(x) x$nonlinear$m$getPars()[1])),
    id = x,
    clade = clade_identification[[x]]$variant_ID[c(1:length(all_growth_estimates[[x]]))],
    N_samples = clade_identification[[x]]$number_samples[c(1:length(all_growth_estimates[[x]]))]
  )
) %>% 
  do.call(what = rbind)
```

## Population size estimates at last timepoint

```{r,echo=F,collapse=T}
final_pop_size %>%
  subset(n_samples >= 5) %>%
  ggplot(aes(x = clade_id,y = exp(N),
             ymin = exp(N025),ymax = exp(N975),
             color = ifelse(clade_id == "Total","Total","NotTotal"))) + 
  geom_point() + 
  geom_linerange() + 
  facet_wrap(~ id,scales = "free") + 
  coord_cartesian(ylim = c(NA,1e10)) + 
  xlab("Clade") + 
  ylab("EPS for final timepoint") + 
  theme_gerstung(base_size = 6) + 
  rotate_x_text(angle = 30) + 
  scale_y_continuous(trans = 'log10') + 
  scale_color_manual(values = c(Total = "grey",NotTotal = "black"),guide = F) 

final_pop_size %>%
  subset(n_samples >= 5) %>%
  ggplot(aes(x = n_samples,y = exp(N),
             ymin = exp(N025),ymax = exp(N975),
             color = ifelse(clade_id == "Total","Total","NotTotal"))) + 
  geom_point() + 
  geom_linerange() + 
  xlab("Number of samples in clade") + 
  ylab("EPS for final timepoint") + 
  theme_gerstung(base_size = 6) + 
  rotate_x_text(angle = 30) + 
  scale_x_continuous(trans = 'log10') +
  scale_y_continuous(trans = 'log10') + 
  coord_cartesian(ylim = c(NA,1e20)) + 
  scale_color_manual(values = c(Total = "grey",NotTotal = "black"),guide = F) 
```

## plotting coefficients for linear and sigmoidal fits

```{r,fig.height=2.5,fig.width=6}
all_growth_estimates_values_df %>%
  group_by(id) %>%
  mutate(clade = factor(clade,levels = clade_identification[[id[1]]]$variant_ID)) %>% 
  mutate(clade_N = seq(1,length(clade))) %>%
  mutate(Color = grepl("DNMT3A",clade)) %>% 
  subset(N_samples >= 5) %>% 
  ggplot(aes(x = clade,y = linear,
             ymin = linear - linear_se,ymax = linear + linear_se,
             colour = Color)) + 
  geom_hline(yintercept = 0.1,size = 0.25) +
  geom_hline(yintercept = 0.3,size = 0.25) +
  geom_linerange(size = 0.25) +
  geom_point(size = 0.25,shape = 3) + 
  facet_grid(~ id,scales = "free_x",space = "free_x") + 
  scale_color_manual(breaks = c(F,T),
                     values = colours[1:2],
                     guide = F) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(ylim = c(0,NA)) +
  ylab("Annual growth") + 
  xlab("Clade") + 
  rotate_x_text(angle = 45) + 
  ggtitle("Plotting coefficients for linear fit") +
  ggsave("figures/model_ch/trees/mitchell_coefficients_linear.pdf",
         height = 2,
         width = 4)

all_growth_estimates_values_df %>%
  group_by(id) %>%
  mutate(clade = factor(clade,levels = clade_identification[[id[1]]]$variant_ID)) %>% 
  mutate(clade_N = seq(1,length(clade))) %>%
  mutate(Color = grepl("DNMT3A",clade)) %>% 
  subset(N_samples >= 5) %>% 
  ggplot(aes(x = clade,y = nonlinear,
             colour = as.factor(Color))) + 
  geom_hline(yintercept = 0.1,size = 0.25) +
  geom_hline(yintercept = 0.3,size = 0.25) +
  geom_point(size = 0.25,shape = 3) + 
  facet_grid(~ id,scales = "free_x",space = "free_x") + 
  scale_color_manual(breaks = c(F,T),
                     values = colours[1:2],
                     guide = F) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(ylim = c(0,NA)) +
  ylab("Annual growth") + 
  xlab("Clade") + 
  rotate_x_text(angle = 45) + 
  ggtitle("Plotting coefficients for sigmoidal fit") +
  ggsave("figures/model_ch/trees/mitchell_coefficients_nonlinear.pdf",
         height = 2,
         width = 4)
```

## Growth rates when using early and late coefficients

```{r,fig.height=2,fig.width=2}
all_growth_estimates_values_df %>%
  subset(N_samples >= 5) %>% 
  gather(key = "key",value = "value",cp_1,cp_2) %>%
  mutate(key = ifelse(key == "cp_1","Early","Late")) %>% 
  ggplot(aes(x = key,y = exp(value) - 1)) + 
  geom_point(size = 0.5) +
  geom_line(aes(group = paste(id,clade)),alpha = 0.4,size = 0.25) + 
  geom_boxplot(size = 0.25,outlier.colour = NA) + 
  theme_gerstung(base_size = 6) + 
  coord_cartesian(ylim = c(0,1.1)) + 
  ylab("Annual growth rate") + 
  xlab("") + 
  ggtitle("Comparing growth rates estimates when\nconsidering early and late growth rates") +
  ggsave("figures/model_ch/trees/mitchell_coefficients_early_late.pdf",
         height = 2.5,
         width = 2)
```

## Trajectories for different types of fit 

```{r,fig.height=2,fig.width=3}
min_clades <- 5
all_growth_estimates_df_var <- all_growth_estimates_df %>% 
  group_by(id,clade) %>% 
  mutate(X = X - min(X)) %>% 
  group_by(clade,id) %>%
  mutate(V = mean(W)) %>% 
  ungroup %>% 
  mutate(alpha_W = W <= 5) %>% 
  subset(n_samples >= min_clades) %>%
  mutate(col = ifelse(grepl("DNMT3A",clade_id),"DNMT3A",
                      ifelse(grepl("Clade",clade_id),"No known driver","Other driver"))) %>%
  group_by(clade,id) %>%
  mutate(C = cumsum(c(0,diff(alpha_W) == 1)))

subset(all_growth_estimates_df_var,alpha_W == T) %>%
  ggplot(aes(x = X,y = exp(Y),colour = col,group = paste(clade,C))) + 
  geom_hline(linetype = 2,size = 0.25,yintercept = 50e3) +
  geom_hline(linetype = 2,size = 0.25,yintercept = 200e3) +
  geom_line(data = all_growth_estimates_df_var,aes(group = clade),alpha = 0.3,size = 0.25) +
  geom_line(size = 0.25) + 
  facet_wrap(~ id) + 
  scale_colour_manual(values = c(DNMT3A = "#BE0032",
                                 `Other driver` = "#18A558",
                                 `No known driver` = "grey"),
                      guide = F) + 
  scale_y_continuous(trans = 'log10',expand = c(0,0),breaks = c(1e-2,1,1e2,1e4,1e6)) +
  scale_x_continuous(expand = c(0,0)) +
  coord_cartesian(ylim = c(0.5,1e7)) +
  theme_gerstung(base_size = 6) + 
  xlab("Time since first coal. (years)") +
  ylab("Neff") +
  theme(strip.text = element_text(margin = margin(b = 0.5))) +
  ggsave("figures/model_ch/trees/mitchell_trajectories.pdf",
         height = 1.5,
         width = 1.8)

all_growth_estimates_df %>% 
  group_by(clade,id) %>% 
  mutate(X = X - min(X)) %>% 
  subset(n_samples >= min_clades) %>%
  ggplot(aes(x = X,y = pred_l,colour = as.factor(clade))) + 
  geom_line(aes(y = Y),size = 0.25) + 
  geom_line(size = 0.25,linetype = 2) + 
  facet_wrap(~ id) + 
  scale_color_manual(breaks = c(1:18),
                     values = c(colours[1:18]),
                     guide = F) + 
  scale_y_continuous(breaks = seq(-21,21,by = 3),labels = 10^seq(-21,21,by = 3)) + 
  theme_gerstung(base_size = 6) + 
  xlab("Time since first coalescence (years)") +
  ylab("Effective population size") +
  ggtitle("Linear fit") +
  ggsave("figures/model_ch/trees/mitchell_trajectories_predicted.pdf",
         height = 2,
         width = 3)

all_growth_estimates_df %>% 
  group_by(clade,id) %>% 
  mutate(X = X - min(X)) %>% 
  subset(n_samples >= min_clades) %>%
  ggplot(aes(x = X,y = pred_nl,colour = as.factor(clade))) + 
  geom_line(aes(y = exp(Y)),size = 0.25) + 
  geom_line(size = 0.25,linetype = 2) + 
  facet_wrap(~ id) + 
  scale_color_manual(breaks = c(1:18),
                     values = c(colours[1:18]),
                     guide = F) + 
  scale_y_continuous(trans = 'log10') + 
  theme_gerstung(base_size = 6) + 
  xlab("Time since first coalescence (years)") +
  ylab("Effective population size") +
  ggtitle("Sigmoidal fit") +
  ggsave("figures/model_ch/trees/mitchell_trajectories_predicted_nl.pdf",
         height = 2,
         width = 3)

nl_sc_pop_size <- data.frame(
  clade = rep(NA,4),
  id = c("KX003","KX004","KX007","KX008"),
  clade_id = rep("Total",4),
  N = sapply(individual_bnpr_estimates,function(x) log(x$summary$quant0.5[1])),
  N025 = sapply(individual_bnpr_estimates,function(x) log(x$summary$quant0.025[1])),
  N975 = sapply(individual_bnpr_estimates,function(x) log(x$summary$quant0.975[1])),
  n_samples = sapply(trees,function(x) length(x$tip.label))
)
```

## Trajectories for different types of fit (using only trajectories fitted at coalescence points)

```{r,fig.height=2,fig.width=3}
min_clades <- 5
all_growth_estimates_df_var <- all_growth_estimates_df %>% 
  group_by(id,clade) %>% 
  mutate(X = X - min(X)) %>% 
  group_by(clade,id) %>%
  mutate(V = mean(W)) %>% 
  ungroup %>% 
  mutate(alpha_W = W <= 5) %>% 
  subset(n_samples >= min_clades) %>%
  mutate(col = ifelse(grepl("DNMT3A",clade_id),"DNMT3A",
                      ifelse(grepl("Clade",clade_id),"No known driver","Other driver"))) %>%
  group_by(clade,id) %>%
  mutate(C = cumsum(c(0,diff(alpha_W) == 1)))

subset(all_growth_estimates_df_var,alpha_W == T) %>%
  ggplot(aes(x = X,y = exp(Y),colour = col,group = paste(clade,C))) + 
  geom_rect(aes(xmin = 0,xmax = max(all_growth_estimates_df_var$X),
                ymin = 50e3,ymax = 200e3),fill = "grey92",colour = NA) +
  geom_line(data = all_growth_estimates_df_var,aes(group = clade),linetype = 3,size = 0.25) +
  geom_line(size = 0.25) + 
  facet_wrap(~ id) + 
  scale_colour_manual(values = c(DNMT3A = "#BE0032",
                                 `Other driver` = "#18A558",
                                 `No known driver` = "grey40"),
                      guide = F) + 
  scale_y_continuous(trans = 'log10',expand = c(0,0),breaks = c(1e-2,1,1e2,1e4,1e6)) +
  coord_cartesian(ylim = c(1e-2,1e7)) +
  theme_gerstung(base_size = 6) + 
  xlab("Time since first coal. (years)") +
  ylab("Neff") +
  theme(strip.text = element_text(margin = margin(b = 0.5))) +
  ggsave("figures/model_ch/trees/mitchell_trajectories.pdf",
         height = 1.5,
         width = 1.8)

all_growth_estimates_df %>% 
  group_by(clade,id) %>% 
  mutate(X = X - min(X)) %>% 
  subset(n_samples >= min_clades) %>%
  ggplot(aes(x = X,y = pred_l,colour = as.factor(clade))) + 
  geom_line(aes(y = Y),size = 0.25) + 
  geom_line(size = 0.25,linetype = 2) + 
  facet_wrap(~ id) + 
  scale_color_manual(breaks = c(1:18),
                     values = c(colours[1:18]),
                     guide = F) + 
  scale_y_continuous(breaks = seq(-21,21,by = 3),labels = 10^seq(-21,21,by = 3)) + 
  theme_gerstung(base_size = 6) + 
  xlab("Time since first coalescence (years)") +
  ylab("Effective population size") +
  ggtitle("Linear fit") +
  ggsave("figures/model_ch/trees/mitchell_trajectories_predicted.pdf",
         height = 2,
         width = 3)

all_growth_estimates_df %>% 
  group_by(clade,id) %>% 
  mutate(X = X - min(X)) %>% 
  subset(n_samples >= min_clades) %>%
  ggplot(aes(x = X,y = pred_nl,colour = as.factor(clade))) + 
  geom_line(aes(y = exp(Y)),size = 0.25) + 
  geom_line(size = 0.25,linetype = 2) + 
  facet_wrap(~ id) + 
  scale_color_manual(breaks = c(1:18),
                     values = c(colours[1:18]),
                     guide = F) + 
  scale_y_continuous(trans = 'log10') + 
  theme_gerstung(base_size = 6) + 
  xlab("Time since first coalescence (years)") +
  ylab("Effective population size") +
  ggtitle("Sigmoidal fit") +
  ggsave("figures/model_ch/trees/mitchell_trajectories_predicted_nl.pdf",
         height = 2,
         width = 3)

all_growth_estimates_df %>% 
  group_by(clade,id) %>% 
  mutate(X = X - min(X)) %>% 
  subset(n_samples >= min_clades) %>%
  ggplot(aes(x = X,y = exp(pred_cp),colour = as.factor(clade))) + 
  geom_line(aes(y = exp(Y)),size = 0.25) + 
  geom_line(size = 0.25,linetype = 2) + 
  facet_wrap(~ id) + 
  scale_color_manual(breaks = c(1:18),
                     values = c(colours[1:18]),
                     guide = F) + 
  scale_y_continuous(trans = 'log10') + 
  theme_gerstung(base_size = 6) + 
  xlab("Time since first coalescence (years)") +
  ylab("Effective population size") +
  ggtitle("Changepoint regression") +
  ggsave("figures/model_ch/trees/mitchell_trajectories_predicted_nl.pdf",
         height = 2,
         width = 3)

nl_sc_pop_size <- data.frame(
  clade = rep(NA,4),
  id = c("KX003","KX004","KX007","KX008"),
  clade_id = rep("Total",4),
  N = sapply(individual_bnpr_estimates,function(x) log(x$summary$quant0.5[1])),
  N025 = sapply(individual_bnpr_estimates,function(x) log(x$summary$quant0.025[1])),
  N975 = sapply(individual_bnpr_estimates,function(x) log(x$summary$quant0.975[1])),
  n_samples = sapply(trees,function(x) length(x$tip.label))
)
```

## Trees and trajectories

```{r}
plot_grid(
  plot_grid(all_plots$KX003$tree,all_plots$KX003$bnpr,ncol = 1,rel_heights = c(1,0.5)),
  plot_grid(all_plots$KX004$tree,all_plots$KX004$bnpr,ncol = 1,rel_heights = c(1,0.5)),
  nrow = 1
) + 
  ggsave("figures/model_ch/trees/mitchell_1.pdf",
         height = 6,
         width = 7)

plot_grid(
  plot_grid(all_plots$KX007$tree,all_plots$KX007$bnpr,ncol = 1,rel_heights = c(1,0.5)),
  plot_grid(all_plots$KX008$tree,all_plots$KX008$bnpr,ncol = 1,rel_heights = c(1,0.5)),
  nrow = 1
) + 
  ggsave("figures/model_ch/trees/mitchell_2.pdf",
         height = 6,
         width = 7)

write.csv(all_growth_estimates_values_df,"data_output/mitchell_bnpr_coefficients.csv",row.names = T)
```

## Representation of early and biphasic growth

```{r,fig.height=0.8,fig.width=0.8}
coefs <- all_growth_estimates_values_df[1,]
bnpr_trajectory <- bnpr_estimates$KX003[[1]]$summary %>%
  mutate(time = ages$KX003 - time)

b_early <- coefs$cp_1
b_late <- coefs$cp_2
X <- seq(min(bnpr_trajectory$time),81)
plotting_scheme_exp <- data.frame(
  time = X,
  traj_true = exp(change_point(X,coefs$offset,b_early,b_late - b_early,coefs$midpoint)),
  traj_early = exp(X * b_early + coefs$offset))

plotting_scheme_exp %>%
  mutate(extrapol = traj_true != traj_early) %>% 
  ggplot(aes(x = time,y = traj_early)) + 
  geom_ribbon(data = bnpr_trajectory,aes(x = time,ymin = quant0.025,ymax = quant0.975),
              inherit.aes = F,
              alpha = 0.2) +
  geom_line(data = bnpr_trajectory,aes(x = time,y = quant0.5),
            inherit.aes = F,
            alpha = 0.75,size = 0.25) +
  geom_point(aes(x = 81,y = coefs$vaf_from_clade * 2e5),
             inherit.aes = F,
             colour = "grey50") +
  geom_line(aes(y = traj_true),colour = "blue",size = 0.25) +
  geom_line(colour = "red",size = 0.25,aes(linetype = extrapol)) + 
  scale_y_continuous(trans = 'log10',breaks = NULL) + 
  ylab("Neff") +
  xlab("Time") + 
  theme_gerstung(base_size = 6) + 
  scale_x_continuous(breaks = NULL) + 
  scale_linetype_manual(values = c(1,3),guide = F) +
  ggsave(filename = sprintf("figures/model_ch/dynamic_coefficients/early_late_difference_trajectory.pdf"),
         useDingbats = F,
         width = 0.8,height = 0.8)
```

```{r,fig.height=3,fig.width=5}
all_vafs <- list()

for (i in 1:nrow(all_growth_estimates_values_df)) {
  coefs <- all_growth_estimates_values_df[i,]
  ID <- coefs$id
  clade_number <- coefs$clade_no

  all_vafs[[i]] <- data.frame(
    vaf = coefs$vaf_from_clade,
    vaf_005 = coefs$vaf_from_clade_005,
    vaf_095 = coefs$vaf_from_clade_095,
    clade = coefs$clade,
    id = ID,
    age = ages[[ID]]
  )
}

all_vafs_df <- all_vafs %>%
  do.call(what = rbind)

low_var_estimates <- all_growth_estimates_df %>%
  mutate(clade = clade_id) %>% 
  group_by(clade,id) %>% 
  mutate(WW = mean(W)) %>%
  subset(WW < 10) 

all_vafs_df <- all_vafs_df %>%
  subset(paste(clade,id) %in% paste(low_var_estimates$clade,low_var_estimates$id)) %>%
  arrange(vaf)

low_var_estimates %>% 
  mutate(extrapol = pred_cp != pred_cp_early) %>% 
  ggplot(aes(x = X)) +
  geom_ribbon(aes(ymin = Y025,ymax = Y975),
              alpha = 0.2) +
  geom_line(aes(y = exp(Y)),size = 0.25) +
  geom_line(aes(y = exp(pred_cp)),colour = "blue") +
  geom_line(aes(y = exp(pred_cp_early),linetype = extrapol),colour = "red") + 
  # geom_linerange(data = all_vafs_df,
  #                aes(x = age,ymin = vaf_005*1000e3,ymax = vaf_095*1000e3)) +
  geom_point(data = all_vafs_df,aes(x = age,y = vaf * 1000e3),size = 0.5) +
  geom_point(data = all_vafs_df,aes(x = age,y = vaf * 200e3),size = 0.5,colour = "red") +
  facet_wrap(~ factor(sprintf("%s\n%s",id,clade),paste(all_vafs_df$id,all_vafs_df$clade,sep = "\n")))+ 
  theme_gerstung(base_size = 6) + 
  scale_y_continuous(trans = 'log10') + 
  theme(strip.text = element_text(margin = margin())) + 
  coord_cartesian(ylim = c(1,1e8)) + 
  xlab("Time (years)") + 
  ylab("Neff") + 
  scale_linetype_manual(values = c(1,3),guide = F) +
  ggsave(filename = sprintf("figures/model_ch/trees/early_late_difference_trajectory_all.pdf"),
         useDingbats = F,
         width = 5,height = 3)
```