---
title: "The Natural History of Clonal Haematopoiesis"
subtitle: "Investigating fit to BNPR"
author: 
    - José Guilherme de Almeida
    - Moritz Gerstung
output: 
  html_document:
    theme: lumen
    highlight: tango
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    number_sections: true
---

In this notebook we inspect how we can use the BNPR fits from `phylodyn` to model annual growth.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Setup

```{r}
source("Scripts/vaf_dynamics_functions.R")
source("Scripts/make_ultrametric.R")

library(parallel)
library(Matrix)
library(castor)
library(phangorn)
library(phylodyn)
library(minpack.lm)
INLA:::inla.dynload.workaround()

change_point <- function(x, b0, m1, m2, delta) { 
  b0 + (x*m1) + (sapply(x-delta, function (t) max(0, t)) * m2)
}

plot_tree <- function(obj) { 
  tree_ultra <- obj$tree_ultra
  tree_ultra$edge.length[is.infinite(tree_ultra$edge.length)] <- 0
  tree_ultra$tip.label <- obj$tree$S
  driver_list <- Filter(function(x) length(x) > 5,obj$driver_list)
  if (length(driver_list) > 1) {
    colours <- RColorBrewer::brewer.pal(length(driver_list),"Set3")
  } else {
    colours <- c("black")
  }
  colour_code <- rep("black",length(tree_ultra$tip.label))
  R <- 1:length(driver_list)
  if (length(driver_list) == 0) {
    R <- numeric(0)
  } 
  for (i in R) {
    D <- driver_list[[i]]
    colour_code[tree_ultra$tip.label %in% unique(D)] <- colours[i]
    print(c(names(driver_list)[i],colours[i]))
  }
  print(names(driver_list))
  print(colours)
  plot(tree_ultra,tip.color = colour_code)
}

expand_edge <- function(tree,edge_id) {
  # takes an edge_id and retrieves the tips which are common to this edge
  n_tips <- length(tree$tip.label)
  edge_ids <- c(edge_id)
  done <- F
  while (done == F) {
    new_edge <- tree$edge[,2][tree$edge[,1] %in% edge_id]
    edge_ids <- c(edge_ids,new_edge)
    edge_id <- new_edge
    if (length(edge_id) == 0) {
      done <- T
    }
  }
  edge_ids <- unique(edge_ids)
  edge_ids <- edge_ids[edge_ids <= n_tips]
  return(edge_ids)
}

trim_tree <- function(tree,l) {
  n_tips <- length(tree$tip.label)
  tree$edge.length <- ifelse(
    tree$edge[,2] <= n_tips,
    tree$edge.length - l,
    tree$edge.length
  )
  tree <- drop.tip(
    tree,
    tree$edge[tree$edge.length < 0] %>% sapply(
      function(x) expand_edge(tree,x)) %>% unlist())
  return(tree)
}

bnpr_at_coalescence <- function(bnpr_estimate) {
  time_at_coal <- bnpr_estimate$coal_times
  X <- bnpr_estimate$summary$time
  Y <- bnpr_estimate$summary$quant0.5
  Y025 <- bnpr_estimate$summary$quant0.025
  Y975 <- bnpr_estimate$summary$quant0.975
  Y_log <- log(na.omit(approx(X,Y,xout = time_at_coal,rule = 2)$y))
  values_at_coalescence <- data.frame(
    X = time_at_coal,
    Y = Y_log,
    Y_exp = approx(X,Y,xout = time_at_coal)$y,
    Y025 = approx(X,Y025,xout = time_at_coal)$y,
    Y975 = approx(X,Y975,xout = time_at_coal)$y
  ) %>%
    na.omit()
  return(values_at_coalescence)
}

bnpr_at_all <- function(bnpr_estimate) {
  X <- bnpr_estimate$summary$time
  Y <- bnpr_estimate$summary$quant0.5
  Y025 <- bnpr_estimate$summary$quant0.025
  Y975 <- bnpr_estimate$summary$quant0.975
  Y_log <- log(Y)
  values_at_coalescence <- data.frame(
    X = X,
    Y = Y_log,
    Y_exp = Y,
    Y025 = Y025,
    Y975 = Y975
  ) %>%
    na.omit()
  return(values_at_coalescence)
}

clade_from_mrca <- function(tree,mrca) {
  edge_idx <- which(tree$edge[,1] == mrca)
  continue <- T
  output <- c(which(tree$edge[,2] == mrca))
  while (continue == T) {
    if (length(edge_idx) > 0) {
      output <- c(output,edge_idx)
      new_edge_idx <- c()
      for (idx in edge_idx) {
        nodes <- tree$edge[idx,2]
        new_idxs <- which(tree$edge[,1] == nodes)
        new_edge_idx <- c(new_edge_idx,new_idxs)
      }
      edge_idx <- new_edge_idx
    } else {
      continue <- F
    }
  }
  return(output)
}

build_tree <- function(sub_data,subsample_size=100,
                       detection_threshold=0) {
  sub_af <- sub_data %>% 
    group_by(CloneID) %>% 
    summarise(N = max(NClones),.groups = "drop") %>%
    arrange(CloneID)
  MaxClone <- max(sub_data$CloneID)
  MaxMut <- max(sub_data$MutID)
  sub_af_sparse <- rep(0,MaxClone)
  sub_af_sparse[sub_af$CloneID] <- sub_af$N 
  
  Presence <- sparseMatrix(i = sub_data$CloneID,
                           j = sub_data$MutID,
                           dims = c(MaxClone,MaxMut))
  Presence[cbind(sub_data$CloneID,sub_data$MutID)] <- 1
  Presence_ <- Presence
  
  if (sum(as.logical(sub_af_sparse)) < subsample_size) {
    sub_af_sparse <- rep(1,length(sub_af_sparse))
  }
  
  S <- sample(MaxClone,subsample_size,replace = F,prob = sub_af_sparse)
  af <- sub_af_sparse[S]
  Presence <- Presence[S,]
  Presence <- as.matrix(rbind(Presence,wt=0))
  r_idx <- colSums(Presence)
  Presence <- Presence[,r_idx > 0]
  Presence_ <- Presence_[,r_idx > 0]
  
  dst <- as.matrix(dist(Presence > 0,method = "manhattan"))
  tree <- root(njs(dst),
               outgroup = "wt",
               resolve.root = T,
               edgelabel = T) 
  tree <- drop.tip(tree,"wt")
  # assign clones not in tree to the closest clone in the tree
  # using Hamming distance (which.max(number of identical bits))
  closest_clone <- (Presence %*% t(Presence_)) %>% apply(2,which.max)
  return(list(tree = tree,S = S,af = af,closest_clone = closest_clone))
}

read_clonex <- function(path,d = 1000) {
  x <- read_tsv(path,
                col_names = c("Gen","NClones","CloneID","MutID"),
                col_types = c(col_integer(),col_integer(),col_integer(),col_integer()),
                progress = F) %>%
    mutate(Driver = MutID <= d)
  return(x)
}
```

# BNPR

In this section we calculate/load all BNPR trajectories for the trees inferred from Wright-Fisher (WF) simulations. These simulations are done for 6 different fitness levels - 0.005, 0.010, 0.015, 0.020, 0.025 and 0.030 - over 800 generations and with a fixed population size (200,000).

## Loading data and inferring trees

The first step is to build the trees, which we then display. One can immediately see that expansions become increasingly prevalent for higher fitness effects, with quite a few cases of a single clone sweeping all of the population.

```{r}
N <- 2e5
N_DRIVERS <- 50
all_file_paths <- list.files(
  "hsc_output_bnpr_complete",pattern = "last_generation",
  full.names = T,recursive = T)
all_driver_file_paths <- list.files(
  "hsc_output_bnpr_complete",pattern = "driver",
  full.names = T,recursive = T)

file_name <- "data_output/simulated_trees_trajectories.RDS"
if (file.exists(file_name) == T) {
  tree_traj <- readRDS(file_name)
  trees <- tree_traj[[1]]
  all_driver_trajectories_raw <- tree_traj[[2]]
} else {
  all_driver_trajectories_raw <- all_driver_file_paths %>%
    lapply(function(x) {
      read.csv(x,header = F) %>%
        arrange(V1,V2) %>%
        select(Gen = V1,Count = V2,Clone = V3,MutID = V4)})
  
  trees <- mclapply(
    all_file_paths,
    mc.cores = 2,
    mc.cleanup = T,
    function(path) {
      s <- str_match(path,"hsc_[0-9.]+") %>%
        gsub(pattern = "hsc_",replacement = "") %>%
        as.numeric()
      system(sprintf('echo "%s"', path)) # prints during mclapply by using bash
      x <- read_clonex(path,d = N_DRIVERS) %>%
        subset(Gen == 800) %>%
        subset(MutID != 0)
      x$R <- as.numeric(str_match(path,"[0-9.]+$"))
      x$s <- s
      x$drift_threshold <- 1 / (N * (s))
      
      driver_list <- x %>% 
        subset(Driver == T) %>% 
        select(CloneID,MutID) 
      driver_list_out <- list()
      for (driver_id in unique(driver_list$MutID)) {
        tmp <- driver_list %>%
          subset(MutID == driver_id) %>%
          select(CloneID) %>%
          unlist
        driver_list_out[[as.character(driver_id)]] <- tmp
      }
      tree <- x %>%
        build_tree()
      ultrametric_tree <- make.ultrametric.tree(tree$tree)
      gc()
      list(
        tree = tree,
        tree_ultra = ultrametric_tree,
        driver_list = driver_list_out,
        path = path
      ) %>%
        return
    }
  )
  
  names(trees) <- gsub('/last_generation','',all_file_paths)
  names(all_driver_trajectories_raw) <- gsub('/driver_trajectory','',all_driver_file_paths)
  saveRDS(object = list(trees,all_driver_trajectories_raw),file = file_name)
}

for (tree_name in names(trees)) {
  trees[[tree_name]]$tree_ultra$edge.length <- trees[[tree_name]]$tree_ultra$edge.length * 800
}
```

```{r}
par(mfrow = c(4,5),mar = c(2,0.5,1,0.5))
for (tree_name in names(trees)) {
  tree <- trees[[tree_name]]
  fitness <- str_match(tree_name,"[0-9.]+")
  plot(tree$tree_ultra,main = fitness)
}
```

## Calculating BNPR trajectories for the whole trees

Here we calculate the actual EPS trajectories using BNPR and display them.

```{r}
file_name <- "data_output/simulated_bnpr_trees.RDS"
if (file.exists(file_name) == T) {
  all_estimates_trees <- readRDS(file_name)
} else {
  all_estimates_trees <- lapply(
    trees,
    function(x) {
      system(sprintf('echo "%s"', x$path)) # prints during mclapply by using bash
      BNPR(x$tree_ultra) %>%
        return
      }
    )
  saveRDS(all_estimates_trees,file_name)
}
```

## Calculating BNPR trajectories for each clade

```{r}
file_name <- "data_output/simulated_bnpr_clades.RDS"
if (file.exists(file_name) == T) {
  all_estimates_full <- readRDS(file_name)
  all_estimates <- all_estimates_full[[1]]
  all_estimates_trimmed_5_years <- all_estimates_full[[2]]
  all_estimates_trimmed_10_years <- all_estimates_full[[3]]
} else {
  all_estimates <- list()
  all_estimates_trimmed_5_years <- list()
  all_estimates_trimmed_10_years <- list()
  i <- 1
  for (obj_name in names(trees)) {
    obj <- trees[[obj_name]]
    print(obj_name)
    s <- str_match(obj_name,"[0-9.]+") %>%
      as.numeric
    Tree <- list(tree = obj$tree_ultra)
    Tree$tree$tip.label <- obj$tree$S
    Tree$tree$edge.length[is.infinite(Tree$tree$edge.length)] <- 0
    tree <- Tree$tree
    tree_ <- tree
    tree_$edge.length <- tree_$edge.length / 800
    driver_id_list <- Filter(function(x) length(x) > 5,obj$driver_list)
    all_drivers <- do.call(c,driver_id_list)
    clades <- cut_tree(tree = tree_,depth = 0.1) %>%
      Filter(f = function(x) length(x) >= 5)
    clades_ <- list()
    for (clade in clades) {
      clade_tips <- tree$tip.label[clade]
      for (driver in names(driver_id_list)) {
        if (all(clade_tips %in% driver_id_list[[driver]])) {
          clades_[[length(clades_)+1]] <- list(
            clade = clade,
            driver = driver
          )
        }
      }
    }
    clades <- clades_
    # clades <- names(driver_id_list) %>%
    #   lapply(function(x) {
    #     return(list(clade = which(tree_$tip.label %in% driver_id_list[[x]]),
    #                 driver = x))
    #     })

    if (!(obj_name %in% names(all_estimates))) {
      bnpr_estimates <- lapply(
        clades,
        #mc.cores = 2,
        #mc.cleanup = T,
        function(clad) {
          #tip_in_clade <- which(Tree$tree$tip.label %in% clad)
          tip_in_clade <- clad$clade
          if (length(tip_in_clade) > 4) {
            sub_tree <- keep.tip(Tree$tree,tip_in_clade) %>%
              multi2di()
            if (!is.null(sub_tree)) {
              bnpr_estimates <- NULL
              bnpr_estimate <- BNPR(sub_tree)
              if (!is.null(bnpr_estimate)) {
                list(
                  tree = tree,
                  bnpr = bnpr_estimate,
                  clade = clad$clade,
                  tree = sub_tree,
                  driver = clad$driver) %>%
                  return
              }
            }
          }
        }
      ) %>%
        Filter(f = function(x) !is.null(x))
      all_estimates[[obj_name]] <- bnpr_estimates
    }
    gc()
  }
  
  # trimmed 5 years
  i <- 1
  for (obj_name in names(trees)) {
    print(obj_name)
    obj <- trees[[obj_name]]
    s <- str_match(obj_name,"[0-9.]+") %>%
      as.numeric
    Tree <- list(tree = obj$tree_ultra)
    Tree$tree$tip.label <- obj$tree$S
    Tree$tree <- trim_tree(Tree$tree,l = 50)
    Tree$tree$edge.length[is.infinite(Tree$tree$edge.length)] <- 0
    tree <- Tree$tree
    tree_ <- tree
    tree_$edge.length <- tree_$edge.length / 800
    driver_id_list <- Filter(function(x) length(x) > 5,obj$driver_list)
    all_drivers <- do.call(c,driver_id_list)
    clades <- cut_tree(tree = tree_,depth = 0.1) %>%
      Filter(f = function(x) length(x) >= 5)
    clades_ <- list()
    for (clade in clades) {
      clade_tips <- tree$tip.label[clade]
      for (driver in names(driver_id_list)) {
        if (all(clade_tips %in% driver_id_list[[driver]])) {
          clades_[[length(clades_)+1]] <- list(
            clade = clade,
            driver = driver
          )
        }
      }
    }
    clades <- clades_
    
    if (!(obj_name %in% names(all_estimates_trimmed_5_years))) {
      bnpr_estimates <- lapply(
        clades,
        #mc.cores = 2,
        #mc.cleanup = T,
        function(clad) {
          #tip_in_clade <- which(Tree$tree$tip.label %in% clad)
          tip_in_clade <- clad$clade
          if (length(tip_in_clade) > 4) {
            sub_tree <- keep.tip(Tree$tree,tip_in_clade) %>%
              multi2di()
            if (!is.null(sub_tree)) {
              bnpr_estimates <- NULL
              bnpr_estimate <- BNPR(sub_tree)
              if (!is.null(bnpr_estimate)) {
                list(
                  tree = tree,
                  bnpr = bnpr_estimate,
                  clade = clad$clade,
                  tree = sub_tree,
                  driver = clad$driver) %>%
                  return
              }
            }
          }
        }
      ) %>%
        Filter(f = function(x) !is.null(x))
      all_estimates_trimmed_5_years[[obj_name]] <- bnpr_estimates
    }
    gc()
  }

  # trimmed 10 years
  i <- 1
  for (obj_name in names(trees)) {
    print(obj_name)
    obj <- trees[[obj_name]]
    s <- str_match(obj_name,"[0-9.]+") %>%
      as.numeric
    Tree <- list(tree = obj$tree_ultra)
    Tree$tree$tip.label <- obj$tree$S
    Tree$tree <- trim_tree(Tree$tree,l = 50)
    Tree$tree$edge.length[is.infinite(Tree$tree$edge.length)] <- 0
    tree <- Tree$tree
    tree_ <- tree
    tree_$edge.length <- tree_$edge.length / 800
    driver_id_list <- Filter(function(x) length(x) > 5,obj$driver_list)
    all_drivers <- do.call(c,driver_id_list)
    clades <- cut_tree(tree = tree_,depth = 0.1) %>%
      Filter(f = function(x) length(x) >= 5)
    clades_ <- list()
    for (clade in clades) {
      clade_tips <- tree$tip.label[clade]
      for (driver in names(driver_id_list)) {
        if (all(clade_tips %in% driver_id_list[[driver]])) {
          clades_[[length(clades_)+1]] <- list(
            clade = clade,
            driver = driver
          )
        }
      }
    }
    clades <- clades_
    
    if (!(obj_name %in% names(all_estimates_trimmed_10_years))) {
      bnpr_estimates <- lapply(
        clades,
        #mc.cores = 2,
        #mc.cleanup = T,
        function(clad) {
          #tip_in_clade <- which(Tree$tree$tip.label %in% clad)
          tip_in_clade <- clad$clade
          if (length(tip_in_clade) > 4) {
            sub_tree <- keep.tip(Tree$tree,tip_in_clade) %>%
              multi2di()
            if (!is.null(sub_tree)) {
              bnpr_estimates <- NULL
              bnpr_estimate <- BNPR(sub_tree)
              if (!is.null(bnpr_estimate)) {
                list(
                  tree = tree,
                  bnpr = bnpr_estimate,
                  clade = clad$clade,
                  tree = sub_tree,
                  driver = clad$driver) %>%
                  return
              }
            }
          }
        }
      ) %>%
        Filter(f = function(x) !is.null(x))
      all_estimates_trimmed_10_years[[obj_name]] <- bnpr_estimates
    }
    gc()
  }
  saveRDS(list(all_estimates,all_estimates_trimmed_5_years,all_estimates_trimmed_10_years),file_name)
}
```

## Timing clones from trees

```{r}
for (x in names(all_estimates)) {
  estimate <- all_estimates[[x]]
  if (length(estimate) > 0) {
    for (y in 1:length(estimate)) {
      m <- MRCA(estimate[[y]]$tree,estimate[[y]]$clade)
      estimate[[y]]$mutation_timing <- time_mutation(
        estimate[[y]]$tree,m)
      all_estimates[[x]][[y]]$mutation_timing <- estimate[[y]]$mutation_timing
    }
  }
}
```

# Modelling BNPR trajectories

Here we model the EPS trajectories using different types of models, namely:

* A log-linear model
* A scaled and shifted sigmoidal curve
* A log-linear model with one changepoint (captures the earlier and later phases of growth)

We also fit the same models to the original WF trajectories one gets for each individual driver. By doing so, we can more accurately verify whether we are able to recapitulate the clonal growth as it is happening with the BNPR trajectories.

## Fitting different models to BNPR trajectory

```{r}
all_fits <- list()
par(mfrow = c(4,5),mar = c(2,2,1,1))
for (obj_name in names(all_estimates)) {
  bnpr_estimates <- all_estimates[[obj_name]] %>%
    Filter(f = function(x) class(x$bnpr) != "try-error")
  new_bnpr_estimates <- list()
  for (x in bnpr_estimates) {
    bnpr_estimate <- x$bnpr
    vaf_from_clade <- length(x$clade) / 100
    Y <- log(bnpr_estimate$summary$mean)
    Y_ <- Y - min(Y,na.rm = T) + 1e-8
    Y025 <- log(bnpr_estimate$summary$quant0.025)
    Y975 <- log(bnpr_estimate$summary$quant0.975)
    X <- (800-bnpr_estimate$summary$time)
    tmp_df <- data.frame(
      X = X,Y = Y,Y_ = Y_,
      Y025 = Y025,Y975 = Y975,
      W = (Y975 - Y025)^2/16
    ) %>% 
      na.omit()
    R <- range(tmp_df$X)
    tmp_df <- tmp_df[!is.infinite(tmp_df$Y),]
    tmp_df <- tmp_df[!is.infinite(tmp_df$Y025),]
    tmp_df <- tmp_df[!is.infinite(tmp_df$Y975),]
    tmp_df$Y_exp <- exp(tmp_df$Y)
    
    tmp_df$W[is.infinite(tmp_df$W)] <- max(tmp_df$W[!is.infinite(tmp_df$W)])
    linear_estimate <- lm(Y ~ X,w = 1 / tmp_df$W,data = tmp_df)
    m <- linear_estimate$coefficients[2]
    linear_estimate_no_weights <- lm(Y ~ X,data = tmp_df)
    
    if (nrow(tmp_df) >= 10 & max(tmp_df$Y_exp) < 1e100) {
      non_linear_estimate <- nlsLM(
        Y_exp ~ SSlogis(X, Asym, b2, b3),
        start = list(Asym = max(tmp_df$Y_exp),b2 = mean(tmp_df$X),b3 = 10),
        control = nls.control(maxiter = 1000,warnOnly = T),
        data = tmp_df,
        algorithm = "port",
        weights = 1/tmp_df$W)
      non_linear_estimate_vaf <- nls(
        Y_exp / Y_exp[1] * vaf_from_clade ~ SSlogis(X,1,theta2,theta3),
        weights = 1/W,data = tmp_df,
        control = nls.control(warnOnly = T,
                              maxiter = 1000,
                              minFactor = 1 / (1024^16)),
        start = c(theta2 = max(X),theta3 = 5))

      x_earliest <- seq(min(tmp_df$X),min(tmp_df$X) + 150,length.out = 20)
      y_earliest <- approx(tmp_df$X,tmp_df$Y,xout = x_earliest)$y
      
      x_latest <- seq(max(tmp_df$X) - 150,max(tmp_df$X),length.out = 20)
      y_latest <- approx(tmp_df$X,tmp_df$Y,xout = x_latest)$y
      
      latest <- lm(y_latest ~ x_latest)
      earliest <- lm(y_earliest ~ x_earliest)

      if (class(non_linear_estimate) == "try-error") {
        non_linear_estimate <- try(nlsLM(
          Y_exp ~ SSlogis(X, Asym, b2, b3),
          start = list(Asym = max(tmp_df$Y_exp),b2 = max(tmp_df$X),b3 = 10),
          control = nls.control(maxiter = 1000,warnOnly = T),
          data = tmp_df,
          algorithm = "port",
          weights = 1/tmp_df$W))
      }
    } else {
      non_linear_estimate <- NULL
      non_linear_estimate_vaf <- NULL
      earliest <- NULL
      latest <- NULL
    }
    
    opt_fn <- function(par) {
      se <- (tmp_df$Y - change_point(tmp_df$X,par[1],par[2],par[3],par[4]))^2
      return(mean(se / tmp_df$W))
    }
    change_point_regression <- optim(
      par = c(linear_estimate$coefficients[1],
              linear_estimate$coefficients[2],
              0,
              mean(X)),
      method = "L-BFGS-B",
      lower = c(NA,0,NA,min(tmp_df$X) + diff(R) * 0.25),
      upper = c(NA,NA,NA,max(tmp_df$X) - diff(R) * 0.25),
      fn = opt_fn)
    x$linear_estimate <- linear_estimate
    x$linear_estimate_no_weights <- linear_estimate_no_weights
    x$change_point_regression <- change_point_regression
    x$non_linear_estimate <- non_linear_estimate
    x$non_linear_estimate_vaf <- non_linear_estimate_vaf
    x$data <- tmp_df
    x$earliest_growth <- earliest
    x$latest_growth <- latest
    new_bnpr_estimates[[length(new_bnpr_estimates) + 1]] <- x
  }
  
  bnpr_estimates <- new_bnpr_estimates %>%
    Filter(f = function(x) !is.null(x))
  
  all_estimates[[obj_name]] <- bnpr_estimates
  for (bnpr in bnpr_estimates) {
    cpr <- bnpr$change_point_regression
    all_fits[[length(all_fits)+1]] <- data.frame(
      x = bnpr$data$X,y = exp(bnpr$data$Y),
      w = bnpr$data$W,
      obj_name,n = length(all_fits)+1,
      clade_size = length(bnpr$clade))

    # if ((cpr$par[2] + cpr$par[3]) < 100) {
    #     plot_BNPR(bnpr$bnpr,main = sprintf("%s",length(bnpr$clade)))
    #     # if (!is.null(bnpr$non_linear_estimate)) {
    #     #   lines(bnpr$data$X,predict(
    #     #     bnpr$non_linear_estimate,newdata = data.frame(X = bnpr$data$X)),
    #     #     col = "red")
    #     # }
    #     lines(800-bnpr$data$X,
    #           exp(change_point(bnpr$data$X,cpr$par[1],cpr$par[2],cpr$par[3],cpr$par[4])),
    #           col="red")
    #   }
    }
}
```

## Fitting different models to BNPR trajectory (5 year trimmed)

```{r}
par(mfrow = c(4,5),mar = c(3,1,1,1))
for (obj_name in names(all_estimates_trimmed_5_years)) {
  bnpr_estimates <- all_estimates_trimmed_5_years[[obj_name]] %>%
    Filter(f = function(x) class(x$bnpr) != "try-error")
  new_bnpr_estimates <- list()
  for (x in bnpr_estimates) {
    bnpr_estimate <- x$bnpr
    Y <- log(bnpr_estimate$summary$mean)
    Y_ <- Y - min(Y,na.rm = T) + 1e-8
    Y025 <- log(bnpr_estimate$summary$quant0.025)
    Y975 <- log(bnpr_estimate$summary$quant0.975)
    X <- (800-bnpr_estimate$summary$time)
    tmp_df <- data.frame(
      X = X,Y = Y,Y_ = Y_,
      Y025 = Y025,Y975 = Y975,
      W = (Y975 - Y025)^2/16
    ) %>% 
      na.omit() 
    R <- range(tmp_df$X)
    tmp_df <- tmp_df[!is.infinite(tmp_df$Y),]
    tmp_df <- tmp_df[!is.infinite(tmp_df$Y025),]
    tmp_df <- tmp_df[!is.infinite(tmp_df$Y975),]
    tmp_df$Y_exp <- exp(tmp_df$Y)
    
    tmp_df$W[is.infinite(tmp_df$W)] <- max(tmp_df$W[!is.infinite(tmp_df$W)])
    linear_estimate <- lm(Y ~ X,w = 1 / tmp_df$W,data = tmp_df)
    m <- linear_estimate$coefficients[2]
    linear_estimate_no_weights <- lm(Y ~ X,data = tmp_df)
    
    if (nrow(tmp_df) >= 10 & max(tmp_df$Y_exp) < 1e100) {
      non_linear_estimate <- try(nlsLM(
        Y_exp ~ SSlogis(X, Asym, b2, b3),
        start = list(Asym = max(tmp_df$Y_exp),b2 = mean(tmp_df$X),b3 = 10),
        control = nls.control(maxiter = 1000,warnOnly = T),
        data = tmp_df,
        algorithm = "port",
        weights = 1/tmp_df$W))
      x_earliest <- seq(min(tmp_df$X),min(tmp_df$X) + 150,length.out = 20)
      y_earliest <- approx(tmp_df$X,tmp_df$Y,xout = x_earliest)$y
      
      x_latest <- seq(max(tmp_df$X) - 150,max(tmp_df$X),length.out = 20)
      y_latest <- approx(tmp_df$X,tmp_df$Y,xout = x_latest)$y
      
      latest <- lm(y_latest ~ x_latest)
      earliest <- lm(y_earliest ~ x_earliest)

      if (class(non_linear_estimate) == "try-error") {
        non_linear_estimate <- try(nlsLM(
          Y_exp ~ SSlogis(X, Asym, b2, b3),
          start = list(Asym = max(tmp_df$Y_exp),b2 = max(tmp_df$X),b3 = 10),
          control = nls.control(maxiter = 1000,warnOnly = T),
          data = tmp_df,
          algorithm = "port",
          weights = 1/tmp_df$W))
      }
    } else {
      non_linear_estimate <- NULL
      earliest <- NULL
      latest <- NULL
    }
    
    opt_fn <- function(par) {
      se <- (tmp_df$Y - change_point(tmp_df$X,par[1],par[2],par[3],par[4]))^2
      return(mean(se / tmp_df$W))
    }
    change_point_regression <- optim(
      par = c(linear_estimate$coefficients[1],
              linear_estimate$coefficients[2],
              0,
              mean(X)),
      method = "L-BFGS-B",
      lower = c(NA,0,NA,min(tmp_df$X) + diff(R) * 0.25),
      upper = c(NA,NA,NA,max(tmp_df$X) - diff(R) * 0.25),
      fn = opt_fn)
    x$linear_estimate <- linear_estimate
    x$linear_estimate_no_weights <- linear_estimate_no_weights
    x$change_point_regression <- change_point_regression
    x$non_linear_estimate <- non_linear_estimate
    x$data <- tmp_df
    x$earliest_growth <- earliest
    x$latest_growth <- latest
    new_bnpr_estimates[[length(new_bnpr_estimates) + 1]] <- x
  }
  
  bnpr_estimates <- new_bnpr_estimates %>%
    Filter(f = function(x) !is.null(x))
  
  all_estimates_trimmed_5_years[[obj_name]] <- bnpr_estimates
}
```


## Fitting different models to BNPR trajectory (10 year trimmed)

```{r}
par(mfrow = c(4,5),mar = c(3,1,1,1))
for (obj_name in names(all_estimates_trimmed_10_years)) {
  bnpr_estimates <- all_estimates_trimmed_10_years[[obj_name]] %>%
    Filter(f = function(x) class(x$bnpr) != "try-error")
  new_bnpr_estimates <- list()
  for (x in bnpr_estimates) {
    bnpr_estimate <- x$bnpr
    Y <- log(bnpr_estimate$summary$mean)
    Y_ <- Y - min(Y,na.rm = T) + 1e-8
    Y025 <- log(bnpr_estimate$summary$quant0.025)
    Y975 <- log(bnpr_estimate$summary$quant0.975)
    X <- (800-bnpr_estimate$summary$time)
    tmp_df <- data.frame(
      X = X,Y = Y,Y_ = Y_,
      Y025 = Y025,Y975 = Y975,
      W = (Y975 - Y025)^2/16
    ) %>% 
      na.omit()
    R <- range(tmp_df$X)
    tmp_df <- tmp_df[!is.infinite(tmp_df$Y),]
    tmp_df <- tmp_df[!is.infinite(tmp_df$Y025),]
    tmp_df <- tmp_df[!is.infinite(tmp_df$Y975),]
    tmp_df$Y_exp <- exp(tmp_df$Y)
    
    tmp_df$W[is.infinite(tmp_df$W)] <- max(tmp_df$W[!is.infinite(tmp_df$W)])
    linear_estimate <- lm(Y ~ X,w = 1 / tmp_df$W,data = tmp_df)
    m <- linear_estimate$coefficients[2]
    linear_estimate_no_weights <- lm(Y ~ X,data = tmp_df)
    
    if (nrow(tmp_df) >= 10 & max(tmp_df$Y_exp) < 1e100) {
      non_linear_estimate <- try(nlsLM(
        Y_exp ~ SSlogis(X, Asym, b2, b3),
        start = list(Asym = max(tmp_df$Y_exp),b2 = mean(tmp_df$X),b3 = 10),
        control = nls.control(maxiter = 1000,warnOnly = T),
        data = tmp_df,
        algorithm = "port",
        weights = 1/tmp_df$W))
      x_earliest <- seq(min(tmp_df$X),min(tmp_df$X) + 150,length.out = 20)
      y_earliest <- approx(tmp_df$X,tmp_df$Y,xout = x_earliest)$y
      
      x_latest <- seq(max(tmp_df$X) - 150,max(tmp_df$X),length.out = 20)
      y_latest <- approx(tmp_df$X,tmp_df$Y,xout = x_latest)$y
      
      latest <- lm(y_latest ~ x_latest)
      earliest <- lm(y_earliest ~ x_earliest)

      if (class(non_linear_estimate) == "try-error") {
        non_linear_estimate <- try(nlsLM(
          Y_exp ~ SSlogis(X, Asym, b2, b3),
          start = list(Asym = max(tmp_df$Y_exp),b2 = max(tmp_df$X),b3 = 10),
          control = nls.control(maxiter = 1000,warnOnly = T),
          data = tmp_df,
          algorithm = "port",
          weights = 1/tmp_df$W))
      }
    } else {
      non_linear_estimate <- NULL
      earliest <- NULL
      latest <- NULL
    }
    
    opt_fn <- function(par) {
      se <- (tmp_df$Y - change_point(tmp_df$X,par[1],par[2],par[3],par[4]))^2
      return(mean(se / tmp_df$W))
    }
    change_point_regression <- optim(
      par = c(linear_estimate$coefficients[1],
              linear_estimate$coefficients[2],
              0,
              mean(X)),
      method = "L-BFGS-B",
      lower = c(NA,0,NA,min(tmp_df$X) + diff(R) * 0.25),
      upper = c(NA,NA,NA,max(tmp_df$X) - diff(R) * 0.25),
      fn = opt_fn)
    x$linear_estimate <- linear_estimate
    x$linear_estimate_no_weights <- linear_estimate_no_weights
    x$change_point_regression <- change_point_regression
    x$non_linear_estimate <- non_linear_estimate
    x$data <- tmp_df
    x$earliest_growth <- earliest
    x$latest_growth <- latest
    new_bnpr_estimates[[length(new_bnpr_estimates) + 1]] <- x
  }
  
  bnpr_estimates <- new_bnpr_estimates %>%
    Filter(f = function(x) !is.null(x))
  
  all_estimates_trimmed_10_years[[obj_name]] <- bnpr_estimates
}
```

### Plotting estimates with distinct trimming

```{r,fig.height=5,fig.width=6}
trajectories_trimmed_0 <- names(all_estimates) %>%
  lapply(function(x) do.call(
    rbind,lapply(all_estimates[[x]],
                 function(y) data.frame(y$data,id = x,clade = paste(y$clade,collapse=','),
                                        clade_size = length(y$clade))))) %>%
  do.call(what = rbind) %>%
  mutate(Trimmed = 0)

trajectories_trimmed_5 <- names(all_estimates_trimmed_5_years) %>%
  lapply(function(x) do.call(
    rbind,lapply(all_estimates_trimmed_5_years[[x]],
                 function(y) data.frame(y$data,id = x,clade = paste(y$clade,collapse=','),
                                        clade_size = length(y$clade))))) %>%
  do.call(what = rbind) %>% 
  mutate(Trimmed = 5) 

trajectories_trimmed_10 <- names(all_estimates_trimmed_10_years) %>%
  lapply(function(x) do.call(
    rbind,lapply(all_estimates_trimmed_10_years[[x]],
                 function(y) data.frame(y$data,id = x,clade = paste(y$clade,collapse=','),
                                        clade_size = length(y$clade))))) %>%
  do.call(what = rbind) %>%
  mutate(Trimmed = 10)

rbind(trajectories_trimmed_0,
      trajectories_trimmed_5,
      trajectories_trimmed_10) %>%
  group_by(clade,id) %>%
  filter(all(c(0,5,10) %in% Trimmed)) %>%
  subset(paste(id,clade) %in% sample(unique(paste(id,clade)),40,replace = F)) %>% 
  group_by(clade,id,Trimmed) %>%
  mutate(w_sum = mean(W)) %>% 
  mutate(X = X - min(X)) %>% 
  mutate(Trimmed = c(`0`="Original",`5`="Trimmed by 5 years",`10`="Trimmed by 10 years")[as.character(Trimmed)]) %>%
  mutate(Trimmed = factor(Trimmed,
                          levels = c("Original","Trimmed by 5 years","Trimmed by 10 years"))) %>% 
  ungroup %>% 
  mutate(unique_id = as.character(as.numeric(as.factor(paste(id,clade))))) %>%
  group_by(clade,id) %>% 
  filter(all(w_sum < 10)) %>% 
  ggplot(aes(x = X,y = exp(Y),group = paste(id,clade,Trimmed),colour = Trimmed,size = Trimmed)) + 
  geom_line(alpha = 0.3) + 
  scale_y_continuous(trans = 'log10') + 
  coord_cartesian(ylim = c(10,1e10)) + 
  facet_wrap(~ reorder(sprintf("id=%s; N=%s",unique_id,clade_size),clade_size)) + 
  theme_gerstung(base_size = 6) +
  theme(strip.text = element_text(margin = margin())) + 
  scale_color_aaas(name = NULL) + 
  scale_size_manual(values = c(0.5,1.0,1.5),name = NULL) +
  xlab("Time since first coalescence (generations)") + 
  ylab("Neff") + 
  theme(legend.position = "bottom",legend.box.spacing = unit(0,"cm")) + 
  ggsave("figures/simulations/trimmed_representation.pdf",height = 5,width = 6)
```

## Fitting different models to BNPR trajectory (only at coalescence)

```{r}
all_estimates_coalescence <- all_estimates
par(mfrow = c(4,5),mar = c(3,1,1,1))
for (obj_name in names(all_estimates_coalescence)) {
  bnpr_estimates <- all_estimates_coalescence[[obj_name]] %>%
    Filter(f = function(x) class(x$bnpr) != "try-error")
  new_bnpr_estimates <- list()
  for (x in bnpr_estimates) {
    x$linear_estimate <- NULL
    x$linear_estimate_no_weights <- NULL
    x$change_point_regression <- NULL
    x$earliest_growth <- NULL
    x$latest_growth <- NULL
    x$data <- NULL
    tmp_df <- bnpr_at_coalescence(x$bnpr) %>%
      mutate(W = (log(Y975) - log(Y025))^2/16,
             Y_ = Y - min(Y,na.rm = T) + 1e-8,
             X = 800 - X) %>%
      na.omit
    tmp_df <- tmp_df[!is.infinite(tmp_df$Y),]
    tmp_df$W[is.infinite(tmp_df$W)] <- max(tmp_df$W[!is.infinite(tmp_df$W)])
    
    if (nrow(tmp_df) >= 5) {
      linear_estimate <- lm(Y ~ X,w = 1 / tmp_df$W,data = tmp_df)
      m <- linear_estimate$coefficients[2]
      linear_estimate_no_weights <- lm(Y ~ X,data = tmp_df)
      
      if (nrow(tmp_df) >= 10 & max(tmp_df$Y_exp) < 1e100) {
        non_linear_estimate <- try({nlsLM(
          Y_exp ~ SSlogis(X, Asym, b2, b3),
          start = list(Asym = max(tmp_df$Y_exp),b2 = mean(tmp_df$X),b3 = 10),
          control = nls.control(maxiter = 1000,warnOnly = T),
          data = tmp_df,
          algorithm = "port",
          weights = 1/tmp_df$W)})
  
        if (class(non_linear_estimate) == "try-error") {
          non_linear_estimate <- try(nlsLM(
            Y_exp ~ SSlogis(X, Asym, b2, b3),
            start = list(Asym = max(tmp_df$Y_exp),b2 = max(tmp_df$X),b3 = 10),
            control = nls.control(maxiter = 1000,warnOnly = T),
            data = tmp_df,
            algorithm = "port",
            weights = 1/tmp_df$W))
        }
      } else {
        non_linear_estimate <- NULL
      }
      
      opt_fn <- function(par) {
        se <- (tmp_df$Y - change_point(tmp_df$X,par[1],par[2],par[3],par[4]))^2
        return(mean(se / tmp_df$W))
      }
      change_point_regression <- optim(
        par = c(linear_estimate$coefficients[1],
                linear_estimate$coefficients[2],
                0,
                mean(tmp_df$X)),
        method = "L-BFGS-B",
        lower = c(NA,0,NA,min(tmp_df$X)),
        upper = c(NA,NA,NA,max(tmp_df$X)),
        fn = opt_fn)
      x$linear_estimate <- linear_estimate
      x$linear_estimate_no_weights <- linear_estimate_no_weights
      x$change_point_regression <- change_point_regression
      x$non_linear_estimate <- non_linear_estimate
      x$data <- tmp_df
      new_bnpr_estimates[[length(new_bnpr_estimates) + 1]] <- x
    }
  }
  
  bnpr_estimates <- new_bnpr_estimates %>%
    Filter(f = function(x) !is.null(x))
  
  all_estimates_coalescence[[obj_name]] <- bnpr_estimates
}
```

```{r}
par(mfrow = c(4,5),mar = c(2,2,1,1))
for (estimate in all_estimates) {
  for (clade in estimate) {
    if (length(clade) > 0) {
      if (!is.null(clade$linear_estimate)) {
        plot_BNPR(clade$bnpr)
        #points(800 - clade$data$X,clade$data$Y_exp,col="red")
        P <- clade$change_point_regression$par
        lines(clade$bnpr$summary$time,
              exp(change_point((800 - clade$bnpr$summary$time),P[1],P[2],P[3],P[4])),
              col = "red")
        }
      }
    }
  }
```

## Fitting different models to driver trajectories from the original WF simulations

```{r}
all_driver_fits <- list()
par(mfrow = c(4,5),mar = c(1.5,2.5,2.5,1))
for (x in names(all_driver_trajectories_raw)) {
  driver_trajectories <- all_driver_trajectories_raw[[x]]
  driver_trajectories$ClosestTip <- trees[[x]]$tree$closest_clone[driver_trajectories$Clone]
  driver_trajectories$Clade <- NA
  if (length(all_estimates[[x]]) > 0) {
      for (i in 1:length(all_estimates[[x]])) {
        estimate <- all_estimates[[x]][[i]]
        driver_trajectories$Clade[driver_trajectories$ClosestTip %in% estimate$clade] <- i
        }
  }
  
  driver_trajectories <- driver_trajectories %>%
    group_by(Clade,Gen,MutID) %>%
    summarise(Count = sum(Count))
  clades <- unique(driver_trajectories$Clade)
  clades <- clades[!is.na(clades)]
  all_driver_fits[[x]] <- list()
  for (clade in clades) {
    tmp_df <- driver_trajectories %>%
      subset(Clade == clade)
    drivers <- unique(tmp_df$MutID)
    tmp_df <- tmp_df %>% 
      group_by(Clade,Gen) %>%
      summarise(Count = sum(Count))
    clade <- tmp_df$Clade[1]
    has_last_gen <- any(tmp_df$Gen == 800)
    large_at_last_gen <- ifelse(
      has_last_gen,
      tmp_df$Count[tmp_df$Gen == 800] > 200,
      F)
    
    if (nrow(tmp_df) >= 5 & has_last_gen & large_at_last_gen) {
      x_earliest <- seq(min(tmp_df$Gen),min(tmp_df$Gen) + 50,length.out = 20)
      y_earliest <- approx(tmp_df$Gen,tmp_df$Count,xout = x_earliest)$y
      
      x_latest <- seq(max(tmp_df$Gen) - 50,max(tmp_df$Gen),length.out = 20)
      y_latest <- approx(tmp_df$Gen,tmp_df$Count,xout = x_latest)$y
      
      latest <- lm(y_latest ~ x_latest)
      earliest <- lm(y_earliest ~ x_earliest)

      linear_sol <- lm(log(Count) ~ Gen,data = tmp_df)
      opt_fn <- function(par) {
        se <- (log(tmp_df$Count) - change_point(tmp_df$Gen,par[1],par[2],par[3],par[4]))^2
        return(mean(se))
      }
      change_point_regression <- optim(
        par = c(linear_sol$coefficients[1],linear_sol$coefficients[2],0,mean(tmp_df$Gen)),
        method = "L-BFGS-B",
        lower = c(NA,0,NA,min(tmp_df$Gen) + diff(range(tmp_df$Gen)) * 0.25),
        upper = c(NA,NA,NA,max(tmp_df$Gen) - diff(range(tmp_df$Gen)) * 0.25),
        fn = opt_fn)
      non_linear_estimate <- nlsLM(
        Count ~ SSlogis(Gen, 2e5, b2, b3),
        start = list(b2 = mean(tmp_df$Gen),b3 = 10),
        control = nls.control(maxiter = 1000,warnOnly = T),
        data = tmp_df,
        upper = c(b2 = 3000,b3 = 1e5),
        algorithm = "port")
      last_vaf <- tmp_df$Count[tmp_df$Gen == 800]/2e5
      all_driver_fits[[x]][[as.character(clade)]] <- list(
        data = tmp_df,
        non_linear_estimate = non_linear_estimate,
        change_point_regression = change_point_regression,
        earliest_driver = earliest,
        latest_driver = latest,
        last_vaf = last_vaf,
        drivers = drivers,
        gen_at_onset = min(tmp_df$Gen))
    }
  }
}
```

## Extracting coefficients from BNPR fits

```{r}
all_bnpr_fits <- names(all_estimates) %>%
  lapply(
    function(x) {
      if (length(all_estimates[[x]]) > 0) {
        tree_eps <- unlist(all_estimates_trees[[x]]$summary[1,c(4,2,6)])
        lapply(1:length(all_estimates[[x]]),function(y) {
          y <- as.numeric(y)
          if (!is.null(all_estimates[[x]][[y]]$non_linear_estimate)) {
            pars <- all_estimates[[x]][[y]]$non_linear_estimate$m$getAllPars()
            tao <- all_estimates[[x]][[y]]$mutation_timing
            cp <- all_estimates[[x]][[y]]$change_point_regression$par
            pop_size_at_tao_cp <- exp((800 - tao)*cp[2])
            earliest <- all_estimates[[x]][[y]]$earliest_growth$coefficients[2]
            latest <- all_estimates[[x]][[y]]$latest_growth$coefficients[2]
            pars_vaf <- all_estimates[[x]][[y]]$non_linear_estimate_vaf$m$getPars()
            O <- data.frame(
              l = all_estimates[[x]][[y]]$linear_estimate$coefficients[2],
              nl = 1 / pars[3],
              nl_asym = pars[1],
              nl_xmid = pars[2],
              l_nw = all_estimates[[x]][[y]]$linear_estimate_no_weights$coefficients[2],
              cp_1 = cp[2],
              cp_2 = cp[3],
              earliest = earliest,
              latest = latest,
              expected = 1/pars_vaf[2],
              changepoint_bnpr = cp[4],
              time_at_onset_low = tao[1],
              time_at_onset_high = tao[2],
              w_sum = mean(all_estimates[[x]][[y]]$data$W,na.rm=T),
              pop_at_detection_low = pop_size_at_tao_cp[2],
              pop_at_detection_high = pop_size_at_tao_cp[1],
              tree_eps_low = tree_eps[1],
              tree_eps_mean = tree_eps[2],
              tree_eps_high = tree_eps[3],
              point_of_saturation = which(
                c(
                  predict(
                    all_estimates[[x]][[y]]$non_linear_estimate,
                    newdata = data.frame(X = seq(1,800)))) > (0.9 * pars[1])) %>% min,
              name = x,
              clade_size = length(all_estimates[[x]][[y]]$clade),
              driver = all_estimates[[x]][[y]]$driver,
              clade = y)
            return(O)
          }
        }) %>%
          do.call(what = rbind)
      } else {
        return(NULL)
      }
    }
    ) %>%
  do.call(what = rbind) %>%
  mutate(fitness = as.numeric(gsub('_',"",str_match(name,"_[0-9.]+_")))) %>%
  group_by(name) %>%
  mutate(N = length(unique(clade))) %>%
  mutate(cp_late = cp_1 + cp_2)

all_bnpr_fits_coalescence <- names(all_estimates_coalescence) %>%
  lapply(
    function(x) {
      if (length(all_estimates_coalescence[[x]]) > 0) {
        lapply(1:length(all_estimates_coalescence[[x]]),function(y) {
          y <- as.numeric(y)
          if (!is.null(all_estimates_coalescence[[x]][[y]]$change_point_regression)) {
            cp <- all_estimates_coalescence[[x]][[y]]$change_point_regression$par
            O <- data.frame(
              l = all_estimates_coalescence[[x]][[y]]$linear_estimate$coefficients[2],
              l_nw = all_estimates_coalescence[[x]][[y]]$linear_estimate_no_weights$coefficients[2],
              cp_1 = cp[2],
              cp_2 = cp[3],
              changepoint_bnpr = cp[4],
              w_sum = mean(all_estimates_coalescence[[x]][[y]]$data$W,na.rm=T),
              name = x,
              clade_size = length(all_estimates_coalescence[[x]][[y]]$clade),
              driver = all_estimates_coalescence[[x]][[y]]$driver,
              clade = y)
            return(O)
          }
        }) %>%
          do.call(what = rbind)
      } else {
        return(NULL)
      }
    }
    ) %>%
  do.call(what = rbind) %>%
  mutate(fitness = as.numeric(gsub('_',"",str_match(name,"_[0-9.]+_")))) %>%
  group_by(name) %>%
  mutate(N = length(unique(clade))) %>%
  mutate(cp_late = cp_1 + cp_2)

all_bnpr_fits_trimmed_5 <- names(all_estimates_trimmed_5_years) %>%
  lapply(
    function(x) {
      if (length(all_estimates[[x]]) > 0) {
        lapply(1:length(all_estimates_trimmed_5_years[[x]]),function(y) {
          y <- as.numeric(y)
          nle <- all_estimates_trimmed_5_years[[x]][[y]]$non_linear_estimate
          if (!is.null(nle) & class(nle) != "try-error") {
            pars <- all_estimates_trimmed_5_years[[x]][[y]]$non_linear_estimate$m$getAllPars()
            cp <- all_estimates_trimmed_5_years[[x]][[y]]$change_point_regression$par
            earliest <- all_estimates_trimmed_5_years[[x]][[y]]$earliest_growth$coefficients[2]
            latest <- all_estimates_trimmed_5_years[[x]][[y]]$latest_growth$coefficients[2]
            O <- data.frame(
              l = all_estimates_trimmed_5_years[[x]][[y]]$linear_estimate$coefficients[2],
              nl = 1 / pars[3],
              nl_asym = pars[1],
              nl_xmid = pars[2],
              l_nw = all_estimates_trimmed_5_years[[x]][[y]]$linear_estimate_no_weights$coefficients[2],
              cp_1 = cp[2],
              cp_2 = cp[3],
              earliest = earliest,
              latest = latest,
              changepoint_bnpr = cp[4],
              w_sum = mean(all_estimates_trimmed_5_years[[x]][[y]]$data$W,na.rm=T),
              point_of_saturation = which(
                c(
                  predict(
                    all_estimates_trimmed_5_years[[x]][[y]]$non_linear_estimate,
                    newdata = data.frame(X = seq(1,800)))) > (0.9 * pars[1])) %>% min,
              name = x,
              clade_size = length(all_estimates_trimmed_5_years[[x]][[y]]$clade),
              driver = all_estimates_trimmed_5_years[[x]][[y]]$driver,
              clade = y)
            return(O)
          }
        }) %>%
          do.call(what = rbind)
      } else {
        return(NULL)
      }
    }
    ) %>%
  do.call(what = rbind) %>%
  mutate(fitness = as.numeric(gsub('_',"",str_match(name,"_[0-9.]+_")))) %>%
  group_by(name) %>%
  mutate(N = length(unique(clade))) %>%
  mutate(cp_late = cp_1 + cp_2)

all_bnpr_fits_trimmed <- names(all_estimates_trimmed_10_years) %>%
  lapply(
    function(x) {
      if (length(all_estimates[[x]]) > 0) {
        lapply(1:length(all_estimates_trimmed_10_years[[x]]),function(y) {
          y <- as.numeric(y)
          nle <- all_estimates_trimmed_10_years[[x]][[y]]$non_linear_estimate
          if (!is.null(nle) & class(nle) != "try-error") {
            pars <- all_estimates_trimmed_10_years[[x]][[y]]$non_linear_estimate$m$getAllPars()
            cp <- all_estimates_trimmed_10_years[[x]][[y]]$change_point_regression$par
            earliest <- all_estimates_trimmed_10_years[[x]][[y]]$earliest_growth$coefficients[2]
            latest <- all_estimates_trimmed_10_years[[x]][[y]]$latest_growth$coefficients[2]
            O <- data.frame(
              l = all_estimates_trimmed_10_years[[x]][[y]]$linear_estimate$coefficients[2],
              nl = 1 / pars[3],
              nl_asym = pars[1],
              nl_xmid = pars[2],
              l_nw = all_estimates_trimmed_10_years[[x]][[y]]$linear_estimate_no_weights$coefficients[2],
              cp_1 = cp[2],
              cp_2 = cp[3],
              earliest = earliest,
              latest = latest,
              changepoint_bnpr = cp[4],
              w_sum = mean(all_estimates_trimmed_10_years[[x]][[y]]$data$W,na.rm=T),
              point_of_saturation = which(
                c(
                  predict(
                    all_estimates_trimmed_10_years[[x]][[y]]$non_linear_estimate,
                    newdata = data.frame(X = seq(1,800)))) > (0.9 * pars[1])) %>% min,
              name = x,
              clade_size = length(all_estimates_trimmed_10_years[[x]][[y]]$clade),
              driver = all_estimates_trimmed_10_years[[x]][[y]]$driver,
              clade = y)
            return(O)
          }
        }) %>%
          do.call(what = rbind)
      } else {
        return(NULL)
      }
    }
    ) %>%
  do.call(what = rbind) %>%
  mutate(fitness = as.numeric(gsub('_',"",str_match(name,"_[0-9.]+_")))) %>%
  group_by(name) %>%
  mutate(N = length(unique(clade))) %>%
  mutate(cp_late = cp_1 + cp_2)
```

## Extracting coefficients from driver fits from the original WF simulations

```{r}
all_driver_fits_df <- names(all_driver_fits) %>% 
  lapply(function(x) {
    if (length(all_driver_fits[[x]]) > 0) {
      lapply(names(all_driver_fits[[x]]),function(y) {
        pars <- all_driver_fits[[x]][[y]]$non_linear_estimate$m$getAllPars()
        earliest <- all_driver_fits[[x]][[y]]$earliest_driver$coefficients[2]
        latest <- all_driver_fits[[x]][[y]]$latest_driver$coefficients[2]
        
        early_cp <- all_driver_fits[[x]][[y]]$change_point_regression$par[2]
        
        m <- all_driver_fits[[x]][[y]]$data %>%
          subset(Gen == min(Gen))
        v <- m$Count / 2e5
        time <- m$Gen
        m <- -log((1 - v)/v) - early_cp * time
        expected_last_vaf <- 1 / (1 + exp(-(m + early_cp * 800)))
        return(
          data.frame(
            xmid = pars[1],
            growth_rate = 1/pars[2],
            early_growth = all_driver_fits[[x]][[y]]$change_point_regression$par[2],
            late_growth = all_driver_fits[[x]][[y]]$change_point_regression$par[3],
            earliest_driver = earliest,
            latest_driver = latest,
            changepoint_wf = all_driver_fits[[x]][[y]]$change_point_regression$par[4],
            point_of_saturation_wf = which(
              c(
                predict(
                  all_driver_fits[[x]][[y]]$non_linear_estimate,
                  newdata = data.frame(Gen = seq(1,800)))) > (0.9 * pars[1])) %>% min,
            name = x,
            clade = y,
            last_vaf = all_driver_fits[[x]][[y]]$last_vaf,
            expected_last_vaf,
            gen_at_onset = all_driver_fits[[x]][[y]]$gen_at_onset
          )
        )
      }) %>%
        do.call(what = rbind)
    }
  }) %>%
  do.call(what = rbind) %>%
  mutate(late_growth = early_growth + late_growth)
```

# Comparing BNPR trajectory models

## Plotting BNPR trajectories

Here, we plot all of the inferred BNPR trajectories for each clade. 

```{r,fig.height=2,fig.width=3}
all_fits %>%
  do.call(what = rbind) %>%
  mutate(fitness = str_match(obj_name,"[0-9.]+")) %>% 
  group_by(n,obj_name) %>%
  mutate(V = mean(w)) %>%
  mutate(x = x - min(x)) %>% 
  subset(V <= 5) %>% 
  mutate(fitness = factor(
    fitness,
    levels = seq(0.005,0.03,by = 0.005),
    labels = sprintf("s=%s",seq(0.005,0.03,by = 0.005)))) %>% 
  ggplot(aes(x = x,y = y)) + 
  geom_line(aes(group = n),alpha = 0.25,size = 0.25) + 
  geom_hline(yintercept = 2e5,size = 0.25) +
  facet_wrap(~ fitness) + 
  scale_y_continuous(trans = 'log10',expand = c(0,0),breaks = c(1e-2,1,1e2,1e4,1e6)) +
  scale_x_continuous(expand = c(0,0)) +
  theme_gerstung(base_size = 6) + 
  theme(strip.text = element_text(margin = margin(b = 0.5))) +
  ylab("Effective population size (EPS)") + 
  xlab("Time since first coal. (generations)") + 
  coord_cartesian(ylim = c(0.5,1e7)) +
  ggsave("figures/simulations/trajectories_BNPR.pdf",height = 1.5,width = 2.3)
```

## Comparing BNPR and WF trajectories

Here we scale the BNPR trajectories so that they match as closely as possible those obtained from the WF trajectories. A fairly good fit between BNPR and WF trajectories is observable for a large fraction of trajectories.

```{r,fig.height=6,fig.width=7}
simulated_trajectories <- list()
bnpr_estimates_ <- list()
for (x in names(all_driver_fits)) {
  if (length(all_driver_fits[[x]]) > 0) {
    for (i in 1:length(all_driver_fits[[x]])) {
      driver_fit <- all_driver_fits[[x]][[i]]$data
      bnpr_sum <- all_estimates[[x]][[i]]$bnpr$summary
      W <- (log(bnpr_sum$quant0.975) - log(bnpr_sum$quant0.025))^2/16
      bnpr_sum$time <- (800 - bnpr_sum$time)
      driver_fit$id <- x
      driver_fit$clade <- i
      bnpr_sum$id <- x
      bnpr_sum$clade <- i
      simulated_trajectories[[length(simulated_trajectories)+1]] <- driver_fit
      bnpr_estimates_[[length(bnpr_estimates_)+1]] <- bnpr_sum
      }
  }
}

# for (x in names(all_driver_fits)) {
#   if (x %in% names(all_estimates)) {
#     for (y in names(all_driver_fits[[x]])) {
#       clade_no <- 1
#       bnpr_sums <- list()
#       largest_clade <- which.max(unlist(lapply(all_estimates[[x]],function(x) length(x$clade))))[1]
#       for (estimate in all_estimates[[x]]) {
#         if (y == estimate$driver) {
#           driver_fit <- all_driver_fits[[x]][[y]]$data
#           bnpr_sum <- estimate$bnpr$summary
#           W <- (log(bnpr_sum$quant0.975) - log(bnpr_sum$quant0.025))^2/16
#           bnpr_sum$time <- (800 - bnpr_sum$time)
#           if (clade_no == largest_clade) {
#             opt_data <- data.frame(
#               a = approx(bnpr_sum$time,y = bnpr_sum$mean,xout = driver_fit$Gen)$y,
#               b = driver_fit$Count) %>%
#               na.omit
#             opt_fn <- function(pars) {
#               l <- (opt_data$a * pars[1] - opt_data$b)^2
#               l <- l
#               return(sum(l))
#             }
#             opt_solution <- optim(c(m = 1),
#                                   opt_fn,
#                                   control = list(maxit = 10000),
#                                   method = "Nelder-Mead")
#             P <- opt_solution$par
#           }
#           driver_fit$id <- x
#           driver_fit$clade <- y
#           bnpr_sum$id <- x
#           bnpr_sum$clade <- y
#           bnpr_sum$clade <- clade_no
#           simulated_trajectories[[length(simulated_trajectories)+1]] <- driver_fit
#           bnpr_sums[[length(bnpr_sums)+1]] <- bnpr_sum
#           clade_no <- clade_no + 1
#         }
#       }
#       for (bnpr_sum in bnpr_sums) {
#         bnpr_sum$transformed_y <- bnpr_sum$mean * P[1]
#         bnpr_sum$transformed_y_low <- bnpr_sum$quant0.025 * P[1]
#         bnpr_sum$transformed_y_high <- bnpr_sum$quant0.975 * P[1]
#         bnpr_estimates_[[length(bnpr_estimates_)+1]] <- bnpr_sum
#       }
#     }
#   }
# }

simulated_trajectories_df <- do.call(rbind,simulated_trajectories)
bnpr_estimates_df <- do.call(rbind,bnpr_estimates_) %>% 
  mutate(w = (log(quant0.975) - log(quant0.025))^2/16) %>% 
  group_by(id, clade) %>% 
  mutate(w_sum = mean(w)) %>%
  filter(w_sum < 5)

ggplot(data = NULL) +
  geom_hline(yintercept = 2e5,size = 0.25,colour = "goldenrod",alpha = 0.7) +
  geom_line(data = simulated_trajectories_df,
            aes(x = Gen/100,y = Count,colour = "Simulated",
                group = paste(clade)),
            size = 0.25,
            alpha = 0.7) + 
  geom_line(data = bnpr_estimates_df,aes(x = time/100,y = mean,
                                         group = paste(clade),colour = "BNPR estimate"),
            size = 0.25,
            alpha = 0.5) + 
  facet_wrap(~ id,scales = "free") + 
  theme_gerstung(base_size = 6) + 
  theme(strip.text = element_blank(),legend.position = "bottom") + 
  scale_y_continuous(trans = 'log10') + 
  coord_cartesian(ylim = c(NA,2e8),c(0,8)) +
  scale_colour_manual(values = c(Simulated = "grey50",`BNPR estimate` = "orchid"),name = NULL) + 
  xlab("Generation (x100)") + 
  ggsave("figures/simulations/trajectories_wf_bnpr.pdf",height = 6,width = 7)
```

```{r,fig.height=2.2,fig.width=3}
set.seed(42)
S <- sample(unique(paste(bnpr_estimates_df$id,bnpr_estimates_df$clade)),
            size = 23,replace = F)

bnpr_ss <- subset(bnpr_estimates_df,paste(id,clade) %in% S)
sim_ss <- subset(simulated_trajectories_df,paste(id,clade) %in% S)

similarity <- lapply(S,function(x) {
  a <- bnpr_ss %>%
    subset(paste(id,clade) == x) %>%
    mutate(age_group = floor(time / 100)) %>%
    group_by(age_group) %>%
    summarise(av_neff = mean(mean))
  b <- sim_ss %>%
    subset(paste(id,clade) == x) %>%
    mutate(age_group = floor(Gen / 100)) %>%
    group_by(age_group,id,clade) %>%
    summarise(av_n = mean(Count))
  merge(a,b,by = "age_group") %>%
    group_by(id,clade) %>%
    summarise(mae = mean(abs(av_neff - av_n))) %>%
    return
}) %>%
  do.call(what = rbind) %>%
  arrange(mae) %>%
  mutate(b = paste(id,clade))

ggplot(data = NULL) +
  geom_hline(yintercept = 2e5,size = 0.25,colour = "goldenrod",alpha = 0.7) +
  geom_line(data = sim_ss,
            aes(x = Gen/100,y = Count,colour = "Simulated",fill = "Simulated",
                group = paste(clade)),
            size = 0.25,
            alpha = 0.9) + 
  geom_ribbon(data = bnpr_ss,
              aes(x = time/100,ymin = quant0.025,ymax = quant0.975,
                  group = paste(clade),fill = "BNPR estimate"),
              alpha = 0.2,
              size = 0) +
  geom_line(data = bnpr_ss,
            aes(x = time/100,y = mean,
                group = paste(clade),colour = "BNPR estimate",fill = "BNPR estimate"),
            size = 0.25,
            alpha = 0.7) + 
  facet_wrap(~ factor(paste(id,clade),levels = similarity$b),scales = "free") + 
  theme_gerstung(base_size = 6) + 
  theme(strip.text = element_blank(),legend.position = "bottom",
        legend.key.size = unit(0.2,"cm"),legend.box.spacing = unit(0,"cm")) + 
  scale_y_continuous(trans = 'log10') + 
  coord_cartesian(ylim = c(1,2e8),xlim = c(0,8)) +
  scale_colour_manual(values = c(Simulated = "black",`BNPR estimate` = "orchid"),name = NULL) + 
  scale_fill_manual(values = c(Simulated = "black",`BNPR estimate` = "orchid"),name = NULL) + 
  xlab("Generation (x100)") + 
  ggsave("figures/simulations/trajectories_wf_bnpr_subset.pdf",height = 2.2,width = 3)
```

## Sensitivity analysis - normal vs. trimmed

Here we see if using trimmed leads to very different estimates. Trimming the data by 5 or even 10 years does not lead to very different estimations for the late growth rate as we propose here. Indeed, the average difference we can expect between trimmed-by-10-years and non-trimmed data is 0.8% growth per generation for late growth estimates and 0.4% growth per generation for early growth estimates.

```{r,fig.height=1,fig.width=2}
all_fits_df <- all_fits %>%
  do.call(what = rbind) %>%
  mutate(fitness = str_match(obj_name,"[0-9.]+")) %>% 
  group_by(n,obj_name) %>%
  mutate(V = mean(w)) %>%
  mutate(x = x - min(x)) %>% 
  subset(V <= 10) %>% 
  mutate(fitness = factor(
    fitness,
    levels = seq(0.005,0.03,by = 0.005),
    labels = sprintf("s=%s",seq(0.005,0.03,by = 0.005))))

diff_fits <- rbind(
  all_bnpr_fits %>%
    select(name,clade,early = cp_1,late = cp_late,earliest,latest,w_sum) %>%
    mutate(Preprocessing = "Original trajectory"),
  all_bnpr_fits_trimmed_5 %>%
    select(name,clade,early = cp_1,late = cp_late,earliest,latest,w_sum) %>%
    mutate(Preprocessing = "Trimmed by 5 years"),
  all_bnpr_fits_trimmed %>%
    select(name,clade,early = cp_1,late = cp_late,earliest,latest,w_sum) %>%
    mutate(Preprocessing = "Trimmed by 10 years")
) %>%
  mutate(Preprocessing = factor(Preprocessing,
                                levels = c("Original trajectory",
                                           "Trimmed by 5 years",
                                           "Trimmed by 10 years"))) %>%
  group_by(clade,name) %>% 
  filter(all(w_sum < 5))

df_benchmark <- rbind(
  diff_fits %>% 
  select(late,Preprocessing,name,clade) %>%
  spread(value = "late",key = "Preprocessing") %>% 
  ungroup %>% 
  summarise(
    `Original-5 years` = (`Original trajectory` - `Trimmed by 5 years`),
    `Original-10 years` = (`Original trajectory` - `Trimmed by 10 years`),
    `5 years-10 years` = (`Trimmed by 5 years` - `Trimmed by 10 years`)) %>%
    mutate(id = "Changepoint late"),
  diff_fits %>% 
    select(early,Preprocessing,name,clade) %>%
    spread(value = "early",key = "Preprocessing") %>% 
    ungroup %>% 
    summarise(
      `Original-5 years` = (`Original trajectory` - `Trimmed by 5 years`),
      `Original-10 years` = (`Original trajectory` - `Trimmed by 10 years`),
      `5 years-10 years` = (`Trimmed by 5 years` - `Trimmed by 10 years`)) %>%
    mutate(id = "Changepoint early"),
  diff_fits %>% 
    select(latest,Preprocessing,name,clade) %>%
    spread(value = "latest",key = "Preprocessing") %>% 
    ungroup %>% 
    summarise(
      `Original-5 years` = (`Original trajectory` - `Trimmed by 5 years`),
      `Original-10 years` = (`Original trajectory` - `Trimmed by 10 years`),
      `5 years-10 years` = (`Trimmed by 5 years` - `Trimmed by 10 years`)) %>%
    mutate(id = "Last 150 generations"),
  diff_fits %>% 
    select(earliest,Preprocessing,name,clade) %>%
    spread(value = "earliest",key = "Preprocessing") %>% 
    ungroup %>% 
    summarise(
      `Original-5 years` = (`Original trajectory` - `Trimmed by 5 years`),
      `Original-10 years` = (`Original trajectory` - `Trimmed by 10 years`),
      `5 years-10 years` = (`Trimmed by 5 years` - `Trimmed by 10 years`)) %>%
    mutate(id = "First 150 generations")) 

df_benchmark %>%
  group_by(id) %>%
  summarise(`Original-5 years` = mean(abs(`Original-5 years`),na.rm=T),
            `Original-10 years` = mean(abs(`Original-10 years`),na.rm=T),
            `5 years-10 years` = mean(abs(`5 years-10 years`),na.rm=T))

df_benchmark %>% 
  gather(key = "key",value = "value",-id) %>% 
  subset(id %in% c("Changepoint late","Changepoint early")) %>%
  subset(grepl("Original",key)) %>% 
  ggplot(aes(x = key, y = value,colour = id)) + 
  geom_jitter(size = 0.25,position = position_dodge(width = 1),alpha = 0.25) +
  geom_boxplot(outlier.alpha = 0,size = 0.25,position = position_dodge(width = 1)) + 
  theme_gerstung(base_size = 6) + 
  coord_flip() + 
  theme(legend.position = "bottom") + 
  scale_colour_aaas(name = NULL) +
  xlab("") + 
  ylab("Growth rate estimate") + 
  guides(colour = guide_legend(nrow = 2)) + 
  theme(legend.key.height = unit(0,"cm"),
        legend.box.spacing = unit(0,"cm")) + 
  scale_y_continuous(limits = c(NA,0)) +
  ggsave("figures/simulations/benchmark_trim.pdf",height = 1,width = 2)
```

## Comparing inference of early and late growth between BNPR fits and driver fits from WF simulations

We define here a measure of excessive observed variance in the BNPR trajectory - whenever the average log-EPS variance is larger than 10 we define these trajectories as "high variance" trajectories, noting that they are trajectories which were obtained from fairly small clades (fewer than 10 tips).

```{r,fig.height=1.5,fig.width=1.5}
bnpr_wf_merge_df <- merge(all_bnpr_fits,all_driver_fits_df,by = c("name","clade")) %>%
  mutate(vaf_at_detection_low = pop_at_detection_low/2e5,
         vaf_at_detection_high = pop_at_detection_high/2e5) %>%
  mutate(vaf_at_detection_low = ifelse(vaf_at_detection_low > 1,1,vaf_at_detection_low),
         vaf_at_detection_high = ifelse(vaf_at_detection_high > 1,1,vaf_at_detection_high)) %>%
  mutate(vaf_contained = last_vaf > vaf_at_detection_low & last_vaf < vaf_at_detection_high) %>%
  mutate(vaf_contained_higher_lower = ifelse(
    vaf_contained == F,
    ifelse(last_vaf > vaf_at_detection_high,"higher","lower"),
    "contained")) %>%
  mutate(col = ifelse(w_sum < 5,"Low variance","High variance")) %>%
  mutate(wf_ratio = late_growth/early_growth,
         bnpr_ratio = cp_late/cp_1) %>% 
  mutate(bnpr_ratio = ifelse(is.infinite(bnpr_ratio),NA,bnpr_ratio),
         wf_ratio = ifelse(is.infinite(wf_ratio),NA,wf_ratio)) 

bnpr_wf_merge_df %>% 
  ggplot(aes(x = w_sum)) +
  geom_density() +
  scale_x_continuous(trans = 'log10') + 
  theme_gerstung(base_size = 6) + 
  xlab("Average variance")

ggplot(bnpr_wf_merge_df) +
  geom_boxplot(aes(x = col,y = clade_size),outlier.size = 0.5) + 
  scale_y_continuous(breaks = seq(0,100,by = 10)) + 
  theme_gerstung(base_size = 6) + 
  xlab("Trajectory type") + 
  ylab("Clade size")
```

## Comparing the expected and observed late growth 

We also want to have a measure of growth deceleration that is more comparable with our inferences from the time series data - based on a sigmoidal fit with a fixed carrying capacity ($VAF = 0.5$) that is then extrapolated to its onset of origin. As such, we fit a sigmoidal to the BNPR that assumes a total carrying capacity of $1$ and that the final EPS corresponds to a proxy for the VAF that is parametrised as $\hat{VAF} = \frac{Number\ Of\ Tips\ In\ The\ Clade}{Total\ Tips\ In\ The\ Tree}$. 

```{r,fig.height=1.5,fig.width=1.5}
expected_late_growth <- function(vaf,id,clade) {
  clade <- as.numeric(clade)
  if (clade <= length(all_estimates_coalescence[[id]])) {
    
    tmp_df <- all_estimates_coalescence[[id]][[clade]]$data %>%
      arrange(X)
    last_eps <- tmp_df$Y_exp[nrow(tmp_df)]
    last_eps <- max(tmp_df$Y_exp)
    tmp_df$normalized_Y <- tmp_df$Y_exp / last_eps * vaf

    nl_fit <- nlsLM(formula = normalized_Y ~ SSlogis(X,1,Xmid,b),
                    start = list(Xmid = mean(tmp_df$X),b = 100),
                    control = nls.control(maxiter = 1000,minFactor = 1/(1024^32),
                                          warnOnly = T),
                    data = tmp_df)
    return(nl_fit$m$getAllPars()[2])
  } else {
    return(NA)
  }
}

bnpr_wf_merge_df$expected_nl <- rep(NA,nrow(bnpr_wf_merge_df))

for (i in 1:nrow(bnpr_wf_merge_df)) {
  sub_bnpr_wf_merge_df <- bnpr_wf_merge_df[i,]
  last_vaf <- sub_bnpr_wf_merge_df$last_vaf
  name <- as.character(sub_bnpr_wf_merge_df$name)
  clade <- sub_bnpr_wf_merge_df$clade
  bnpr_wf_merge_df$expected_nl[i] <- 1/expected_late_growth(last_vaf,name,clade)
}

bnpr_wf_merge_df %>% 
  subset(col == "Low variance") %>% 
  mutate(expected_late_ratio_wf = late_growth/growth_rate,
         expected_late_ratio = cp_late / expected_nl) %>%
  mutate(expected_late_ratio = ifelse(expected_late_ratio < -1,-1,expected_late_ratio)) %>% 
  ggplot(aes(x = ifelse(expected_late_ratio_wf < 0.8,"Saturating","Near constant"),
             y = expected_late_ratio)) +
  geom_hline(yintercept = 0.5) +
  geom_jitter(width = 0.3,size = 0.5,alpha = 0.8) +
  geom_boxplot(outlier.alpha = 0,alpha = 0.8) + 
  theme_gerstung(base_size = 6) +
  scale_y_continuous(breaks = c(-1,0,1,2),
                     labels = c("<-1","0","1","2")) + 
  ylab("BNPR late/expected growth ratio") +
  xlab("")
```

```{r,fig.height=1.5,fig.width=0.7}
bnpr_wf_merge_df %>% 
  subset(col == "Low variance") %>% 
  mutate(expected_late_ratio = cp_late / expected_nl) %>% 
  mutate(expected_late_ratio = ifelse(expected_late_ratio < -1,-1,expected_late_ratio)) %>% 
  ggplot(aes(y = expected_late_ratio,x=0)) + 
  geom_jitter(size = 0.5,alpha = 0.5,width = 0.05) + 
  geom_boxplot(size = 0.25,width = 0.1,outlier.size = 0,alpha = 0.8) + 
  theme_gerstung(base_size = 6) + 
  geom_hline(yintercept = 1,linetype = 2) +
  #coord_cartesian(ylim = c(0,5)) + 
  scale_y_continuous(breaks = c(-1,0,1,2),
                     labels = c("<-1","0","1","2")) + 
  ylab("Observed/expected late growth ratio") +
  xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  ggsave("figures/simulations/expected_late_growth_ratio.pdf",height = 1.5,width = 0.7)
```

## Benchmarking different models for BNPR fits

Finally we compare all of the methods specified above and make statements regarding their usability in predicting the true fitness of each clade. We note that log-linear fits have a tendency to underestimate the growth rate of clones with large fitness values, whereas sigmoidal curves have a tendency to do the opposite - overestimate the growth rate of clones with small fitness values. The early growth rate stands as the most adequate solution for our problem, with $R^2 > 0.3$ for fits with low variance. We must note that we still observe an $RMSE = 0.01$ which would, assuming 10 generations per year, correspond to a variation of $\pm 10%$. As such, these inferences should account for this considerable level of uncertainty that is associated with each estimate.

```{r,fig.height=5,fig.width=4.7}
fits <- all_bnpr_fits
fits <- fits %>% 
  mutate(col = ifelse(w_sum < 5,"Low variance","High variance")) 

limits <- c(0,0.05)
plot_list <- list()
metric_list <- list()

for (col_name in c("l","cp_1","cp_late","nl")) {
  sub_data <- fits %>%
    ungroup() %>%
    select(fitness,R = col_name) %>%
    na.omit()
  sub_data_low <- fits %>%
    subset(col == "Low variance") %>%
    ungroup() %>%
    select(fitness,R = col_name) %>%
    na.omit() 
  
  plot_list[[col_name]] <- fits %>%
    select(fitness,R = col_name,col = col) %>% 
    ggplot(aes(x = fitness,y = R,group = paste(fitness),colour = col)) +
    geom_jitter(width = 0.001,size = 0.5) +
    geom_boxplot(outlier.alpha = 0,size = 0.25,alpha = 0.7,colour = "black") +
    theme_gerstung(base_size = 6) +
    coord_cartesian(xlim = limits,
                    ylim = limits) +
    geom_abline(slope = 1) +
    ylab("Inferred growth") +
    xlab("Simulated growth") + 
    scale_colour_lancet(guide = F)
  
  metric_list[[col_name]] <- data.frame(
    method = col_name,
    s = c("All trajectories","Low variance"),
    R = c(cor(sub_data$fitness,sub_data$R),
          cor(sub_data_low$fitness,sub_data_low$R)),
    mse = c(mean((sub_data$fitness - sub_data$R)^2),
            mean((sub_data_low$fitness - sub_data_low$R)^2))
  )
}

metric_df <- do.call(rbind,metric_list) %>%
  mutate(
    
    met = list(l = "Linear\ngrowth",
               cp_1 = "Early linear\ngrowth",
               cp_late = "Late linear\ngrowth",
               nl = "Sigmoidal\ngrowth")[method] %>%
      unlist
  )

r2_plot <- ggplot(data = metric_df,aes(x = reorder(met,-R^2),y = R^2,fill = s)) + 
  geom_bar(stat = "identity",position = "dodge") + 
  xlab("") + 
  ylab("R2") + 
  theme_gerstung(base_size = 6) +
  coord_flip() + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_fill_lancet(guide = F)

mse_plot <- ggplot(data = metric_df,aes(x = reorder(met,-mse),y = sqrt(mse),fill = s)) + 
  geom_bar(stat = "identity",position = "dodge") + 
  xlab("") + 
  ylab("RMSE") + 
  theme_gerstung(base_size = 6) +
  coord_flip() + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_fill_lancet(guide = F)

plot_grid(
  plot_list$l + ggtitle("Linear growth"),
  plot_list$cp_1 + ggtitle("Early linear growth"),
  plot_list$cp_late + ggtitle("Late linear growth"),
  plot_list$nl + ggtitle("Sigmoidal growth"),
  r2_plot,
  mse_plot,
  rel_heights = c(1,1,0.6),
  ncol = 2) + 
  ggsave("figures/simulations/validation_BNPR.pdf",height = 5,width = 4.7,useDingbats=FALSE)

write.csv(bnpr_wf_merge_df,"data_output/simulated_bnpr_coefficients.csv",row.names = F)
```