---
title: "The Natural History of Clonal Haematopoiesis"
subtitle: "Investigating different methods to estimate EPS"
author: 
    - José Guilherme de Almeida
    - Moritz Gerstung
output: 
  html_document:
    theme: lumen
    highlight: tango
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    number_sections: true
---

In this notebook we inspect how different methods to estimate EPS compare.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Setup

```{r}
source("Scripts/vaf_dynamics_functions.R")
source("Scripts/make_ultrametric.R")

library(parallel)
library(Matrix)
library(castor)
library(phangorn)
library(phylodyn)
library(minpack.lm)
INLA:::inla.dynload.workaround()

change_point <- function(x, b0, m1, m2, delta) { 
  b0 + (x*m1) + (sapply(x-delta, function (t) max(0, t)) * m2)
}

plot_tree <- function(obj) { 
  tree_ultra <- obj$tree_ultra
  tree_ultra$edge.length[is.infinite(tree_ultra$edge.length)] <- 0
  tree_ultra$tip.label <- obj$tree$S
  driver_list <- Filter(function(x) length(x) > 5,obj$driver_list)
  if (length(driver_list) > 1) {
    colours <- RColorBrewer::brewer.pal(length(driver_list),"Set3")
  } else {
    colours <- c("black")
  }
  colour_code <- rep("black",length(tree_ultra$tip.label))
  R <- 1:length(driver_list)
  if (length(driver_list) == 0) {
    R <- numeric(0)
  } 
  for (i in R) {
    D <- driver_list[[i]]
    colour_code[tree_ultra$tip.label %in% unique(D)] <- colours[i]
  }
  plot(tree_ultra,tip.color = colour_code)
}

clade_from_mrca <- function(tree,mrca) {
  edge_idx <- which(tree$edge[,1] == mrca)
  continue <- T
  output <- c(which(tree$edge[,2] == mrca))
  while (continue == T) {
    if (length(edge_idx) > 0) {
      output <- c(output,edge_idx)
      new_edge_idx <- c()
      for (idx in edge_idx) {
        nodes <- tree$edge[idx,2]
        new_idxs <- which(tree$edge[,1] == nodes)
        new_edge_idx <- c(new_edge_idx,new_idxs)
      }
      edge_idx <- new_edge_idx
    } else {
      continue <- F
    }
  }
  return(output)
}

build_tree <- function(sub_data,subsample_size=100,
                       detection_threshold=0) {
  sub_af <- sub_data %>% 
    group_by(CloneID) %>% 
    summarise(N = max(NClones),.groups = "drop") %>%
    arrange(CloneID)
  MaxClone <- max(sub_data$CloneID)
  MaxMut <- max(sub_data$MutID)
  sub_af_sparse <- rep(0,MaxClone)
  sub_af_sparse[sub_af$CloneID] <- sub_af$N 
  
  Presence <- sparseMatrix(i = sub_data$CloneID,
                           j = sub_data$MutID,
                           dims = c(MaxClone,MaxMut))
  Presence[cbind(sub_data$CloneID,sub_data$MutID)] <- 1

  if (sum(as.logical(sub_af_sparse)) < subsample_size) {
    sub_af_sparse <- rep(1,length(sub_af_sparse))
  }
  
  S <- sample(MaxClone,subsample_size,replace = F,prob = sub_af_sparse)
  af <- sub_af_sparse[S]
  Presence <- Presence[S,]
  Presence <- as.matrix(rbind(Presence,wt=0))
  Presence <- Presence[,colSums(Presence) > 0]
  
  dst <- as.matrix(dist(Presence > 0,method = "manhattan"))
  tree <- root(njs(dst),
               outgroup = "wt",
               resolve.root = T,
               edgelabel = T) 
  tree <- drop.tip(tree,"wt")
  return(list(tree = tree,S = S,af = af))
}

read_clonex <- function(path,d = 1000) {
  x <- read_tsv(path,
                col_names = c("Gen","NClones","CloneID","MutID"),
                col_types = c(col_integer(),col_integer(),col_integer(),col_integer()),
                progress = F) %>%
    mutate(Driver = MutID <= d)
  return(x)
}

trajectory_from_td <- function(x=0:1000, td=0, s=0.01, N=2e5, g=1){
  s <- s/g
  t0 <- td - 1/(s)
  t <- x-t0
  t <- t*g
  y <- pmax(0,t)
  if (s==0)
    return(y)
  td <- 1/(s)
  te <- log(N*s -1)/s + td
  y[t>td] <- N/(1+exp(-(s*(t[t>td]-te))))
  return(y)
}

trajectory_from_t0 <- function(x=0:1000, t0=0, s=0.01, N=2e5, g=1){
  s <- s/g
  t <- x-t0
  t <- t*g
  y <- pmax(0,t)
  if (s==0)
    return(y)
  td <- 1/(s)
  te <- log(N*s -1)/s + td
  y[t>td] <- N/(1+exp(-(s*(t[t>td]-te))))
  return(y)
}
```

# Population size estimation

In this section we calculate/load all BNPR trajectories for the trees inferred from Wright-Fisher (WF) simulations. These simulations are done for 6 different fitness levels - 0.005, 0.010, 0.015, 0.020, 0.025 and 0.030 - over 800 generations and with a fixed population size (200,000).

## Loading data and inferring trees

The first step is to build the trees, which we then display. One can immediately see that expansions become increasingly prevalent for higher fitness effects, with quite a few cases of a single clone sweeping all of the population.

```{r}
N <- 2e5
N_DRIVERS <- 50
all_file_paths <- list.files(
  "hsc_output_bnpr_complete",pattern = "last_generation",
  full.names = T,recursive = T)
all_driver_file_paths <- list.files(
  "hsc_output_bnpr_complete",pattern = "driver",
  full.names = T,recursive = T)

file_name <- "data_output/simulated_trees_trajectories.RDS"
if (file.exists(file_name) == T) {
  tree_traj <- readRDS(file_name)
  trees <- tree_traj[[1]]
  all_driver_trajectories <- tree_traj[[2]]
} else {
  all_driver_trajectories <- all_driver_file_paths %>%
    lapply(function(x) {
      read.csv(x,header = F) %>%
        arrange(V1,V2) %>%
        select(Gen = V1,MutID = V2,Count = V3)})
  
  trees <- mclapply(
    all_file_paths,
    mc.cores = 6,
    mc.cleanup = T,
    function(path) {
      s <- str_match(path,"hsc_[0-9.]+") %>%
        gsub(pattern = "hsc_",replacement = "") %>%
        as.numeric()
      system(sprintf('echo "%s"', path)) # prints during mclapply by using bash
      x <- read_clonex(path,d = N_DRIVERS) %>%
        subset(Gen == 800) %>%
        subset(MutID != 0)
      x$R <- as.numeric(str_match(path,"[0-9.]+$"))
      x$s <- s
      x$drift_threshold <- 1 / (N * (s))
      
      driver_list <- x %>% 
        subset(Driver == T) %>% 
        select(CloneID,MutID) 
      driver_list_out <- list()
      for (driver_id in unique(driver_list$MutID)) {
        tmp <- driver_list %>%
          subset(MutID == driver_id) %>%
          select(CloneID) %>%
          unlist
        driver_list_out[[as.character(driver_id)]] <- tmp
      }
      tree <- x %>%
        build_tree()
      ultrametric_tree <- make.ultrametric.tree(tree$tree)
      gc()
      list(
        tree = tree,
        tree_ultra = ultrametric_tree,
        driver_list = driver_list_out,
        path = path
      ) %>%
        return
    }
  )
  
  names(trees) <- gsub('/last_generation','',all_file_paths)
  names(all_driver_trajectories) <- gsub('/driver_trajectory','',all_driver_file_paths)
  saveRDS(object = list(trees,all_driver_trajectories),file = file_name)
}
```

```{r}
par(mfrow = c(4,5),mar = c(2,0.5,1,0.5))
for (tree_name in names(trees)) {
  tree <- trees[[tree_name]]
  fitness <- str_match(tree_name,"[0-9.]+")
  plot(tree$tree_ultra,main = fitness)
}
```

## Calculating BNPR trajectories for the whole trees

Here we calculate the actual EPS trajectories using BNPR and display them.

```{r}
file_name <- "data_output/simulated_bnpr_trees.RDS"
if (file.exists(file_name) == T) {
  all_estimates_trees <- readRDS(file_name)
} else {
  all_estimates_trees <- lapply(
    trees,
    function(x) {
      system(sprintf('echo "%s"', x$path)) # prints during mclapply by using bash
      BNPR(x$tree_ultra) %>%
        return
      }
    )
  saveRDS(all_estimates_trees,file_name)
}
```

## Calculating `mcmc.popsize` and skyline trajectories for the whole trees

We accompany our BNPR estimates with those obtained through `mcmc.popsize` and skyline estimations to prove that what we are observing are not artefacts specific to BNPR. Interestingly, `mcmc.popsize` appears to suffer from its own bias - in some trajectories a very large initial EPS followed by a dip is observable. This holds for both possible priors - constant population size and a skyline-process population size, with the latter ameliorating this effect slightly.

```{r}
all_estimates_trees_skyline <- mclapply(
    trees,
    mc.cores = 2,
    mc.cleanup = T,
    function(x) skyline(x$tree_ultra,epsilon = 0.01)
    )

all_estimates_trees_skyline_wide <- mclapply(
    trees,
    mc.cores = 2,
    mc.cleanup = T,
    function(x) skyline(x$tree_ultra,epsilon = 0.05)
    )

all_estimates_trees_mcmc <- mclapply(
    trees,
    mc.cores = 2, 
    mc.cleanup = T,
    function(x) mcmc.popsize(x$tree_ultra,nstep=10000,thinning = 100,
                             progress.bar = F,
                             burn.in = 1000,lambda = 0.1,
                             method.prior.heights = "constant")
    )

all_estimates_trees_mcmc_skyline <- mclapply(
    trees,
    mc.cores = 2, 
    mc.cleanup = T,
    function(x) mcmc.popsize(x$tree_ultra,nstep=10000,thinning = 100,
                             progress.bar = F,
                             burn.in = 1000,lambda = 0.1,
                             method.prior.heights = "skyline")
    )
```

```{r}
par(mfrow = c(4,5),mar = c(1.5,2,2,1))
for (name in 1:length(all_estimates_trees)) {
  plot_BNPR(all_estimates_trees[[name]])
  lines(all_estimates_trees_skyline[[name]],col = "green")
  lines(all_estimates_trees_skyline_wide[[name]],col = "#026440")
  lines(extract.popsize(all_estimates_trees_mcmc[[name]]),col = "red")
  lines(extract.popsize(all_estimates_trees_mcmc_skyline[[name]]),col = "purple")
}
```

## Calculating BNPR trajectories for each clade

```{r}
file_name <- "data_output/simulated_bnpr_clades.RDS"
if (file.exists(file_name) == T) {
  all_estimates <- readRDS(file_name)
} else {
  all_estimates <- list()
  i <- 1
  for (obj_name in names(trees)) {
    obj <- trees[[obj_name]]
    print(obj_name)
    s <- str_match(obj_name,"[0-9.]+") %>%
      as.numeric
    Tree <- list(tree = obj$tree_ultra)
    Tree$tree$tip.label <- obj$tree$S
    Tree$tree$edge.length[is.infinite(Tree$tree$edge.length)] <- 0
    tree <- Tree$tree
    driver_id_list <- Filter(function(x) length(x) > 5,obj$driver_list)
    all_drivers <- do.call(c,driver_id_list)
    clades <- cut_tree(tree = tree,depth = 0.1) %>%
      Filter(f = function(x) length(x) >= 5)
    clades_ <- list()
    for (clade in clades) {
      clade_tips <- tree$tip.label[clade]
      for (driver in names(driver_id_list)) {
        if (any(clade_tips %in% driver_id_list[[driver]])) {
          clades_[[length(clades_)+1]] <- list(
            clade = clade,
            driver = driver
          )
        }
      }
    }
    clades <- clades_
    
    if (!(obj_name %in% names(all_estimates))) {
      bnpr_estimates <- mclapply(
        clades,
        mc.cores = 2,
        mc.cleanup = T,
        function(clad) {
          #tip_in_clade <- which(Tree$tree$tip.label %in% clad)
          tip_in_clade <- clad$clade
          if (length(tip_in_clade) > 4) {
            sub_tree <- keep.tip(Tree$tree,tip_in_clade) %>%
              multi2di()
            if (!is.null(sub_tree)) {
              bnpr_estimates <- NULL
              bnpr_estimate <- BNPR(sub_tree)
              if (!is.null(bnpr_estimate)) {
                list(
                  tree = tree,
                  bnpr = bnpr_estimate,
                  clade = clad$clade,
                  tree = sub_tree,
                  driver = clad$driver) %>%
                  return
              }
            }
          }
        }
      ) %>%
        Filter(f = function(x) !is.null(x))
      all_estimates[[obj_name]] <- bnpr_estimates
    }
    gc()
  }
  saveRDS(all_estimates,file_name)
}
```

## Calculating `mcmc.popsize` trajectories for each clade

```{r}
all_estimates_mcmc <- list()
i <- 1
par(mfrow = c(4,5),mar=c(1,2,2,1))
for (obj_name in names(all_estimates)) {
  print(obj_name)
  if (!(obj_name %in% names(all_estimates_mcmc))) {
    obj <- all_estimates[[obj_name]]
    estimates <- list()
    for (child_obj in obj) {
      Tr <- child_obj$tree %>%
        keep.tip(child_obj$clade) %>%
        multi2di()
      pr <- ifelse(length(child_obj$tree$tip.labels) > 10,"skyline","constant")
      pop_size_eps <- mcmc.popsize(Tr,10000,100,1000,
                                   lambda = 0.1,
                                   progress.bar = F,
                                   method.prior.heights = pr)
      estimates[[length(estimates)+1]] <- list(pop_size_eps = pop_size_eps)
    }
    estimates <- estimates %>%
     Filter(f = function(x) !is.null(x))
    all_estimates_mcmc[[obj_name]] <- estimates
  }
  gc()
}
```

## Calculating skyline trajectories for each clade

```{r}
all_estimates_skyline <- list()
names_done <- names(all_estimates_skyline)
i <- 1
par(mfrow = c(4,5),mar=c(1,2,2,1))
for (obj_name in names(all_estimates)) {
  print(obj_name)
  if (!(obj_name %in% names_done)) {
    obj <- all_estimates[[obj_name]]
    estimates <- list()
    for (child_obj in obj) {
      Tr <- child_obj$tree %>%
        keep.tip(child_obj$clade) %>%
        multi2di()
      sk <- skyline(Tr,epsilon = 0.02)
      estimates[[length(estimates)+1]] <- list(skyline = sk)
    }
    estimates <- estimates %>% 
      Filter(f = function(x) !is.null(x))
    all_estimates_skyline[[obj_name]] <- estimates
  }
  gc()
}
```

## Comparing BNPR, `mcmc.popsize` and skyline trajectories

### For clades with 10 or fewer tips

```{r}
# fewer than 10 tips per clade
par(mfrow = c(4,5),mar = c(1,1,1,1))
for (name in names(all_estimates)) {
  obj <- all_estimates[[name]]
  obj_mcmc <- all_estimates_mcmc[[name]]
  obj_skyl <- all_estimates_skyline[[name]]
  
  if (length(obj) > 0) {
    for (i in 1:length(obj)) {
      if (length(obj[[i]]$clade) <= 10) {
        bnpr_traj <- obj[[i]]$bnpr
        mcmc_traj <- obj_mcmc[[i]]$pop_size_eps
        skyl_traj <- obj_skyl[[i]]$skyline
        
        plot_BNPR(bnpr_traj)
        try(lines(extract.popsize(mcmc_traj),col = "red"))
        lines(skyl_traj,col = "green")
      }
    }
  }
}
```

### For clades with 10-50 tips

```{r}
# 10-50 tips
par(mfrow = c(4,5),mar = c(1,1,1,1))
for (name in names(all_estimates)) {
  obj <- all_estimates[[name]]
  obj_mcmc <- all_estimates_mcmc[[name]]
  obj_skyl <- all_estimates_skyline[[name]]
  
  if (length(obj) > 0) {
    for (i in 1:length(obj)) {
      if (length(obj[[i]]$clade) < 50 & length(obj[[i]]$clade) > 10) {
        bnpr_traj <- obj[[i]]$bnpr
        mcmc_traj <- obj_mcmc[[i]]$pop_size_eps
        skyl_traj <- obj_skyl[[i]]$skyline
        
        plot_BNPR(bnpr_traj)
        try(lines(extract.popsize(mcmc_traj),col = "red"))
        lines(skyl_traj,col = "green")
      }
    }
  }
}
```

### For clades with 50+ tips

```{r}
# 50-100 clades
par(mfrow = c(4,5),mar = c(1,1,1,1))
for (name in names(all_estimates)) {
  obj <- all_estimates[[name]]
  obj_mcmc <- all_estimates_mcmc[[name]]
  obj_skyl <- all_estimates_skyline[[name]]
  
  if (length(obj) > 0) {
    for (i in 1:length(obj)) {
      if (length(obj[[i]]$clade) >= 50) {
        bnpr_traj <- obj[[i]]$bnpr
        mcmc_traj <- obj_mcmc[[i]]$pop_size_eps
        skyl_traj <- obj_skyl[[i]]$skyline
        
        plot_BNPR(bnpr_traj)
        try(lines(extract.popsize(mcmc_traj),col = "red"))
        # extracting population size fails for a few cases
        lines(skyl_traj,col = "green")
      }
    }
  }
}
```

### Figure with examples

```{r}
set.seed(4242)
samples_to_plot <- sample(names(all_estimates),replace = F,size = 15)

data_for_plot <- samples_to_plot %>%
  lapply(
    function(i) {
      data_list <- list()
      obj <- all_estimates[[i]]
      obj_mcmc <- all_estimates_mcmc[[i]]
      obj_skyl <- all_estimates_skyline[[i]]
      
      if (length(obj) > 0) {
        for (j in 1:length(obj)) {
          try({
            bnpr_traj <- obj[[j]]$bnpr
            mcmc_traj <- obj_mcmc[[j]]$pop_size_eps
            skyl_traj <- obj_skyl[[j]]$skyline
            
            pop_size <- extract.popsize(mcmc_traj) %>%
              as.matrix 
            
            data_list[[j]] <- data.frame(
              X = bnpr_traj$summary$time,
              Y = bnpr_traj$summary$quant0.5,
              Y025 = bnpr_traj$summary$quant0.025,
              Y975 = bnpr_traj$summary$quant0.975,
              MCMC_Y = approx(x = pop_size[,1],
                              y = pop_size[,3],
                              xout = bnpr_traj$summary$time)$y,
              MCMC_Y025 = approx(x = pop_size[,1],
                                 y = pop_size[,4],
                                 xout = bnpr_traj$summary$time)$y,
              MCMC_Y975 = approx(x = pop_size[,1],
                                 y = pop_size[,5],
                                 xout = bnpr_traj$summary$time)$y,
              SKYL_Y = approx(x = skyl_traj$time,
                              y = skyl_traj$population.size,
                              xout = bnpr_traj$summary$time,rule = 2)$y,
              id = i,
              clade_no = j,
              clade_size = length(obj[[j]]$clade)
            )
          })
        }
      }
      return(do.call(rbind,data_list))
    }
  )
```

```{r}
data_for_plot %>%
  do.call(what = rbind) %>%
  mutate(fitness = str_match(id,pattern = "[0-9.]+")) %>%
  mutate(traj_id = as.numeric(as.factor(paste(id,clade_no)))) %>%
  ggplot(aes(x = 800 - X)) + 
  geom_ribbon(aes(ymin = Y025,ymax = Y975,fill = "BNPR"),alpha = 0.3) + 
  geom_line(aes(y = Y,colour = "BNPR"),size = 0.25) + 
  geom_ribbon(aes(ymin = MCMC_Y025,ymax = MCMC_Y975,fill = "mcmc.popsize"),alpha = 0.3) + 
  geom_line(aes(y = MCMC_Y,colour = "mcmc.popsize"),size = 0.25) + 
  geom_line(aes(y = SKYL_Y,colour = "skyline",fill = "skyline"),size = 0.25) + 
  facet_wrap(~ reorder(sprintf("id=%s (N=%s; s=%s)",traj_id, clade_size,fitness),clade_size)) + 
  scale_y_continuous(trans = 'log10') + 
  coord_cartesian(ylim = c(1e-1,1e8)) + 
  scale_colour_aaas(name = NULL) +
  scale_fill_aaas(name = NULL) +
  theme_gerstung(base_size = 6) + 
  theme(strip.text = element_text(margin = margin()),
        legend.position = "bottom",
        legend.key.size = unit(0.1,"cm")) + 
  xlab("Time (generations)") + 
  ylab("Neff") + 
  ggsave(filename = sprintf("figures/simulations/eps_comparison.pdf"),
         useDingbats = F,
         width = 5,height = 3)
```

## Comparing fits derived from BNPR, `mcmc.popsize` and skyline

There is quite good agreement between log-linear trajectories fitted to skyline estimation and BNPR. This does not hold as well when comparing `mcmc.popsize` and BNPR - probably due to the initial dip in population size, the EPS estimates derived from `mcmc.popsize` lead to a slight underestimation of log-linear growth trajectories. As for early/late growth rate comparisons we observe little agreement between BNPR and `mcmc.popsize` or skyline estimation.

```{r,fig.height=2,fig.width=4}
linear_estimates <- list()
early_late_estimates <- list()
idx <- 1
for (x in names(all_estimates)) {
  obj <- all_estimates[[x]]
  obj_mcmc <- all_estimates_mcmc[[x]]
  obj_skyl <- all_estimates_skyline[[x]]
  linear_estimates[[x]] <- list()
  early_late_estimates[[x]] <- list()
  if (length(obj) > 0){
    for (i in 1:length(obj)) {
      try(
        {
          bnpr_data <- data.frame(
            X = (1 - obj[[i]]$bnpr$summary$time),
            Y = log(obj[[i]]$bnpr$summary$quant0.5))
          ps <- extract.popsize(obj_mcmc[[i]]$pop_size_eps)
          X_ <- seq(min(obj_skyl[[i]]$skyline$time),max(obj_skyl[[i]]$skyline$time),length.out = 100)
          sk_Y <- obj_skyl[[i]]$skyline$population.size[as.numeric(cut(X_,c(0,obj_skyl[[i]]$skyline$time)))]
          skyl_data <- data.frame(
            X = (1 - X_),
            Y = log(sk_Y)
            ) %>% 
            na.omit
          mcmc_data <- data.frame(
            X = (1 - ps[,1]),
            Y = log(ps[,3])) %>%
            na.omit
          linear_estimates[[x]][[i]] <- list(
            id = idx,
            bnpr = lm(Y ~ X,data = bnpr_data),
            skyl = lm(Y ~ X,data = skyl_data),
            mcmc = lm(Y ~ X,data = mcmc_data)
          )
          early_late_estimates[[x]][[i]] <- list(
            id = idx,
            bnpr = list(
              late = lm(Y ~ X,data = bnpr_data[1:20,]),
              early = lm(Y ~ X,data = bnpr_data[nrow(bnpr_data) - c(19:0),])),
            skyl = list(
              late = lm(Y ~ X,data = skyl_data[1:20,]),
              early = lm(Y ~ X,data = skyl_data[nrow(skyl_data) - c(19:0),])),
            mcmc = list(
              late = lm(Y ~ X,data = mcmc_data[1:40,]),
              early = lm(Y ~ X,data = mcmc_data[nrow(mcmc_data) - c(39:0),]))
            )
          }
      )
    }
  }
  idx <- idx + 1
}

linear_estimates_bnpr_mcmc_skyl <- linear_estimates %>% 
    lapply(function(x) do.call(rbind,
                               lapply(x, function(y) c(y$id,
                                                       y$bnpr$coefficient[2],
                                                       y$mcmc$coefficient[2],
                                                       y$skyl$coefficient[2])))) %>%
  do.call(what = rbind) %>%
  as.data.frame()
colnames(linear_estimates_bnpr_mcmc_skyl) <- c("id","BNPR","mcmc.popsize","Skyline")
linear_estimates_bnpr_mcmc_skyl$id <- names(all_estimates)[linear_estimates_bnpr_mcmc_skyl$id]
linear_estimates_bnpr_mcmc_skyl$fitness <- linear_estimates_bnpr_mcmc_skyl$id %>%
  str_match("[0-9.]+") %>% 
  as.numeric

early_estimates_bnpr_mcmc_skyl <- early_late_estimates %>% 
    lapply(function(x) do.call(rbind,
                               lapply(x, function(y) c(y$id,
                                                       y$bnpr$early$coefficient[2],
                                                       y$mcmc$early$coefficient[2],
                                                       y$skyl$early$coefficient[2])))) %>%
  do.call(what = rbind) %>%
  as.data.frame()
colnames(early_estimates_bnpr_mcmc_skyl) <- c("id","BNPR","mcmc.popsize","Skyline")
early_estimates_bnpr_mcmc_skyl$id <- names(all_estimates)[early_estimates_bnpr_mcmc_skyl$id]
early_estimates_bnpr_mcmc_skyl$fitness <- early_estimates_bnpr_mcmc_skyl$id %>%
  str_match("[0-9.]+") %>% 
  as.numeric

late_estimates_bnpr_mcmc_skyl <- early_late_estimates %>% 
  Filter(f = function(f) !is.null(x)) %>% 
  lapply(function(x) do.call(rbind,
                             lapply(x, function(y) c(y$id,
                                                     y$bnpr$late$coefficient[2],
                                                     ifelse(!is.null(y$bnpr$late),summary(y$bnpr$late)$coefficients[2,2],NA),
                                                     y$mcmc$late$coefficient[2],
                                                     y$skyl$late$coefficient[2])))) %>%
  do.call(what = rbind) %>%
  as.data.frame()
colnames(late_estimates_bnpr_mcmc_skyl) <- c("id","BNPR","BNPR_se","mcmc.popsize","Skyline")
late_estimates_bnpr_mcmc_skyl$id <- names(all_estimates)[late_estimates_bnpr_mcmc_skyl$id]
late_estimates_bnpr_mcmc_skyl$fitness <- late_estimates_bnpr_mcmc_skyl$id %>%
  str_match("[0-9.]+") %>% 
  as.numeric

A <- linear_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = Skyline,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.05,0.05),
                  ylim = c(-0.05,0.05))

B <- early_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = Skyline,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.05,0.05),
                  ylim = c(-0.05,0.05))

C <- late_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = Skyline,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.05,0.05),
                  ylim = c(-0.05,0.05))

D <- linear_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = mcmc.popsize,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.05,0.05),
                  ylim = c(-0.05,0.05))

E <- early_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = mcmc.popsize,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.05,0.05),
                  ylim = c(-0.05,0.05))

FF <- late_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = mcmc.popsize,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.05,0.05),
                  ylim = c(-0.05,0.05))

plot_grid(A,B,C,D,E,FF,nrow = 2)

data.frame(
  BNPR = c(cor(linear_estimates_bnpr_mcmc_skyl$fitness,linear_estimates_bnpr_mcmc_skyl$BNPR),
           cor(early_estimates_bnpr_mcmc_skyl$fitness,early_estimates_bnpr_mcmc_skyl$BNPR))^2,
  Skyline = c(cor(linear_estimates_bnpr_mcmc_skyl$fitness,linear_estimates_bnpr_mcmc_skyl$Skyline),
              cor(early_estimates_bnpr_mcmc_skyl$fitness,early_estimates_bnpr_mcmc_skyl$Skyline))^2,
  mcmc.popsize = c(cor(linear_estimates_bnpr_mcmc_skyl$fitness,linear_estimates_bnpr_mcmc_skyl$mcmc.popsize),
                   cor(early_estimates_bnpr_mcmc_skyl$fitness,early_estimates_bnpr_mcmc_skyl$mcmc.popsize))^2,
  fit = c("Log-linear","Early log-linear"))
```

```{r,fig.height=1,fig.width=2.2}
data.frame(
  BNPR = c(cor(linear_estimates_bnpr_mcmc_skyl$fitness,linear_estimates_bnpr_mcmc_skyl$BNPR),
           cor(early_estimates_bnpr_mcmc_skyl$fitness,early_estimates_bnpr_mcmc_skyl$BNPR))^2,
  Skyline = c(cor(linear_estimates_bnpr_mcmc_skyl$fitness,linear_estimates_bnpr_mcmc_skyl$Skyline),
              cor(early_estimates_bnpr_mcmc_skyl$fitness,early_estimates_bnpr_mcmc_skyl$Skyline))^2,
  mcmc.popsize = c(cor(linear_estimates_bnpr_mcmc_skyl$fitness,linear_estimates_bnpr_mcmc_skyl$mcmc.popsize),
                   cor(early_estimates_bnpr_mcmc_skyl$fitness,early_estimates_bnpr_mcmc_skyl$mcmc.popsize))^2,
  fit = c("Log-linear","Early log-linear")) %>%
  gather(key = "key",value = "value",BNPR,Skyline,mcmc.popsize) %>%
  ggplot(aes(x = key,y = fit,fill = value)) +
  geom_tile() + 
  theme_gerstung(base_size = 6) + 
  theme(legend.position = "bottom",legend.key.height = unit(0.1,"cm"),
        legend.key.width = unit(0.4,"cm"),
        legend.box.spacing = unit(0,"cm")) + 
  scale_y_discrete(expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  ylab("") + 
  xlab("") +
  scale_fill_material(palette = "deep-purple",name = "R2") + 
  ggsave(filename = sprintf("figures/simulations/eps_comparison_heatmap.pdf"),
         useDingbats = F,
         width = 2.2,height = 1)
```