---
title: "The Natural History of Clonal Haematopoiesis"
subtitle: "Investigating fit to BNPR"
author: 
    - José Guilherme de Almeida
    - Moritz Gerstung
output: 
  html_document:
    theme: lumen
    highlight: tango
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    number_sections: true
---

In this notebook we inspect how we can use the BNPR fits from `phylodyn` to model annual growth.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Setup

```{r}
source("Scripts/vaf_dynamics_functions.R")
source("Scripts/make_ultrametric.R")

library(parallel)
library(Matrix)
library(castor)
library(phangorn)
library(phylodyn)
library(minpack.lm)
INLA:::inla.dynload.workaround()

change_point <- function(x, b0, m1, m2, delta) { 
  b0 + (x*m1) + (sapply(x-delta, function (t) max(0, t)) * m2)
}

plot_tree <- function(obj) { 
  tree_ultra <- obj$tree_ultra
  tree_ultra$edge.length[is.infinite(tree_ultra$edge.length)] <- 0
  tree_ultra$tip.label <- obj$tree$S
  driver_list <- Filter(function(x) length(x) > 5,obj$driver_list)
  if (length(driver_list) > 1) {
    colours <- RColorBrewer::brewer.pal(length(driver_list),"Set3")
  } else {
    colours <- c("black")
  }
  colour_code <- rep("black",length(tree_ultra$tip.label))
  R <- 1:length(driver_list)
  if (length(driver_list) == 0) {
    R <- numeric(0)
  } 
  for (i in R) {
    D <- driver_list[[i]]
    colour_code[tree_ultra$tip.label %in% unique(D)] <- colours[i]
  }
  plot(tree_ultra,tip.color = colour_code)
}

clade_from_mrca <- function(tree,mrca) {
  edge_idx <- which(tree$edge[,1] == mrca)
  continue <- T
  output <- c(which(tree$edge[,2] == mrca))
  while (continue == T) {
    if (length(edge_idx) > 0) {
      output <- c(output,edge_idx)
      new_edge_idx <- c()
      for (idx in edge_idx) {
        nodes <- tree$edge[idx,2]
        new_idxs <- which(tree$edge[,1] == nodes)
        new_edge_idx <- c(new_edge_idx,new_idxs)
      }
      edge_idx <- new_edge_idx
    } else {
      continue <- F
    }
  }
  return(output)
}

build_tree <- function(sub_data,subsample_size=100,
                       detection_threshold=0) {
  sub_af <- sub_data %>% 
    group_by(CloneID) %>% 
    summarise(N = max(NClones),.groups = "drop") %>%
    arrange(CloneID)
  MaxClone <- max(sub_data$CloneID)
  MaxMut <- max(sub_data$MutID)
  sub_af_sparse <- rep(0,MaxClone)
  sub_af_sparse[sub_af$CloneID] <- sub_af$N 
  
  Presence <- sparseMatrix(i = sub_data$CloneID,
                           j = sub_data$MutID,
                           dims = c(MaxClone,MaxMut))
  Presence[cbind(sub_data$CloneID,sub_data$MutID)] <- 1

  if (sum(as.logical(sub_af_sparse)) < subsample_size) {
    sub_af_sparse <- rep(1,length(sub_af_sparse))
  }
  
  S <- sample(MaxClone,subsample_size,replace = F,prob = sub_af_sparse)
  af <- sub_af_sparse[S]
  Presence <- Presence[S,]
  Presence <- as.matrix(rbind(Presence,wt=0))
  Presence <- Presence[,colSums(Presence) > 0]
  
  dst <- as.matrix(dist(Presence > 0,method = "manhattan"))
  tree <- root(njs(dst),
               outgroup = "wt",
               resolve.root = T,
               edgelabel = T) 
  tree <- drop.tip(tree,"wt")
  return(list(tree = tree,S = S,af = af))
}

read_clonex <- function(path,d = 1000) {
  x <- read_tsv(path,
                col_names = c("Gen","NClones","CloneID","MutID"),
                col_types = c(col_integer(),col_integer(),col_integer(),col_integer()),
                progress = F) %>%
    mutate(Driver = MutID <= d)
  return(x)
}

trajectory_from_td <- function(x=0:1000, td=0, s=0.01, N=2e5, g=1){
  s <- s/g
  t0 <- td - 1/(s)
  t <- x-t0
  t <- t*g
  y <- pmax(0,t)
  if (s==0)
    return(y)
  td <- 1/(s)
  te <- log(N*s -1)/s + td
  y[t>td] <- N/(1+exp(-(s*(t[t>td]-te))))
  return(y)
}

trajectory_from_t0 <- function(x=0:1000, t0=0, s=0.01, N=2e5, g=1){
  s <- s/g
  t <- x-t0
  t <- t*g
  y <- pmax(0,t)
  if (s==0)
    return(y)
  td <- 1/(s)
  te <- log(N*s -1)/s + td
  y[t>td] <- N/(1+exp(-(s*(t[t>td]-te))))
  return(y)
}
```

# BNPR

In this section we calculate/load all BNPR trajectories for the trees inferred from Wright-Fisher (WF) simulations. These simulations are done for 6 different fitness levels - 0.005, 0.010, 0.015, 0.020, 0.025 and 0.030 - over 800 generations and with a fixed population size (200,000).

## Loading data and inferring trees

The first step is to build the trees, which we then display. One can immediately see that expansions become increasingly prevalent for higher fitness effects, with quite a few cases of a single clone sweeping all of the population.

```{r}
N <- 2e5
N_DRIVERS <- 20
all_file_paths <- list.files(
  "hsc_output_bnpr_complete",pattern = "last_generation",
  full.names = T,recursive = T)
all_driver_file_paths <- list.files(
  "hsc_output_bnpr_complete",pattern = "driver",
  full.names = T,recursive = T)

file_name <- "data_output/simulated_trees_trajectories.RDS"
if (file.exists(file_name) == T) {
  tree_traj <- readRDS(file_name)
  trees <- tree_traj[[1]]
  all_driver_trajectories <- tree_traj[[2]]
} else {
  all_driver_trajectories <- all_driver_file_paths %>%
    lapply(function(x) {
      read.csv(x,header = F) %>%
        arrange(V1,V2) %>%
        select(Gen = V1,MutID = V2,Count = V3)})
  
  trees <- mclapply(
    all_file_paths,
    mc.cores = 6,
    mc.cleanup = T,
    function(path) {
      s <- str_match(path,"hsc_[0-9.]+") %>%
        gsub(pattern = "hsc_",replacement = "") %>%
        as.numeric()
      system(sprintf('echo "%s"', path)) # prints during mclapply by using bash
      x <- read_clonex(path,d = N_DRIVERS) %>%
        subset(Gen == 800) %>%
        subset(MutID != 0)
      x$R <- as.numeric(str_match(path,"[0-9.]+$"))
      x$s <- s
      x$drift_threshold <- 1 / (N * (s))
      
      driver_list <- x %>% 
        subset(Driver == T) %>% 
        select(CloneID,MutID) 
      driver_list_out <- list()
      for (driver_id in unique(driver_list$MutID)) {
        tmp <- driver_list %>%
          subset(MutID == driver_id) %>%
          select(CloneID) %>%
          unlist
        driver_list_out[[as.character(driver_id)]] <- tmp
      }
      tree <- x %>%
        build_tree()
      ultrametric_tree <- make.ultrametric.tree(tree$tree)
      gc()
      list(
        tree = tree,
        tree_ultra = ultrametric_tree,
        driver_list = driver_list_out,
        path = path
      ) %>%
        return
    }
  )
  
  names(trees) <- gsub('/last_generation','',all_file_paths)
  names(all_driver_trajectories) <- gsub('/driver_trajectory','',all_driver_file_paths)
  saveRDS(object = list(trees,all_driver_trajectories),file = file_name)
}
```

```{r}
par(mfrow = c(4,5),mar = c(2,0.5,1,0.5))
for (tree_name in names(trees)) {
  tree <- trees[[tree_name]]
  fitness <- str_match(tree_name,"[0-9.]+")
  plot(tree$tree_ultra,main = fitness)
}
```

## Calculating BNPR trajectories for the whole trees

Here we calculate the actual EPS trajectories using BNPR and display them.

```{r}
file_name <- "data_output/simulated_bnpr_trees.RDS"
if (file.exists(file_name) == T) {
  all_estimates_trees <- readRDS(file_name)
} else {
  all_estimates_trees <- mclapply(
    trees,
    mc.cores = 2, # do not go over 2 cores here, may lead to OOM errors
    mc.cleanup = T,
    function(x) {
      system(sprintf('echo "%s"', x$path)) # prints during mclapply by using bash
      BNPR(x$tree_ultra) %>%
        return
      }
    )
  saveRDS(all_estimates_trees,file_name)
}
```

## Calculating `mcmc.popsize` and skyline trajectories for the whole trees

We accompany our BNPR estimates with those obtained through `mcmc.popsize` and skyline estimations to prove that what we are observing are not artefacts specific to BNPR. Interestingly, `mcmc.popsize` appears to suffer from its own bias - in some trajectories a very large initial EPS followed by a dip is observable. This holds for both possible priors - constant population size and a skyline-process population size, with the latter ameliorating this effect slightly.

```{r}
all_estimates_trees_skyline <- mclapply(
    trees,
    mc.cores = 2,
    mc.cleanup = T,
    function(x) skyline(x$tree_ultra,epsilon = 0.01)
    )

all_estimates_trees_skyline_wide <- mclapply(
    trees,
    mc.cores = 2,
    mc.cleanup = T,
    function(x) skyline(x$tree_ultra,epsilon = 0.05)
    )

all_estimates_trees_mcmc <- mclapply(
    trees,
    mc.cores = 2, 
    mc.cleanup = T,
    function(x) mcmc.popsize(x$tree_ultra,nstep=10000,thinning = 100,
                             progress.bar = F,
                             burn.in = 1000,lambda = 0.1,
                             method.prior.heights = "constant")
    )

all_estimates_trees_mcmc_skyline <- mclapply(
    trees,
    mc.cores = 2, 
    mc.cleanup = T,
    function(x) mcmc.popsize(x$tree_ultra,nstep=10000,thinning = 100,
                             progress.bar = F,
                             burn.in = 1000,lambda = 0.1,
                             method.prior.heights = "skyline")
    )
```

```{r}
par(mfrow = c(4,5),mar = c(1.5,2,2,1))
for (name in names(all_estimates_trees)) {
  plot_BNPR(all_estimates_trees[[name]])
  lines(all_estimates_trees_skyline[[name]],col = "green")
  lines(all_estimates_trees_skyline_wide[[name]],col = "#026440")
  lines(extract.popsize(all_estimates_trees_mcmc[[name]]),col = "red")
  lines(extract.popsize(all_estimates_trees_mcmc_skyline[[name]]),col = "purple")
}
```

## Calculating BNPR trajectories for each clade

```{r}
file_name <- "data_output/simulated_bnpr_clades.RDS"
if (file.exists(file_name) == T) {
  all_estimates <- readRDS(file_name)
} else {
  all_estimates <- list()
  i <- 1
  for (obj_name in names(trees)) {
    obj <- trees[[obj_name]]
    print(obj_name)
    s <- str_match(obj_name,"[0-9.]+") %>%
      as.numeric
    Tree <- list(tree = obj$tree_ultra)
    Tree$tree$tip.label <- obj$tree$S
    Tree$tree$edge.length[is.infinite(Tree$tree$edge.length)] <- 0
    tree <- Tree$tree
    driver_id_list <- Filter(function(x) length(x) > 5,obj$driver_list)
    all_drivers <- do.call(c,driver_id_list)
    clades <- cut_tree(tree = tree,depth = 0.1) %>%
      Filter(f = function(x) length(x) >= 5)
    clades_ <- list()
    for (clade in clades) {
      clade_tips <- tree$tip.label[clade]
      for (driver in names(driver_id_list)) {
        if (any(clade_tips %in% driver_id_list[[driver]])) {
          clades_[[length(clades_)+1]] <- list(
            clade = clade,
            driver = driver
          )
        }
      }
    }
    clades <- clades_
    
    if (!(obj_name %in% names(all_estimates))) {
      bnpr_estimates <- mclapply(
        clades,
        mc.cores = 2,
        mc.cleanup = T,
        function(clad) {
          #tip_in_clade <- which(Tree$tree$tip.label %in% clad)
          tip_in_clade <- clad$clade
          if (length(tip_in_clade) > 4) {
            sub_tree <- keep.tip(Tree$tree,tip_in_clade) %>%
              multi2di()
            if (!is.null(sub_tree)) {
              bnpr_estimates <- NULL
              bnpr_estimate <- BNPR(sub_tree)
              if (!is.null(bnpr_estimate)) {
                list(
                  tree = tree,
                  bnpr = bnpr_estimate,
                  clade = clad$clade,
                  tree = sub_tree,
                  driver = clad$driver) %>%
                  return
              }
            }
          }
        }
      ) %>%
        Filter(f = function(x) !is.null(x))
      all_estimates[[obj_name]] <- bnpr_estimates
    }
    gc()
  }
  saveRDS(all_estimates,file_name)
}
```

## Calculating `mcmc.popsize` trajectories for each clade

```{r}
all_estimates_mcmc <- list()
i <- 1
par(mfrow = c(4,5),mar=c(1,2,2,1))
for (obj_name in names(all_estimates)) {
  print(obj_name)
  if (!(obj_name %in% names(all_estimates_mcmc))) {
    obj <- all_estimates[[obj_name]]
    estimates <- list()
    for (child_obj in obj) {
      Tr <- child_obj$tree %>%
        keep.tip(child_obj$clade) %>%
        multi2di()
      pr <- ifelse(length(child_obj$tree$tip.labels) > 10,"skyline","constant")
      pop_size_eps <- mcmc.popsize(Tr,10000,100,1000,
                                   lambda = 0.1,
                                   progress.bar = F,
                                   method.prior.heights = pr)
      estimates[[length(estimates)+1]] <- list(pop_size_eps = pop_size_eps)
    }
    estimates <- estimates %>%
     Filter(f = function(x) !is.null(x))
    all_estimates_mcmc[[obj_name]] <- estimates
  }
  gc()
}
```

## Calculating skyline trajectories for each clade

```{r}
all_estimates_skyline <- list()
names_done <- names(all_estimates_skyline)
i <- 1
par(mfrow = c(4,5),mar=c(1,2,2,1))
for (obj_name in names(all_estimates)) {
  print(obj_name)
  if (!(obj_name %in% names_done)) {
    obj <- all_estimates[[obj_name]]
    estimates <- list()
    for (child_obj in obj) {
      Tr <- child_obj$tree %>%
        keep.tip(child_obj$clade) %>%
        multi2di()
      sk <- skyline(Tr,epsilon = 0.02)
      estimates[[length(estimates)+1]] <- list(skyline = sk)
    }
    estimates <- estimates %>% 
      Filter(f = function(x) !is.null(x))
    all_estimates_skyline[[obj_name]] <- estimates
  }
  gc()
}
```

## Comparing BNPR, `mcmc.popsize` and skyline trajectories

### For clades with 10 or fewer tips

```{r}
# fewer than 10 tips per clade
par(mfrow = c(4,5),mar = c(1,1,1,1))
for (name in names(all_estimates)) {
  obj <- all_estimates[[name]]
  obj_mcmc <- all_estimates_mcmc[[name]]
  obj_skyl <- all_estimates_skyline[[name]]
  
  if (length(obj) > 0) {
    for (i in 1:length(obj)) {
      if (length(obj[[i]]$clade) <= 10) {
        bnpr_traj <- obj[[i]]$bnpr
        mcmc_traj <- obj_mcmc[[i]]$pop_size_eps
        skyl_traj <- obj_skyl[[i]]$skyline
        
        plot_BNPR(bnpr_traj)
        try(lines(extract.popsize(mcmc_traj),col = "red"))
        lines(skyl_traj,col = "green")
      }
    }
  }
}
```

### For clades with 10-50 tips

```{r}
# 10-50 tips
par(mfrow = c(4,5),mar = c(1,1,1,1))
for (name in names(all_estimates)) {
  obj <- all_estimates[[name]]
  obj_mcmc <- all_estimates_mcmc[[name]]
  obj_skyl <- all_estimates_skyline[[name]]
  
  if (length(obj) > 0) {
    for (i in 1:length(obj)) {
      if (length(obj[[i]]$clade) < 50 & length(obj[[i]]$clade) > 10) {
        bnpr_traj <- obj[[i]]$bnpr
        mcmc_traj <- obj_mcmc[[i]]$pop_size_eps
        skyl_traj <- obj_skyl[[i]]$skyline
        
        plot_BNPR(bnpr_traj)
        try(lines(extract.popsize(mcmc_traj),col = "red"))
        lines(skyl_traj,col = "green")
      }
    }
  }
}
```

### For clades with 50+ tips

```{r}
# 50-100 clades
par(mfrow = c(4,5),mar = c(1,1,1,1))
for (name in names(all_estimates)) {
  obj <- all_estimates[[name]]
  obj_mcmc <- all_estimates_mcmc[[name]]
  obj_skyl <- all_estimates_skyline[[name]]
  
  if (length(obj) > 0) {
    for (i in 1:length(obj)) {
      if (length(obj[[i]]$clade) >= 50) {
        bnpr_traj <- obj[[i]]$bnpr
        mcmc_traj <- obj_mcmc[[i]]$pop_size_eps
        skyl_traj <- obj_skyl[[i]]$skyline
        
        plot_BNPR(bnpr_traj)
        try(lines(extract.popsize(mcmc_traj),col = "red"))
        # extracting population size fails for a few cases
        lines(skyl_traj,col = "green")
      }
    }
  }
}
```

### Comparing fits derived from BNPR, `mcmc.popsize` and skyline

There is quite good agreement between log-linear trajectories fitted to skyline estimation and BNPR. This does not hold as well when comparing `mcmc.popsize` and BNPR - probably due to the initial dip in population size, the EPS estimates derived from `mcmc.popsize` lead to a slight underestimation of log-linear growth trajectories. As for early/late growth rate comparisons we observe little agreement between BNPR and `mcmc.popsize` or skyline estimation.

```{r,fig.height=4,fig.width=6}
linear_estimates <- list()
early_late_estimates <- list()
idx <- 1
for (x in names(all_estimates)) {
  obj <- all_estimates[[x]]
  obj_mcmc <- all_estimates_mcmc[[x]]
  obj_skyl <- all_estimates_skyline[[x]]
  linear_estimates[[x]] <- list()
  early_late_estimates[[x]] <- list()
  if (length(obj) > 0){
    for (i in 1:length(obj)) {
      try(
        {
          bnpr_data <- data.frame(
            X = (1 - obj[[i]]$bnpr$summary$time)*800,
            Y = log(obj[[i]]$bnpr$summary$quant0.5))
          ps <- extract.popsize(obj_mcmc[[i]]$pop_size_eps)
          X_ <- seq(min(obj_skyl[[i]]$skyline$time),max(obj_skyl[[i]]$skyline$time),length.out = 100)
          sk_Y <- obj_skyl[[i]]$skyline$population.size[as.numeric(cut(X_,c(0,obj_skyl[[i]]$skyline$time)))]
          skyl_data <- data.frame(
            X = (1 - X_)*800,
            Y = log(sk_Y)
            ) %>% 
            na.omit
          mcmc_data <- data.frame(
            X = (1 - ps[,1])*800,
            Y = log(ps[,3])) %>%
            na.omit
          linear_estimates[[x]][[i]] <- list(
            id = idx,
            bnpr = lm(Y ~ X,data = bnpr_data),
            skyl = lm(Y ~ X,data = skyl_data),
            mcmc = lm(Y ~ X,data = mcmc_data)
          )
          early_late_estimates[[x]][[i]] <- list(
            id = idx,
            bnpr = list(
              late = lm(Y ~ X,data = bnpr_data[1:20,]),
              early = lm(Y ~ X,data = bnpr_data[nrow(bnpr_data) - c(19:0),])),
            skyl = list(
              late = lm(Y ~ X,data = skyl_data[1:20,]),
              early = lm(Y ~ X,data = skyl_data[nrow(skyl_data) - c(19:0),])),
            mcmc = list(
              late = lm(Y ~ X,data = mcmc_data[1:40,]),
              early = lm(Y ~ X,data = mcmc_data[nrow(mcmc_data) - c(39:0),]))
            )
          }
      )
    }
  }
  idx <- idx + 1
}

linear_estimates_bnpr_mcmc_skyl <- linear_estimates %>% 
    lapply(function(x) do.call(rbind,
                               lapply(x, function(y) c(y$id,
                                                       y$bnpr$coefficient[2],
                                                       y$mcmc$coefficient[2],
                                                       y$skyl$coefficient[2])))) %>%
  do.call(what = rbind) %>%
  as.data.frame()
colnames(linear_estimates_bnpr_mcmc_skyl) <- c("id","BNPR","mcmc.popsize","Skyline")
linear_estimates_bnpr_mcmc_skyl$id <- names(all_estimates)[linear_estimates_bnpr_mcmc_skyl$id]
linear_estimates_bnpr_mcmc_skyl$fitness <- linear_estimates_bnpr_mcmc_skyl$id %>%
  str_match("[0-9.]+") %>% 
  as.numeric

early_estimates_bnpr_mcmc_skyl <- early_late_estimates %>% 
    lapply(function(x) do.call(rbind,
                               lapply(x, function(y) c(y$id,
                                                       y$bnpr$early$coefficient[2],
                                                       y$mcmc$early$coefficient[2],
                                                       y$skyl$early$coefficient[2])))) %>%
  do.call(what = rbind) %>%
  as.data.frame()
colnames(early_estimates_bnpr_mcmc_skyl) <- c("id","BNPR","mcmc.popsize","Skyline")
early_estimates_bnpr_mcmc_skyl$id <- names(all_estimates)[early_estimates_bnpr_mcmc_skyl$id]
early_estimates_bnpr_mcmc_skyl$fitness <- early_estimates_bnpr_mcmc_skyl$id %>%
  str_match("[0-9.]+") %>% 
  as.numeric

late_estimates_bnpr_mcmc_skyl <- early_late_estimates %>% 
  Filter(f = function(f) !is.null(x)) %>% 
  lapply(function(x) do.call(rbind,
                             lapply(x, function(y) c(y$id,
                                                     y$bnpr$late$coefficient[2],
                                                     ifelse(!is.null(y$bnpr$late),summary(y$bnpr$late)$coefficients[2,2],NA),
                                                     y$mcmc$late$coefficient[2],
                                                     y$skyl$late$coefficient[2])))) %>%
  do.call(what = rbind) %>%
  as.data.frame()
colnames(late_estimates_bnpr_mcmc_skyl) <- c("id","BNPR","BNPR_se","mcmc.popsize","Skyline")
late_estimates_bnpr_mcmc_skyl$id <- names(all_estimates)[late_estimates_bnpr_mcmc_skyl$id]
late_estimates_bnpr_mcmc_skyl$fitness <- late_estimates_bnpr_mcmc_skyl$id %>%
  str_match("[0-9.]+") %>% 
  as.numeric

A <- linear_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = Skyline,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.1,0.1),
                  ylim = c(-0.1,0.1))

B <- early_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = Skyline,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.1,0.1),
                  ylim = c(-0.1,0.1))

C <- late_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = Skyline,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.1,0.1),
                  ylim = c(-0.1,0.1))

D <- linear_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = mcmc.popsize,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.1,0.1),
                  ylim = c(-0.1,0.1))

E <- early_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = mcmc.popsize,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.1,0.1),
                  ylim = c(-0.1,0.1))

FF <- late_estimates_bnpr_mcmc_skyl %>%
  ggplot() +
  geom_point(aes(x = mcmc.popsize,y = BNPR),size = 0.5) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(-0.1,0.1),
                  ylim = c(-0.1,0.1))

plot_grid(A,B,C,D,E,FF,nrow = 2)

data.frame(
  BNPR = c(cor(linear_estimates_bnpr_mcmc_skyl$fitness,linear_estimates_bnpr_mcmc_skyl$BNPR),
           cor(early_estimates_bnpr_mcmc_skyl$fitness,early_estimates_bnpr_mcmc_skyl$BNPR))^2,
  Skyline = c(cor(linear_estimates_bnpr_mcmc_skyl$fitness,linear_estimates_bnpr_mcmc_skyl$Skyline),
              cor(early_estimates_bnpr_mcmc_skyl$fitness,early_estimates_bnpr_mcmc_skyl$Skyline))^2,
  mcmc.popsize = c(cor(linear_estimates_bnpr_mcmc_skyl$fitness,linear_estimates_bnpr_mcmc_skyl$mcmc.popsize),
                   cor(early_estimates_bnpr_mcmc_skyl$fitness,early_estimates_bnpr_mcmc_skyl$mcmc.popsize))^2,
  fit = c("Log-linear","Early log-linear"))
```

## Timing clones from trees

```{r}
for (x in names(all_estimates)) {
  estimate <- all_estimates[[x]]
  if (length(estimate) > 0) {
    for (y in 1:length(estimate)) {
      m <- MRCA(estimate[[y]]$tree,estimate[[y]]$clade)
      estimate[[y]]$mutation_timing <- time_mutation(
        estimate[[y]]$tree,m)
      all_estimates[[x]][[y]]$mutation_timing <- estimate[[y]]$mutation_timing
    }
  }
}
```

# Modelling BNPR trajectories

Here we model the EPS trajectories using different types of models, namely:

* A log-linear model
* A scaled and shifted sigmoidal curve
* A log-linear model with one changepoint (captures the earlier and later phases of growth)

We also fit the same models to the original WF trajectories one gets for each individual driver. By doing so, we can more accurately verify whether we are able to recapitulate the clonal growth as it is happening with the BNPR trajectories.

## Fitting different models to BNPR trajectory

```{r}
all_fits <- list()
par(mfrow = c(4,5),mar = c(3,1,1,1))
for (obj_name in names(all_estimates)) {
  bnpr_estimates <- all_estimates[[obj_name]] %>%
    Filter(f = function(x) class(x$bnpr) != "try-error")
  new_bnpr_estimates <- list()
  for (x in bnpr_estimates) {
    bnpr_estimate <- x$bnpr
    Y <- log(bnpr_estimate$summary$mean)
    Y_ <- Y - min(Y,na.rm = T) + 1e-8
    Y025 <- log(bnpr_estimate$summary$quant0.025)
    Y975 <- log(bnpr_estimate$summary$quant0.975)
    X <- (1-bnpr_estimate$summary$time) * 800
    tmp_df <- data.frame(
      X = X,Y = Y,Y_ = Y_,
      Y025 = Y025,Y975 = Y975,
      W = (Y975 - Y025)^2/16
    ) %>% 
      na.omit()
    tmp_df <- tmp_df[!is.infinite(tmp_df$Y),]
    tmp_df$Y_exp <- exp(tmp_df$Y)
    
    tmp_df$W[is.infinite(tmp_df$W)] <- max(tmp_df$W[!is.infinite(tmp_df$W)])
    linear_estimate <- lm(Y ~ X,w = 1 / tmp_df$W,data = tmp_df)
    m <- linear_estimate$coefficients[2]
    linear_estimate_no_weights <- lm(Y ~ X,data = tmp_df)
    
    if (nrow(tmp_df) >= 10 & max(tmp_df$Y_exp) < 1e100) {
      non_linear_estimate <- nlsLM(
        Y_exp ~ SSlogis(X, Asym, b2, b3),
        start = list(Asym = max(tmp_df$Y_exp),b2 = mean(tmp_df$X),b3 = 10),
        control = nls.control(maxiter = 1000,warnOnly = T),
        data = tmp_df,
        algorithm = "port",
        weights = 1/tmp_df$W)
      x_earliest <- seq(min(tmp_df$X),min(tmp_df$X) + 50,length.out = 20)
      y_earliest <- approx(tmp_df$X,tmp_df$Y,xout = x_earliest)$y
      
      x_latest <- seq(max(tmp_df$X) - 50,max(tmp_df$X),length.out = 20)
      y_latest <- approx(tmp_df$X,tmp_df$Y,xout = x_latest)$y
      
      latest <- lm(y_latest ~ x_latest)
      earliest <- lm(y_earliest ~ x_earliest)

      if (class(non_linear_estimate) == "try-error") {
        non_linear_estimate <- try(nlsLM(
          Y_exp ~ SSlogis(X, Asym, b2, b3),
          start = list(Asym = max(tmp_df$Y_exp),b2 = max(tmp_df$X),b3 = 10),
          control = nls.control(maxiter = 1000,warnOnly = T),
          data = tmp_df,
          algorithm = "port",
          weights = 1/tmp_df$W))
      }
    } else {
      non_linear_estimate <- NULL
      earliest <- NULL
      latest <- NULL
    }
    
    opt_fn <- function(par) {
      se <- (tmp_df$Y - change_point(tmp_df$X,par[1],par[2],par[3],par[4]))^2
      return(mean(se / tmp_df$W))
    }
    change_point_regression <- optim(
      par = c(linear_estimate$coefficients[1],
              linear_estimate$coefficients[2],
              0,
              mean(X)),
      method = "L-BFGS-B",
      lower = c(NA,0,NA,0),
      upper = c(NA,NA,NA,800),
      fn = opt_fn)
    x$linear_estimate <- linear_estimate
    x$linear_estimate_no_weights <- linear_estimate_no_weights
    x$change_point_regression <- change_point_regression
    x$non_linear_estimate <- non_linear_estimate
    x$data <- tmp_df
    x$earliest_growth <- earliest
    x$latest_growth <- latest
    new_bnpr_estimates[[length(new_bnpr_estimates) + 1]] <- x
  }
  
  bnpr_estimates <- new_bnpr_estimates %>%
    Filter(f = function(x) !is.null(x))
  
  all_estimates[[obj_name]] <- bnpr_estimates
  for (bnpr in bnpr_estimates) {
    cpr <- bnpr$change_point_regression
    all_fits[[length(all_fits)+1]] <- data.frame(
      x = bnpr$data$X,y = exp(bnpr$data$Y),
      obj_name,n = length(all_fits)+1,
      clade_size = length(bnpr$clade))

    if ((cpr$par[2] + cpr$par[3]) < 0) {
        # plot_BNPR(bnpr$bnpr,main = sprintf("%s",length(bnpr$clade)))
        # plot(bnpr$data$X,exp(bnpr$data$Y),col = "black")
        # if (!is.null(bnpr$non_linear_estimate)) {
        #   lines(bnpr$data$X,predict(
        #     bnpr$non_linear_estimate,newdata = data.frame(X = bnpr$data$X)),
        #     col = "red")
        # }
        # lines(1-bnpr$data$X/800,
        #       exp(change_point(bnpr$data$X,cpr$par[1],cpr$par[2],cpr$par[3],cpr$par[4])),
        #       col="red")
      }
    }
}
```

## Fitting different models to driver trajectories from the original WF simulations

```{r}
all_driver_fits <- list()
par(mfrow = c(4,5),mar = c(1.5,2.5,2.5,1))
for (x in names(all_driver_trajectories)) {
  driver_trajectories <- all_driver_trajectories[[x]]
  driver_trajectories <- driver_trajectories %>%
    arrange(MutID,Gen) %>%
    group_by(MutID) %>%
    mutate(Break = c(T,diff(Gen) != 20)) %>%
    mutate(Component = cumsum(Break)) %>%
    mutate(MutID_C = sprintf("%s_%s",MutID,Component))
  drivers <- unique(driver_trajectories$MutID_C)
  all_driver_fits[[x]] <- list()
  for (driver in drivers) {
    tmp_df <- driver_trajectories %>%
      subset(MutID_C == driver)
    driver <- tmp_df$MutID[1]
    has_last_gen <- any(tmp_df$Gen == 800)
    large_at_last_gen <- ifelse(
      has_last_gen,
      tmp_df$Count[tmp_df$Gen == 800] > 200,
      F)
    
    if (nrow(tmp_df) >= 5 & has_last_gen & large_at_last_gen) {
      opt_fn <- function(par) {
        se <- (log(tmp_df$Count) - change_point(tmp_df$Gen,par[1],par[2],par[3],par[4]))^2
        return(mean(se))
      }
      
      x_earliest <- seq(min(tmp_df$Gen),min(tmp_df$Gen) + 50,length.out = 20)
      y_earliest <- approx(tmp_df$Gen,tmp_df$Count,xout = x_earliest)$y
      
      x_latest <- seq(max(tmp_df$Gen) - 50,max(tmp_df$Gen),length.out = 20)
      y_latest <- approx(tmp_df$Gen,tmp_df$Count,xout = x_latest)$y
      
      latest <- lm(y_latest ~ x_latest)
      earliest <- lm(y_earliest ~ x_earliest)

      linear_sol <- lm(log(Count) ~ Gen,data = tmp_df)
      change_point_regression <- optim(
        par = c(linear_sol$coefficients[1],linear_sol$coefficients[2],0,mean(tmp_df$Gen)),
        method = "L-BFGS-B",
        lower = c(NA,0,NA,0),
        upper = c(NA,NA,NA,800),
        fn = opt_fn)
      non_linear_estimate <- nlsLM(
        Count ~ SSlogis(Gen, Asym, b2, b3),
        start = list(Asym = max(tmp_df$Count),b2 = mean(tmp_df$Gen),b3 = 10),
        control = nls.control(maxiter = 1000,warnOnly = T),
        data = tmp_df,
        upper = c(Asym = 2e5,b2 = 3000,b3 = 1e5),
        algorithm = "port")
      last_vaf <- tmp_df$Count[tmp_df$Gen == 800]/2e5
      all_driver_fits[[x]][[as.character(driver)]] <- list(
        data = tmp_df,
        non_linear_estimate = non_linear_estimate,
        change_point_regression = change_point_regression,
        earliest_driver = earliest,
        latest_driver = latest,
        last_vaf = last_vaf,
        gen_at_onset = min(tmp_df$Gen))
    }
  }
}
```

## Extracting coefficients from BNPR fits

```{r}
all_bnpr_fits <- names(all_estimates) %>%
  lapply(
    function(x) {
      if (length(all_estimates[[x]]) > 0) {
        tree_eps <- unlist(all_estimates_trees[[x]]$summary[1,c(4,2,6)])
        lapply(1:length(all_estimates[[x]]),function(y) {
          y <- as.numeric(y)
          if (!is.null(all_estimates[[x]][[y]]$non_linear_estimate)) {
            pars <- all_estimates[[x]][[y]]$non_linear_estimate$m$getAllPars()
            tao <- all_estimates[[x]][[y]]$mutation_timing * 800
            cp <- all_estimates[[x]][[y]]$change_point_regression$par
            pop_size_at_tao_cp <- exp((800 - tao)*cp[2])
            earliest <- all_estimates[[x]][[y]]$earliest_growth$coefficients[2]
            latest <- all_estimates[[x]][[y]]$latest_growth$coefficients[2]
            O <- data.frame(
              l = all_estimates[[x]][[y]]$linear_estimate$coefficients[2],
              nl = 1 / pars[3],
              nl_asym = pars[1],
              nl_xmid = pars[2],
              l_nw = all_estimates[[x]][[y]]$linear_estimate_no_weights$coefficients[2],
              cp_1 = cp[2],
              cp_2 = cp[3],
              earliest = earliest,
              latest = latest,
              changepoint_bnpr = cp[4],
              time_at_onset_low = tao[1],
              time_at_onset_high = tao[2],
              w_sum = mean(all_estimates[[x]][[y]]$data$W,na.rm=T),
              pop_at_detection_low = pop_size_at_tao_cp[2],
              pop_at_detection_high = pop_size_at_tao_cp[1],
              tree_eps_low = tree_eps[1],
              tree_eps_mean = tree_eps[2],
              tree_eps_high = tree_eps[3],
              point_of_saturation = which(
                c(
                  predict(
                    all_estimates[[x]][[y]]$non_linear_estimate,
                    newdata = data.frame(X = seq(1,800)))) > (0.9 * pars[1])) %>% min,
              name = x,
              clade_size = length(all_estimates[[x]][[y]]$clade),
              driver = all_estimates[[x]][[y]]$driver,
              clade = y)
            return(O)
          }
        }) %>%
          do.call(what = rbind)
      } else {
        return(NULL)
      }
    }
    ) %>%
  do.call(what = rbind) %>%
  mutate(fitness = as.numeric(gsub('_',"",str_match(name,"_[0-9.]+_")))) %>%
  group_by(name) %>%
  mutate(N = length(unique(clade))) %>%
  mutate(cp_late = cp_1 + cp_2)
```

## Extracting coefficients from driver fits from the original WF simulations

```{r}
all_driver_fits_df <- names(all_driver_fits) %>% 
  lapply(function(x) {
    if (length(all_driver_fits[[x]]) > 0) {
      lapply(names(all_driver_fits[[x]]),function(y) {
        pars <- all_driver_fits[[x]][[y]]$non_linear_estimate$m$getAllPars()
        earliest <- all_driver_fits[[x]][[y]]$earliest_driver$coefficients[2]
        latest <- all_driver_fits[[x]][[y]]$latest_driver$coefficients[2]
        
        early_cp <- all_driver_fits[[x]][[y]]$change_point_regression$par[2]
        
        m <- all_driver_fits[[x]][[y]]$data %>%
          subset(Gen == min(Gen))
        v <- m$Count / 2e5
        time <- m$Gen
        m <- -log((1 - v)/v) - early_cp * time
        expected_last_vaf <- 1 / (1 + exp(-(m + early_cp * 800)))
        return(
          data.frame(
            Asym = pars[1],
            xmid = pars[2],
            growth_rate = 1/pars[3],
            early_growth = all_driver_fits[[x]][[y]]$change_point_regression$par[2],
            late_growth = all_driver_fits[[x]][[y]]$change_point_regression$par[3],
            earliest_driver = earliest,
            latest_driver = latest,
            changepoint_wf = all_driver_fits[[x]][[y]]$change_point_regression$par[4],
            point_of_saturation_wf = which(
              c(
                predict(
                  all_driver_fits[[x]][[y]]$non_linear_estimate,
                  newdata = data.frame(Gen = seq(1,800)))) > (0.9 * pars[1])) %>% min,
            name = x,
            driver = y,
            last_vaf = all_driver_fits[[x]][[y]]$last_vaf,
            expected_last_vaf,
            gen_at_onset = all_driver_fits[[x]][[y]]$gen_at_onset
          )
        )
      }) %>%
        do.call(what = rbind)
    }
  }) %>%
  do.call(what = rbind) %>%
  mutate(late_growth = early_growth + late_growth)
```

# Comparing BNPR trajectory models

## Plotting BNPR trajectories

Here, we plot all of the inferred BNPR trajectories for each clade. 

```{r,fig.height=2,fig.width=3}
all_fits %>%
  do.call(what = rbind) %>%
  mutate(fitness = str_match(obj_name,"[0-9.]+")) %>% 
  group_by(n) %>%
  mutate(x = x - min(x)) %>% 
  filter(min(y) > 1e-6 & max(y) < 1e8) %>%
  mutate(fitness = factor(
    fitness,
    levels = seq(0.005,0.03,by = 0.005),
    labels = sprintf("s=%s",seq(0.005,0.03,by = 0.005)))) %>% 
  ggplot(aes(x = x,y = y)) + 
  geom_line(aes(group = n),alpha = 0.25,size = 0.25) + 
  facet_wrap(~ fitness) + 
  scale_y_continuous(trans = 'log10',expand = c(0,0),breaks = c(1e-2,1,1e2,1e4,1e6)) +
  scale_x_continuous(expand = c(0,0)) +
  theme_gerstung(base_size = 6) + 
  theme(strip.text = element_text(margin = margin(b = 0.5))) +
  ylab("Effective population size (EPS)") + 
  xlab("Time since first coal. (generations)") + 
  coord_cartesian(ylim = c(1e-2,1e7)) +
  ggsave("figures/simulations/trajectories_BNPR.pdf",height = 1.5,width = 2.3)
```

## Comparing BNPR and WF trajectories

Here we scale the BNPR trajectories so that they match as closely as possible those obtained from the WF trajectories. A fairly good fit between BNPR and WF trajectories is observable for a large fraction of trajectories.

```{r,fig.height=6,fig.width=7}
simulated_trajectories <- list()
bnpr_estimates_ <- list()
for (x in names(all_driver_fits)) {
  if (x %in% names(all_estimates)) {
    for (y in names(all_driver_fits[[x]])) {
      clade_no <- 1
      bnpr_sums <- list()
      largest_clade <- which.max(unlist(lapply(all_estimates[[x]],function(x) length(x$clade))))[1]
      for (estimate in all_estimates[[x]]) {
        if (y == estimate$driver) {
          driver_fit <- all_driver_fits[[x]][[y]]$data
          bnpr_sum <- estimate$bnpr$summary
          W <- (log(bnpr_sum$quant0.975) - log(bnpr_sum$quant0.025))^2/16
          bnpr_sum$time <- (1 - bnpr_sum$time)*800
          if (clade_no == largest_clade) {
            opt_data <- data.frame(
              a = approx(bnpr_sum$time,y = bnpr_sum$mean,xout = driver_fit$Gen)$y,
              b = driver_fit$Count) %>%
              na.omit
            opt_fn <- function(pars) {
              l <- (opt_data$a * pars[1] - opt_data$b)^2
              l <- l
              return(sum(l))
            }
            opt_solution <- optim(c(m = 0.001,b = 1),
                                  opt_fn,
                                  control = list(maxit = 10000),
                                  method = "Nelder-Mead")
            P <- opt_solution$par
          }
          driver_fit$id <- x
          driver_fit$driver <- y
          bnpr_sum$id <- x
          bnpr_sum$driver <- y
          bnpr_sum$clade <- clade_no
          simulated_trajectories[[length(simulated_trajectories)+1]] <- driver_fit
          bnpr_sums[[length(bnpr_sums)+1]] <- bnpr_sum
          clade_no <- clade_no + 1
        }
      }
      for (bnpr_sum in bnpr_sums) {
        bnpr_sum$transformed_y <- bnpr_sum$mean * P[1]
        bnpr_sum$transformed_y_low <- bnpr_sum$quant0.025 * P[1]
        bnpr_sum$transformed_y_high <- bnpr_sum$quant0.975 * P[1]
        bnpr_estimates_[[length(bnpr_estimates_)+1]] <- bnpr_sum
      }
    }
  }
}

simulated_trajectories_df <- do.call(rbind,simulated_trajectories)
bnpr_estimates_df <- do.call(rbind,bnpr_estimates_)
ggplot(data = NULL) +
  geom_hline(yintercept = 2e5,size = 0.25,colour = "goldenrod",alpha = 0.7) +
  geom_line(data = simulated_trajectories_df,
            aes(x = Gen/100,y = Count,colour = "Simulated",
                group = paste(MutID_C,driver)),
            size = 0.25,
            alpha = 0.7) + 
  geom_line(data = bnpr_estimates_df,aes(x = time/100,y = transformed_y,
                                         group = paste(clade,driver),colour = "BNPR estimate"),
            size = 0.25,
            alpha = 0.5) + 
  facet_wrap(~ id,scales = "free") + 
  theme_gerstung(base_size = 6) + 
  theme(strip.text = element_blank(),legend.position = "bottom") + 
  scale_y_continuous(trans = 'log10') + 
  coord_cartesian(ylim = c(NA,2e8),c(0,8)) +
  scale_colour_manual(values = c(Simulated = "grey50",`BNPR estimate` = "orchid"),name = NULL) + 
  xlab("Generation (x100)") + 
  ggsave("figures/simulations/trajectories_wf_bnpr.pdf",height = 6,width = 7)
```

## Comparing inference of early and late growth between BNPR fits and driver fits from WF simulations

We define here a measure of excessive observed variance in the BNPR trajectory - whenever the average log-EPS variance is larger than 10 we define these trajectories as "high variance" trajectories, noting that they are trajectories which were obtained from fairly small clades (fewer than 10 tips).

```{r,fig.height=1.5,fig.width=1.5}
bnpr_wf_merge_df <- merge(all_bnpr_fits,all_driver_fits_df,by = c("name","driver")) %>%
  mutate(vaf_at_detection_low = pop_at_detection_low/2e5,
         vaf_at_detection_high = pop_at_detection_high/2e5) %>%
  mutate(vaf_at_detection_low = ifelse(vaf_at_detection_low > 1,1,vaf_at_detection_low),
         vaf_at_detection_high = ifelse(vaf_at_detection_high > 1,1,vaf_at_detection_high)) %>%
  mutate(vaf_contained = last_vaf > vaf_at_detection_low & last_vaf < vaf_at_detection_high) %>%
  mutate(vaf_contained_higher_lower = ifelse(
    vaf_contained == F,
    ifelse(last_vaf > vaf_at_detection_high,"higher","lower"),
    "contained")) %>%
  mutate(col = ifelse(w_sum < 10,"Low variance","High variance")) %>%
  mutate(wf_ratio = late_growth/early_growth,bnpr_ratio = cp_late/cp_1) %>% 
  mutate(bnpr_ratio = ifelse(is.infinite(bnpr_ratio),1,bnpr_ratio),
         wf_ratio = ifelse(is.infinite(wf_ratio),1,wf_ratio))


bnpr_wf_merge_df %>% 
  ggplot(aes(x = w_sum)) +
  geom_density() +
  scale_x_continuous(trans = 'log10') + 
  theme_gerstung(base_size = 6) + 
  xlab("Average variance")

ggplot(bnpr_wf_merge_df) +
  geom_boxplot(aes(x = col,y = clade_size),outlier.size = 0.5) + 
  scale_y_continuous(breaks = seq(0,100,by = 10)) + 
  theme_gerstung(base_size = 6) + 
  xlab("Trajectory type") + 
  ylab("Clade size")
```

```{r,fig.height=1.4,fig.width=4.5}
early_v_early_plot <- bnpr_wf_merge_df %>% 
  ggplot(aes(x = early_growth,y = cp_1,colour = col)) + 
  geom_abline(slope = 1,linetype = 2, size = 0.25) +
  geom_point(size = 0.5) + 
  geom_smooth(method = "lm",alpha = 0.5,size = 0.25) +
  theme_gerstung(base_size = 6) + 
  coord_cartesian(xlim = c(0,0.07),ylim = c(0,0.07))  + 
  xlab("WF early growth") + 
  ylab("BNPR early growth") + 
  theme(legend.position = "bottom",
        legend.key.size = unit(0,"cm")) + 
  scale_colour_lancet() + 
  scale_colour_lancet(name = NULL)

late_v_late_plot <- bnpr_wf_merge_df %>% 
  ggplot(aes(x = late_growth,y = cp_late,colour = col)) + 
  geom_abline(slope = 1,linetype = 2, size = 0.25) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm",alpha = 0.5,size = 0.25) +
  theme_gerstung(base_size = 6) +
  coord_cartesian(xlim = c(-0.03,0.04),ylim = c(-0.03,0.04)) + 
  xlab("WF late growth") + 
  ylab("BNPR late growth") + 
  theme(legend.position = "bottom",
        legend.key.size = unit(0,"cm")) + 
  scale_colour_lancet() + 
  scale_colour_lancet(name = NULL)

cp_v_cp_plot <- bnpr_wf_merge_df %>% 
  ggplot(aes(x = changepoint_wf,y = changepoint_bnpr,colour = col)) + 
  geom_abline(slope = 1,linetype = 2, size = 0.25) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm",alpha = 0.5,size = 0.25) +
  theme_gerstung(base_size = 6) +
  coord_cartesian(xlim = c(200,800),ylim = c(200,800)) + 
  xlab("WF change point") + 
  ylab("BNPR change point") + 
  theme(legend.position = "bottom",
        legend.key.size = unit(0,"cm")) + 
  scale_colour_lancet(name = NULL)

plot_grid(early_v_early_plot + theme(legend.position = "none"),
          late_v_late_plot + theme(legend.position = "none"),
          cp_v_cp_plot + theme(legend.position = "none"),
          align = "hv",
          nrow = 1) %>%
  plot_grid(get_legend(early_v_early_plot),rel_heights = c(1,0.05),ncol = 1) + 
  ggsave("figures/simulations/compare_cp_wf_bnpr.pdf",height = 1.5,width = 4.5)
```

## Comparing the growth ratio between early and late phases of growth

We observe here that the ratio defined by $\frac{LateGrowth}{EarlyGrowth}$ for both WF and BNPR independently fits quite closely to the identity, thus proving that we are capable of detecting this reasonably well using the BNPR trajectories alone.

```{r,fig.height=2,fig.width=2}
bnpr_wf_merge_df %>% 
  subset(early_growth > 0 & cp_1 > 0) %>% 
  ggplot(aes(x = late_growth/early_growth,y = cp_late/cp_1)) + 
  geom_abline(slope = 1,linetype = 2,size = 0.25) +
  geom_point(size = 0.5,alpha = 0.4) + 
  geom_smooth(method = "lm",colour = "red4",alpha = 0.5,size = 0.25) +
  coord_cartesian(xlim = c(-1,4),ylim = c(-1,4)) + 
  theme_gerstung(base_size = 6) + 
  xlab("WF growth ratio") + 
  ylab("BNPR growth ratio")

bnpr_wf_merge_df %>%
  lm(formula = bnpr_ratio ~ wf_ratio) %>% 
  summary

table(all_driver_fits_df$last_vaf < all_driver_fits_df$expected_last_vaf) / nrow(all_driver_fits_df)
```

## Comparing the expected and observed late growth 

We also want to have a measure of growth deceleration that is more comparable with our inferences from the time series data - based on a sigmoidal fit with a fixed carrying capacity ($VAF = 0.5$) that is then extrapolated to its onset of origin. As such, we fit a sigmoidal to the BNPR that assumes a total carrying capacity of $1$ and that the final EPS corresponds to a proxy for the VAF that is parametrised as $\hat{VAF} = \frac{Number\ Of\ Tips\ In\ The\ Clade}{Total\ Tips\ In\ The\ Tree}$. 

```{r,fig.height=1.5,fig.width=1.5}
expected_late_growth <- function(vaf,id,clade) {
  clade <- as.numeric(clade)
  if (clade <= length(all_estimates[[id]])) {
    
    tmp_df <- all_estimates[[id]][[clade]]$data %>%
      arrange(X)
    last_eps <- tmp_df$Y_exp[nrow(tmp_df)]
    last_eps <- max(tmp_df$Y_exp)
    tmp_df$normalized_Y <- tmp_df$Y_exp / last_eps * vaf

    nl_fit <- nls(formula = normalized_Y ~ SSlogis(X,1,Xmid,b),
                  start = list(Xmid = mean(tmp_df$X),b = 10),
                  control = nls.control(maxiter = 1000,minFactor = 1/(1024^32),
                                        warnOnly = T),
                  data = tmp_df)
    return(nl_fit$m$getAllPars()[2])
  } else {
    return(NA)
  }
}

bnpr_wf_merge_df$expected_nl <- rep(NA,nrow(bnpr_wf_merge_df))

for (i in 1:nrow(bnpr_wf_merge_df)) {
  sub_bnpr_wf_merge_df <- bnpr_wf_merge_df[i,]
  last_vaf <- sub_bnpr_wf_merge_df$last_vaf
  name <- as.character(sub_bnpr_wf_merge_df$name)
  clade <- sub_bnpr_wf_merge_df$clade
  bnpr_wf_merge_df$expected_nl[i] <- 1/expected_late_growth(last_vaf,name,clade)
}

bnpr_wf_merge_df %>% 
  subset(col == "Low variance") %>% 
  mutate(expected_late_ratio_wf = late_growth/growth_rate,
         expected_late_ratio = cp_late / expected_nl) %>%
  mutate(expected_late_ratio = ifelse(expected_late_ratio < -1,-1,expected_late_ratio)) %>% 
  ggplot(aes(x = expected_late_ratio_wf < 0.8,y = expected_late_ratio)) +
  geom_jitter(width = 0.3,size = 0.5,alpha = 0.7) +
  geom_boxplot(outlier.alpha = 0) + 
  theme_gerstung(base_size = 6) +
  scale_y_continuous(breaks = c(-1,0,1,2),
                     labels = c("<-1","0","1","2")) + 
  ylab("BNPR late/expected growth ratio") +
  xlab("WF late/expected growth ratio < 0.8")
```

```{r,fig.height=1.5,fig.width=0.7}
bnpr_wf_merge_df %>% 
  subset(col == "Low variance") %>% 
  mutate(expected_late_ratio = cp_late / expected_nl) %>% 
  mutate(expected_late_ratio = ifelse(expected_late_ratio < -1,-1,expected_late_ratio)) %>% 
  ggplot(aes(y = expected_late_ratio,x=0)) + 
  geom_jitter(size = 0.5,alpha = 0.5,width = 0.05) + 
  geom_boxplot(size = 0.25,width = 0.1,outlier.size = 0,alpha = 0.8) + 
  theme_gerstung(base_size = 6) + 
  geom_hline(yintercept = 1,linetype = 2) +
  #coord_cartesian(ylim = c(0,5)) + 
  scale_y_continuous(breaks = c(-1,0,1,2),
                     labels = c("<-1","0","1","2")) + 
  ylab("Observed/expected late growth ratio") +
  xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  ggsave("figures/simulations/expected_late_growth_ratio.pdf",height = 1.5,width = 0.7)
```

## Comparing inference of sigmoidal fit between BNPR fits and driver fits from WF simulations

As mentioned, we also fit sigmoidal curves to each trajectory and here we assess whether they can be used for inference.

```{r,fig.height=3.6,fig.width=4.8}
bnpr_wf_merge_sigmoid_df <- bnpr_wf_merge_df
asymptote_graph <- bnpr_wf_merge_sigmoid_df %>%
  ggplot(aes(x = Asym,y = nl_asym,colour = col)) + 
  geom_point(size = 0.5) + 
  scale_y_continuous(trans = 'log10') +
  scale_x_continuous(trans = 'log10') +
  theme_gerstung(base_size = 6) +
  xlab("Simulated asymptote") +
  ylab("BNPR asymptote") + 
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(NA,1e6)) + 
  scale_color_lancet(name = NULL)

midpoint_graph <- bnpr_wf_merge_sigmoid_df %>%
  ggplot(aes(x = xmid,y = nl_xmid,colour = col)) + 
  geom_abline(slope = 1,size = 0.25,linetype = 2) +
  geom_smooth(method = "lm",size = 0.25,alpha = 0.25) +
  geom_point(size = 0.5) + 
  coord_cartesian(xlim = c(0,2000),ylim = c(0,2000)) + 
  theme_gerstung(base_size = 6) +
  xlab("Simulated midpoint") +
  ylab("BNPR midpoint") + 
  theme(legend.position = "none") + 
  scale_color_lancet(name = NULL)

point_of_saturation_plot <- bnpr_wf_merge_sigmoid_df %>%
  subset(!is.infinite(point_of_saturation) & !is.infinite(point_of_saturation_wf)) %>%
  ggplot(aes(x = point_of_saturation,y = point_of_saturation_wf,colour = col)) + 
  geom_abline(slope = 1,size = 0.25,linetype = 2) +
  geom_smooth(method = "lm",size = 0.25,alpha = 0.25) +
  geom_point(size = 0.5) +
  coord_cartesian(xlim = c(200,800),ylim = c(200,800)) + 
  theme_gerstung(base_size = 6) + 
  xlab("Simulated generation at saturation") +
  ylab("BNPR generation at saturation") + 
  theme(legend.position = "none") + 
  scale_color_lancet(name = NULL)

fitness_plot <- bnpr_wf_merge_sigmoid_df %>%
  ggplot(aes(x = growth_rate,y = nl,colour = col)) +
  geom_abline(slope = 1,size = 0.25,linetype = 2) +
  geom_smooth(method = "lm",size = 0.25,alpha = 0.25) +
  geom_point(size = 0.5) +
  coord_cartesian() + 
  theme_gerstung(base_size = 6) + 
  xlab("Fitness inferred from simulation") +
  ylab("Fitness inferred from BNPR") +
  theme(legend.key.width = unit(0.1,"cm"),
        legend.key.height = unit(0.3,"cm")) + 
  scale_color_lancet(name = NULL) + 
  coord_cartesian(ylim = c(NA,0.05),xlim = c(NA,NA))

plot_grid(
  asymptote_graph,midpoint_graph,point_of_saturation_plot,
  fitness_plot + theme(legend.position = "none"),
  nrow = 2,align = "hv",
  rel_heights = c(1,1)
) %>% 
  plot_grid(plot_grid(ggplot() + theme_nothing(),
                      get_legend(fitness_plot + theme(legend.position = "bottom")),
                      ggplot() + theme_nothing(),nrow = 1),
            rel_heights = c(1,0.1),ncol = 1) + 
  ggsave("figures/simulations/compare_wf_bnpr.pdf",height = 3.6,width = 4.8)

truth <- is.finite(bnpr_wf_merge_df$point_of_saturation_wf)
pred <- is.finite(bnpr_wf_merge_df$point_of_saturation)
acc <- sum(truth == pred) / length(truth)
prec <- sum(truth == pred & truth == 1) / (sum(pred == 1))
recall <- sum(truth == pred & truth == 1) / (sum(truth == 1))

truth_low_var <- is.finite(
  subset(bnpr_wf_merge_df,col == "Low variance")$point_of_saturation_wf)
pred_low_var <- is.finite(
  subset(bnpr_wf_merge_df,col == "Low variance")$point_of_saturation)
acc_low_var <- sum(truth_low_var == pred_low_var) / length(truth_low_var)
prec_low_var <- sum(truth_low_var == pred_low_var & truth_low_var == 1)/(sum(pred_low_var == 1))
recall_low_var <- sum(truth_low_var == pred_low_var & truth_low_var == 1)/(sum(truth_low_var == 1))

data.frame(
  Accuracy = c(acc,acc_low_var),
  Precision = c(prec,prec_low_var),
  Recall = c(recall,recall_low_var),
  Data = c("All trajectories","Low variance trajectories")
)
```

## Benchmarking different models for BNPR fits

Finally we compare all of the methods specified above and make statements regarding their usability in predicting the true fitness of each clade. We note that log-linear fits have a tendency to underestimate the growth rate of clones with large fitness values, whereas sigmoidal curves have a tendency to do the opposite - overestimate the growth rate of clones with small fitness values. The early growth rate stands as the most adequate solution for our problem, with $R^2 > 0.3$ for fits with low variance. We must note that we still observe an $RMSE = 0.01$ which would, assuming 10 generations per year, correspond to a variation of $\pm 10%$. As such, these inferences should account for this considerable level of uncertainty that is associated with each estimate.

```{r,fig.height=5,fig.width=4.7}
all_bnpr_fits <- all_bnpr_fits %>% 
  mutate(col = ifelse(w_sum < 10,"Low variance","High variance")) 

limits <- c(0,0.05)
plot_list <- list()
metric_list <- list()

for (col_name in c("l","cp_1","cp_late","nl")) {
  sub_data <- all_bnpr_fits %>%
    ungroup() %>%
    select(fitness,R = col_name) %>%
    na.omit()
  sub_data_low <- all_bnpr_fits %>%
    subset(col == "Low variance") %>%
    ungroup() %>%
    select(fitness,R = col_name) %>%
    na.omit() 
  
  plot_list[[col_name]] <- all_bnpr_fits %>%
    select(fitness,R = col_name,col = col) %>% 
    ggplot(aes(x = fitness,y = R,group = paste(fitness),colour = col)) +
    geom_jitter(width = 0.001,size = 0.5) +
    geom_boxplot(outlier.alpha = 0,size = 0.25,alpha = 0.7,colour = "black") +
    theme_gerstung(base_size = 6) +
    coord_cartesian(xlim = limits,
                    ylim = limits) +
    geom_abline(slope = 1) +
    ylab("Inferred growth") +
    xlab("Simulated growth") + 
    scale_colour_lancet(guide = F)
  
  metric_list[[col_name]] <- data.frame(
    method = col_name,
    s = c("All trajectories","Low variance"),
    R = c(cor(sub_data$fitness,sub_data$R),
          cor(sub_data_low$fitness,sub_data_low$R)),
    mse = c(mean((sub_data$fitness - sub_data$R)^2),
            mean((sub_data_low$fitness - sub_data_low$R)^2))
  )
}

metric_df <- do.call(rbind,metric_list) %>%
  mutate(
    
    met = list(l = "Linear\ngrowth",
               cp_1 = "Early linear\ngrowth",
               cp_late = "Late linear\ngrowth",
               nl = "Sigmoidal\ngrowth")[method] %>%
      unlist
  )

r2_plot <- ggplot(data = metric_df,aes(x = reorder(met,-R^2),y = R^2,fill = s)) + 
  geom_bar(stat = "identity",position = "dodge") + 
  xlab("") + 
  ylab("R2") + 
  theme_gerstung(base_size = 6) +
  coord_flip() + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_fill_lancet(guide = F)

mse_plot <- ggplot(data = metric_df,aes(x = reorder(met,-mse),y = sqrt(mse),fill = s)) + 
  geom_bar(stat = "identity",position = "dodge") + 
  xlab("") + 
  ylab("RMSE") + 
  theme_gerstung(base_size = 6) +
  coord_flip() + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_fill_lancet(guide = F)

plot_grid(
  plot_list$l + ggtitle("Linear growth"),
  plot_list$cp_1 + ggtitle("Early linear growth"),
  plot_list$cp_late + ggtitle("Late linear growth"),
  plot_list$nl + ggtitle("Sigmoidal growth"),
  r2_plot,
  mse_plot,
  rel_heights = c(1,1,0.6),
  ncol = 2) + 
  ggsave("figures/simulations/validation_BNPR.pdf",height = 5,width = 4.7)

write.csv(bnpr_wf_merge_df,"data_output/simulated_bnpr_coefficients.csv",row.names = F)
```