---
title: "The Natural History and Fitness Landscape of Clonal Haematopoiesis"
subtitle: "Growth coefficient inference and ages at clone foundation"
author: 
    - José Guilherme de Almeida
    - Moritz Gerstung
output: 
  html_document:
    theme: lumen
    highlight: tango
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    number_sections: true
--- 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,dpi = 300)
```

```{r import_functions, include=FALSE}
source("scripts/vaf_dynamics_functions.R")
model_file_name <- "models/model_F5_full.RDS"
dir.create("figures/data_exploration/",showWarnings = F)
dir.create("figures/overdispersion/",showWarnings = F)
set.seed(42)
```

# Problem setting 

We have a cohort with roughly $400$ individuals from Sardinia. These individuals were target-sequenced at a high depth (1000X), which enables us to detect clonal haematopoiesis with a high precision, and at different time points, which in theory enables the modelling of how VAF progress. Ultimately, this enables us to infer the future trajectory of VAF values and determine the effect size underlying this. To model VAF, hierarchical Bayes modelling is used and, in general terms, each VAF is modelled as a binomial distribution whose size is the coverage and probability best described as the linear combination of one or more effects and an offset per individual.

```{r map sardinia,fig.height=8,fig.width=8,fig.align="center",results='hide',fig.keep='all',warning=FALSE,message=FALSE}
m_s <- map_sardinia()
m_s
```

# Data preparation

Data preparation here is concerned mostly with the following aspects:

* Selecting genes or domains based on their significance for our analysis 
* Excluding individuals who have (had) haematological malignancies
* Determining mutations which occur across more than 1 individual
* Defining subsets within our data 

```{r data_preparation}
source("scripts/prepare_data.R")
```

# Data exploration

An initial inspection of the trajectories does not allow us to make any ambitious claims on their dynamics considering age as the only determinant of dynamics. 

Some conclusions can be drawn at this point:

* No apparent single dynamics underlies most VAF trajectories when considering the age of individuals, but this becomes neater when using relative timepoints (relative time instead of absolute time) - this points towards the necessity of an individual-based offset. This also makes the likeliness of other factors (i.e. other mutations) affecting the dynamics of VAF higher
* There is some apparently very similar dynamics for different VAFs within the same individual, which may be a good indicator of the existence of a slope offset on a *per* individual basis for all clones. This may also helps us to decipher any possible interaction there may exist between clones

```{r plot_vaf_trajectories_per_individual, fig.width=14, fig.height=10}
full_data %>% subset(single_occurring == F) %>% 
  mutate(q005=qbeta(p=0.05,MUTcount_Xadj+1,TOTALcount+1-MUTcount_Xadj),
         q095=qbeta(p=0.95,MUTcount_Xadj+1,TOTALcount+1-MUTcount_Xadj)) %>% 
  select(TOTALcount,MUTcount_Xadj,SardID,Age,VAF,q005,q095,Phase,amino_acid_change) %>% 
  
  ggplot(aes(x = Age,y = MUTcount_Xadj/(TOTALcount),ymin = q005,ymax = q095,color = as.factor(SardID))) + 
  geom_linerange(alpha = 0.7) + 
  geom_line() + 
  geom_point() + 
  facet_wrap(~ amino_acid_change,scales = 'free') + 
  theme_gerstung() + 
  scale_color_discrete(guide=FALSE) +
  ylab("VAF") + 
  ggtitle("Trajectory view for each site with more than two mutations (one color per individual)") + 
  ggsave(useDingbats=FALSE,"figures/data_exploration/vaf_trajectories_per_individual.pdf",width=14, height=10)
```

```{r plot_vaf_trajectories_per_site, fig.width=28, fig.height=30}
full_data %>% #subset(single_occurring == F) %>% 
  mutate(q005=qbeta(p=0.05,MUTcount_Xadj+1,TOTALcount+1-MUTcount_Xadj),
         q095=qbeta(p=0.95,MUTcount_Xadj+1,TOTALcount+1-MUTcount_Xadj)) %>% 
  select(TOTALcount,MUTcount_Xadj,SardID,Age,VAF,q005,q095,Phase,amino_acid_change) %>% 
  ggplot(aes(x = Age,y = MUTcount_Xadj/(TOTALcount),ymin = q005,ymax = q095,color = amino_acid_change)) + 
  geom_linerange(alpha = 0.7) + 
  geom_line() + 
  geom_point() + 
  facet_wrap(~ SardID,scales = 'free') + 
  theme_gerstung(base_size=15) + 
  scale_color_discrete(guide=FALSE) +
  ylab("VAF") + 
  ggtitle("Trajectory view for each individual (one color per mutation)") + 
  scale_x_continuous(n.breaks = 4,
                     breaks = function(x) floor(x[1]) + seq(1,x[2],by = round((x[2] - x[1])/5))) + 
  ggsave(useDingbats=FALSE,"figures/data_exploration/vaf_trajectories_per_site.pdf",width=28, height=30)
```

```{r, fig.width=7,fig.height=5}
unpack_square_matrix <- function(square_matrix,V) {
  square_matrix <- as.data.frame(square_matrix)
  colnames(square_matrix) <- V
  square_matrix$labels_2 <- V
  square_matrix %>% 
    gather(key = "labels_1",value = "value",-labels_2) %>%
    return
}

tmp <- full_data %>%
  select(amino_acid_change,SardID,Gene,single_occurring) %>%
  mutate(Gene = str_match(Gene,"[A-Z0-9]+")) %>% 
  distinct()
usm <- full_formatted_data$unique_site_multiple
genes <- names(gene_colours) %>% 
  str_match('[A-Z0-9]+') %>%
  unique %>%
  sort

co_occurrence_genes <- matrix(0,length(genes),length(genes))
co_occurrence_sites <- matrix(0,length(usm),length(usm))
for (individual in unique(full_data$SardID)) {
  individual_data <- tmp %>% 
    subset(SardID==individual) 
  individual_sites <- individual_data %>%
    subset(single_occurring == F) %>% 
    select(amino_acid_change) %>% 
    unlist() %>%
    unique
  individual_genes <- individual_data %>%
    select(Gene) %>% 
    unlist %>% 
    unique
  if (length(individual_sites) >= 2) {
    site_combinations <- t(combn(individual_sites,2))
    for (i in 1:nrow(site_combinations)) {
      x <- site_combinations[i,]
      co_occurrence_sites[usm == x[1],
                          usm == x[2]] <- co_occurrence_sites[usm == x[1],usm == x[2]] + 1
    }
  }
  if (length(individual_genes) >= 2) {
    gene_combinations <- t(combn(individual_genes,2))
    for (i in 1:nrow(gene_combinations)) {
      x <- gene_combinations[i,]
      co_occurrence_genes[genes == x[1],
                          genes == x[2]] <- co_occurrence_genes[genes == x[1],genes == x[2]] + 1
    }
  }
}

data_unique_gene <- full_data %>%
  select(SardID,Gene) %>% 
  distinct
chi_squared_pvalues <- combn(unique(str_match(full_data$Gene,'[A-Z0-9]+')),2) %>% 
  t %>%
  apply(1,function(x) {
    tmp <- data_unique_gene %>% 
      group_by(SardID) %>%
      summarise(
        HasA = and(sum(Gene %in% x[1])>0,sum(Gene %in% x[2])==0),
        HasB = and(sum(Gene %in% x[1])==0,sum(Gene %in% x[2])>0),
        HasNone = and(sum(Gene %in% x[1])==0,sum(Gene %in% x[2])==0),
        HasBoth = and(sum(Gene %in% x[1])>0,sum(Gene %in% x[2])>0)
      ) %>%
      gather("key","value",HasA,HasB,HasNone,HasBoth) %>%
      group_by(key) %>%
      summarise(N = sum(value)) 
    O <- fisher.test(matrix(c(tmp$N[4],tmp$N[2],tmp$N[1],tmp$N[3]),nrow=2)) 
    return(c(O$estimate[1],pValue=O$p.value,
             GeneA=x[1],GeneB=x[2],
             HasGeneA=tmp$N[1],HasGeneB=tmp$N[2],
             HasNone=tmp$N[4],HasBoth=tmp$N[3]))
  }) %>% 
  t %>% 
  as_tibble() %>%
  mutate(`odds ratio` = as.numeric(`odds ratio`),
         pValue = as.numeric(pValue))

chi_squared_pvalues$pValue_adj <- p.adjust(chi_squared_pvalues$pValue)
chi_squared_pvalues %>%
  subset(pValue_adj < 0.05)

total_counts <- full_data %>%
  select(SardID,labels_1=Gene) %>%
  mutate(labels_1 = str_match(labels_1,"[A-Z0-9]+")) %>%
  distinct() %>% 
  group_by(labels_1) %>% 
  summarise(value = length(SardID)) %>%
  mutate(labels_2 = 'Total',labels_1 = labels_1,value = value)

co_occurrence_genes %>%
  unpack_square_matrix(genes) %>%
  merge(total_counts[,-3],by = 'labels_1') %>%
  merge(total_counts[,-3],by.x = 'labels_2',by.y = 'labels_1') %>%
  mutate(
    intersection = value.x,
    union = value.y + value - value.x
  ) %>%
  select(labels_1,labels_2,intersection,union) %>%
  mutate(IoU = intersection / union) %>% 
  mutate(IoU = ifelse(labels_1 == labels_2,1,IoU)) %>% 
  filter(IoU > 0) %>% 
  ggplot(aes(x = labels_1,y = labels_2,fill = IoU,label = round(IoU,2))) +
  geom_tile() + 
  geom_text(aes(colour = IoU),size=3) +
  theme_gerstung(base_size = 15) + 
  rotate_x_text() + 
  scale_fill_material(palette = "deep-purple",
                      name = "Jaccard\nIndex",
                      na.value="white") + 
  scale_colour_material(palette = "purple",
                        na.value="white",
                        trans = "reverse",
                        guide=F) + 
  xlab("") + 
  ylab("") +
  theme(panel.grid = element_blank()) + 
  scale_y_discrete() + 
  ggtitle("Jaccard index for different gene pairs")

co_occurrence_sites %>% max
```

```{r recurrence_prevalence,warning=F,message=F,fig.height=4,fig.width=7}
NMutIndividual <- full_data %>%
  group_by(SardID) %>%
  summarise(N = length(unique(amino_acid_change)))

NMutGeneIndividual <- full_data %>%
  mutate(Gene = str_match(Gene,"[0-9A-Z]+")) %>%
  group_by(SardID,Gene) %>% 
  summarise(N = length(unique(amino_acid_change)))

PrevalenceGenes <- full_data %>%
  mutate(Gene = str_match(Gene,"[0-9A-Z]+")) %>%
  group_by(Gene) %>% 
  summarise(N = length(unique(paste(amino_acid_change,SardID)))) %>%
  ungroup %>%
  mutate(P = N / sum(N))

simulations <- list()
for (i in 1:1000) {
  simulations[[i]] <- NMutIndividual %>%
    apply(1,
      function(x) {
        S <- rmultinom(1,x[2],PrevalenceGenes$P) %>%
          t 
        }
    ) %>% 
    t %>%
    apply(
      1,
      function(x) {
        idx <- which(x > 0)
        return(cbind(Gene = idx,Count = x[idx]))
      }
    ) %>%
    do.call(what = rbind) %>% 
    as.tibble() %>%
    group_by(Gene,Count) %>% 
    summarise(N = length(Gene))
}

simulation_df <- simulations %>% 
  do.call(what = rbind)

ecdf_list <- list()
for (x in 1:length(PrevalenceGenes$Gene)) {
  ecdf_list[[PrevalenceGenes$Gene[x]]] <- S <- simulation_df %>% 
    subset(Gene == x) %>% 
    apply(1,function(x) rep(x[2],x[3])) %>% unlist %>%
    ecdf
}

NTests <- length(unique(paste(NMutGeneIndividual$Gene,NMutGeneIndividual$N)))
NMutGene <- NMutGeneIndividual %>% 
  group_by(Gene,N) %>%
  summarise(Total = length(unique(SardID))) 
  
NMutGeneIndividual %>% 
  rowwise %>% 
  mutate(Prob = 1 - ecdf_list[[Gene]](N)) %>% 
  group_by(Gene,N,Prob) %>%
  summarise(Total = length(unique(SardID))) %>% 
  ungroup() %>% 
  mutate(Significant = ifelse(Prob < 0.05 / NTests,"Yes","No")) %>% 
  ggplot(aes(x = Gene,y = N, shape = Significant)) +
  geom_point(aes(size=Total,color = log(Prob))) + 
  # geom_label(aes(x = Gene,y = N, label = Total),colour = 'black',vjust=-0.5,
  #            label.padding = unit(0,"cm"),label.r = unit(0,"cm"),label.size = unit(0,"cm")) + 
  geom_text(aes(x = Gene,y = N, label = Total),colour = 'black',vjust=-0.6,fontface="bold") + 
  theme_gerstung(base_size = 15) + 
  scale_color_gradient2(low = "red",mid = "grey",high = "lightblue",
                        midpoint = log(0.05 / NTests),na.value = "red",
                        name = 'log(p)') + 
  rotate_x_text() + 
  ylab("Number of unique mutations\nper individual") + 
  scale_size(range = c(2,6),guide = F) +
  scale_shape(name = sprintf("p < %s",0.05 / NTests)) +
  scale_y_continuous(limits = c(1,8.5))
```

```{r summary_figure_first,fig.height=5,fig.width=10}
fc <- function(V) {
  L <- length(V)
  return(V[2:L] / V[1:(L-1)])
}

pd <- position_dodge(0.5)
age_per_phase_plot <- full_data %>%
  group_by(SardID,Phase,Age) %>%
  summarise(
    NMut = length(unique(amino_acid_change))
  ) %>%
  ggplot(aes(x = Phase,y = Age,group = SardID)) + 
  geom_line(alpha = 0.2,
            position=pd) + 
  geom_point(aes(colour = NMut,size = NMut),alpha = 0.5,
             position=pd) + 
  scale_size(name = "No. of\nmutations",
             limits = c(0,10),
             breaks = c(1,3,6,9)) + 
  theme_gerstung(base_size = 15) +
  scale_color_gradient(guide = "legend",
                       low = "goldenrod",
                       high = "brown4",
                       name = "No. of\nmutations",
                       limits = c(0,10),
                       breaks = c(1,3,6,9))

n_mutations_age <- full_data %>% 
  group_by(SardID,Phase,Age) %>%
  summarise(
    NMut = length(unique(amino_acid_change))
  ) %>%
  ggplot(aes(x = Age,y = NMut)) + 
  geom_point(alpha = 0.2,position = position_jitter(height = 0.2)) + 
  geom_smooth(linetype = 2,colour = "red3",size = 1) + 
  theme_gerstung(base_size = 15) +
  scale_y_continuous(breaks = seq(0,10,2)) + 
  ylab("No. of mutations")

mutations_per_gene_time <- load_data_keep() %>%
  mutate(AgeInterval = floor(Age / 10) * 10,
         Gene = str_match(Gene,'[A-Z0-9]+')) %>%
  subset(AgeInterval > 50 & AgeInterval < 100) %>% 
  subset(Gene %in% c("ASXL1","DNMT3A","SF3B1","U2AF1")) %>% 
  group_by(AgeInterval) %>%
  mutate(Total = length(unique(SardID))) %>%
  group_by(Gene,AgeInterval,Total) %>%
  summarise(NMut = length(unique(SardID))) %>% 
  ggplot(aes(x = AgeInterval,y = NMut / Total,group = Gene,colour = Gene)) + 
  geom_line(alpha = 0.6,size = 1) + 
  scale_color_manual(values = gene_colours) + 
  ylab("Prevalnece") + 
  xlab("Age") +
  theme_gerstung(base_size = 15)

N_tp <- 10
trajectory_example_data <- data.frame(
  coefficient = 0.2,
  b_clone = rnorm(N_tp,sd=0.02)
) %>%
  apply(1,FUN=function(x) inv.logit((x[1]+x[2]) * seq(0,15,by = 3)-5)) %>%
  t %>%
  as.data.frame() %>% 
  mutate(ID = as.character(seq(1:N_tp))) %>% 
  gather(key = "key",value = "value",-ID) %>%
  mutate(Time = seq(0,15,by = 3)[as.numeric(str_match(key,"[0-9]"))]) %>% 
  select(-key)
trajectory_example_data_genetic <- data.frame(
  value = inv.logit(0.2 * seq(0,15,by = 3)-5),
  Time = seq(0,15,by = 3)
)

trajectory_example <- trajectory_example_data %>%
  ggplot(aes(x = Time,y = value)) +
  geom_line(aes(group = ID),
            alpha = 0.5) + 
  geom_path(data = trajectory_example_data_genetic,
               colour = "red4",
               size = 1,
               alpha = 0.8,
               lineend = "round"
               ) + 
  xlab("Years since study entry") + 
  ylab("VAF") + 
  theme_gerstung(base_size = 15) + 
  theme() + 
  annotate(x = 0,
           y = max(trajectory_example_data$value)*0.7,
           geom = "text",
           colour = "red4",
           label = "Genetic effect",
           hjust = 0,
           size = 5,
           fontface = "bold") +
  annotate(x = 0,
           y = max(trajectory_example_data$value)*0.6,
           geom = "text",
           colour = "black",
           label = "Individual trajectories",
           hjust = 0,
           size = 5)

plot_grid(age_per_phase_plot,n_mutations_age,mutations_per_gene_time,
          trajectory_example,
          labels = c("A","B","C","D"),
          nrow = 2,rel_widths = c(1,0.8))

## Change A and B to barplot with age brackets (0 mutations should be gray probably)
```

# Modelling 

```{r defining_site_domain_gene_lists}
site_list <- full_formatted_data$unique_site_multiple
domain_list <- full_formatted_data$unique_domain
gene_list <- full_formatted_data$unique_gene
```

```{r creating_result_lists}
r_values_list <- list()
statistics_lists <- list()
```

We model the counts of mutation $i$ in individual $j$ - clone $c_{ji}$ - at age $t$ as the product of a beta-binomial distribution such that $counts_{c_{ji}} \sim BB(cov_{c_{ji}}(t),\alpha_{c_{ji}}(t),\beta)$, where $cov_{c_{ji}}$ is the coverage for mutation $c_{ji}$ at timepoint $t$, $\beta$ is the technical overdispersion and $\alpha = \frac{\beta p}{1 - p(t)}$, where $p(t)$ is the expected value for the counts at age $t$. 

Here we work with a relatively simple formulation of clone growth - for a given clone $i$ in individual $j$ we consider there to be a *genetic* growth effect ($b_{genetic_{i}} = b_{gene_i} + b_{domain_i} + b_{site_i}$), a *mutation-independent* growth effect ($b_{c_{ji}}$) and a *mutation-independent* offset ($u_{ji}$). Only the *growth effects* mentioned here depend on age. These two growth effects and offset are linearly combined and inverse-logit transformed to yield a probability such that $p_{ji}(t) = ilogit(b_{genetic_{i}} = (b_{genetic_{i}} + b_{c_{ji}}) * t + u_{ji})$. Further model specifications are specified below:

$$
b_{gene_i} \sim N(0,1)
\\
b_{site_i} \sim N(0,0.1)
\\
u_{ji} \sim Uniform(-50,0)
\\
b_{clone_{ji}} \sim N(0,0.05)
\\
p_i = ilogit((b_{gene_i} + b_{domain_i} + b_{site_i} + b_{clone_{ji}}) * age + u_{ji})
\\
\beta \sim N(\mu=\mu_{\beta},\sigma=\sigma_{\beta})
\\
\alpha_{SR} = \frac{\beta p}{1 - p}
\\
counts \sim BB(coverage,alpha,beta)
$$

## Modelling the technical overdispersion

Technical replicate data shows some level of technical overdispersion. This can be modelled in our current model by changing our binomial modelling of the counts to a beta binomial, with a fixed $\beta$ parameter and an $\alpha$ parameter that depends on $\beta$ and on the probability that is inferred by modelling, particularly by using the formulation for the expected value of the beta distribution $E[X] = \frac{\alpha}{\alpha+\beta}$. 

Considering the above, we can model the counts as $counts \sim BB(n =  coverage,alpha = \frac{\beta p}{1 - p},beta = \beta)$

```{r visualising overdispersion,fig.height=14,fig.width=14}
one_hot_encode <- function(factor_vector) {
  factor_levels <- sort(as.character(unique(factor_vector)))
  print(factor_levels)
  output <- sapply(
    factor_vector,
    function(x) {
      as.numeric(factor_levels == x)
    }
  )
  return(t(output))
}

tru_q_data <- read.table('data/TruQ.txt',header = T)
replicate_data <- read.table('data/Replicates_all.txt',header = T) %>%
  mutate(VAF = ifelse(CHR == "X", VAF / 2,VAF),
         MUTcount = ifelse(CHR == "X", MUTcount / 2,MUTcount))
ages <- merge(full_data,replicate_data,by = c("SardID","Phase")) %>%
  select(Age, SardID, Phase) %>%
  unique
replicated_data_multiple <- replicate_data %>%
  group_by(SardID,variant) %>%
  summarise(count = length(unique(Phase))) %>%
  subset(count > 1)
replicate_data <- merge(
  replicate_data,
  ages,
  by = c("SardID","Phase"),
  all = T
) %>% 
  subset(SardID %in% replicated_data_multiple$SardID & variant %in% replicated_data_multiple$variant)

tru_q_data %>% 
  ggplot(aes(x = VAF_expected,y = VAF_observed)) + 
  geom_point(aes(colour = as.factor(Replicate))) + 
  geom_errorbar(
    aes(
      ymin = qbeta(0.05,MUTcount + 1,TOTALcount - MUTcount + 1),
      ymax = qbeta(0.95,MUTcount + 1,TOTALcount - MUTcount + 1)
  )) + 
  theme_gerstung(base_size = 15) +
  facet_wrap(~ Gene) + 
  ggsci::scale_color_lancet(name = "Replicate") +
  theme(legend.position = 'bottom') + 
  xlab("VAF expected") +
  ylab("VAF observed") + 
  ggtitle("TruQ data") + 
  ggsave(useDingbats=FALSE,"figures/overdispersion/overdispersion_vis_truq.pdf",height=14,width=14)

replicate_data %>% 
  ggplot(aes(x = as.factor(Phase),y = VAF)) + 
  geom_point(aes(colour = as.factor(Replicate)),
             size = 2) + 
  geom_linerange(
    aes(
      ymin = qbeta(0.05,MUTcount + 1,TOTALcount - MUTcount + 1),
      ymax = qbeta(0.95,MUTcount + 1,TOTALcount - MUTcount + 1)
    )) + 
  theme_gerstung(base_size = 15) +
  facet_wrap(~ paste(Gene,variant,SardID,sep = '-'),scale = 'free') + 
  ggsci::scale_color_lancet(name = "Replicate") +
  theme(legend.position = 'bottom') + 
  xlab("Phase") +
  ylab("VAF") + 
  ggtitle("Replicate data from Sardinian samples") + 
  ggsave(useDingbats=FALSE,"figures/overdispersion/overdispersion_vis_sard.pdf",height=14,width=14)
```

The invervals here are estimated using a beta distribution ($Beta(alpha = count+1,beta = coverage - count + 1)$).

### Estimating $\beta$ with technical replicates

Using technical replicates for known DNA quantities ($truQ$) and for actual patient data ($SR$), we can posit the following model:

$$
p_{truQ} = observed\ VAF_{truQ} + 1/coverage_{truQ}
\\
\beta ~ \sim exponential(0.001)
\\
\alpha_{truQ} = \frac{\beta p_{truQ}}{1 - p_{truQ}}
\\
expected\ VAF_{truQ} + 1/coverage_{truQ} \sim Beta(alpha,beta)
$$

We need to adjust the expected and observed VAFs with the equivalent of having one count due to the fact that beta distributions only permit $\alpha>0$ and $\beta>0$.

$$
p_{SR} = growth\ coefficient = effect_{gene} + individual_{offset}
\\
\alpha_{SR} = \frac{\beta p_{SR}}{1 - p_{SR}}
\\
observed\ counts_{SR} \sim BB(coverage + 1,alpha,beta)
$$

Here, $p_{SR}$ is parametrized as $p_{SR} = ilogit((b_{gene_{i}} + b_{c_{ji}}) * t + u_{ji})$. 

For this subsubsection, the model is specified in `scripts/estimating_overdispersion.R`.

We used this model to infer the $\mu_{\beta}$ and $\sigma_{\beta}$ mentioned in the model specification above.

## Modelling clone growth as a product of genetic and mutation-independent effects

The model specified in `scripts/F2_clone_effect.R` was executed using `scripts/run_model_f2.R`. The model parameters are stored in `models/model_F2_full.RDS` and will be used for the remainder of this section.

```{r model_f2_load_values}
values_model <- readRDS(model_file_name)
model_id <- gsub('_$','',str_match(model_file_name,'model_.*_'))
dir.create("figures/",showWarnings = F)
dir.create(sprintf("figures/%s",model_id),showWarnings = F)
c('age_at_clone_foundation','dynamic_coefficients',
  'inferred_trajectories','overdispersion',
  'statistical_analysis') %>%
  sapply(
    function(x) dir.create(sprintf("figures/%s/%s",model_id,x),showWarnings = F)
  )
```

``` {r model_f2_dynamic_visualization_preparation, fig.width = 3.2,fig.height = 6}
val_subset <- values_model$training_subset

X_val <- val_subset$counts[values_model$gene_idxs,]
gene_idxs <- values_model$gene_idxs

interference_idxs <- values_model$interference_idxs
u_idx <- values_model$u_idx

b_gene_values_val <- values_model$b_gene_values[[1]]$values[values_model$b_gene_values[[1]]$labels == 'mean'] %>% 
  matrix(ncol=1) 
b_site_values_val <- values_model$b_site_values[[1]]$values[values_model$b_site_values[[1]]$labels == 'mean'] %>% 
  matrix(ncol=1) 
beta_val <- values_model$beta_values[[1]]$values[values_model$beta_values[[1]]$labels == 'mean'] %>% 
  matrix(ncol=1) 
b_clone_values_val <- values_model$b_clone[[1]]$values[values_model$b_clone[[1]]$labels == 'mean'] %>% 
  matrix(ncol=1)

u_values_val <- values_model$u_values[[1]]$values[values_model$u_values[[1]]$labels == 'mean'] %>%
  matrix(nrow=1) 

ae_gene <- val_subset$gene_to_site_indicator[values_model$gene_idxs,] %*% b_gene_values_val
ae_site <- val_subset$site_multiple_to_site_indicator[values_model$gene_idxs,] %*% b_site_values_val

age_effect_coef <- ae_gene + ae_site
if ("b_domain_values" %in% names(values_model)) {
  b_domain_values_val <- values_model$b_domain_values[[1]]$values[values_model$b_domain_values[[1]]$labels == 'mean'] %>% 
    matrix(ncol=1) 
  ae_domain <- val_subset$domain_to_site_indicator[values_model$gene_idxs,] %*% b_domain_values_val
  age_effect_coef <- age_effect_coef + ae_domain
}
age_effect <- age_effect_coef[interference_idxs$site] + b_clone_values_val[interference_idxs$ind_site]
age_effect <- age_effect * (val_subset$ages[interference_idxs$ind_age] - values_model$min_age)

offset_per_individual <- u_values_val[interference_idxs$ind_site]
r <- age_effect + offset_per_individual
mu_val <- inv.logit(r) * 0.5

r_values <- data.frame(
  mu_val = mu_val,
  true = val_subset$counts[gene_idxs,][cbind(interference_idxs$site,interference_idxs$ind_age)],
  age = val_subset$ages[interference_idxs$ind_age],
  coverage = val_subset$coverage[gene_idxs,][cbind(interference_idxs$site,interference_idxs$ind_age)],
  individual = val_subset$unique_individual_true[interference_idxs$ind],
  coefficient_095 = values_model$b_values[[1]]$values[values_model$b_values[[1]]$labels == 'HDPI_high'][interference_idxs$site],
  coefficient = values_model$b_values[[1]]$values[values_model$b_values[[1]]$labels == 'mean'][interference_idxs$site],
  coefficient_var = values_model$b_values[[1]]$values[values_model$b_values[[1]]$labels == 'var'][interference_idxs$site],
  coefficient_005 = values_model$b_values[[1]]$values[values_model$b_values[[1]]$labels == 'HDPI_low'][interference_idxs$site],
  site = val_subset$unique_site[values_model$gene_idxs][interference_idxs$site],
  b_clone_005 = c(values_model$b_clone[[1]]$values[values_model$b_clone[[1]]$labels == 'HDPI_low'][interference_idxs$ind_site]),
  b_clone = c(values_model$b_clone[[1]]$values[values_model$b_clone[[1]]$labels == 'mean'][interference_idxs$ind_site]),
  b_clone_095 = c(values_model$b_clone[[1]]$values[values_model$b_clone[[1]]$labels == 'HDPI_high'][interference_idxs$ind_site]),
  u_values_005 = c(values_model$u_values[[1]]$values[values_model$u_values[[1]]$labels == 'HDPI_low'][interference_idxs$ind_site]),
  u_values = c(values_model$u_values[[1]]$values[values_model$u_values[[1]]$labels == 'mean'][interference_idxs$ind_site]),
  u_values_095 = c(values_model$u_values[[1]]$values[values_model$u_values[[1]]$labels == 'HDPI_high'][interference_idxs$ind_site])
) %>%
  as.data.frame() %>%
  apply(1,FUN = function(x){
    mu_val <- as.numeric(x[1])
    cov <- as.numeric(x[4])
    alpha_value <- (beta_val * mu_val) / (1 - mu_val)
    Q <- extraDistr::rbbinom(n = 1000,
                             size = cov,
                             alpha = alpha_value,
                             beta = beta_val) %>%
      quantile(c(0.05,0.50,0.95))
    names(Q) <- c("pred_005","pred","pred_095")
    return(c(x,Q))
  }) %>% 
  t %>%
  as.data.frame() %>% 
  mutate(genes = sapply(as.character(site),function(x) unlist(strsplit(x,'-'))[[1]])) %>%
  group_by(site) %>%
  mutate(n = length(pred)) %>%
  ungroup() %>%
  mutate(
    mu_val = as.numeric(as.character(mu_val)),
    true = as.numeric(as.character(true)),
    age = as.numeric(as.character(age)),
    coverage = as.numeric(as.character(coverage)),
    individual = as.numeric(as.character(individual)),
    coefficient_005 = as.numeric(as.character(coefficient_005)),
    coefficient = as.numeric(as.character(coefficient)),
    coefficient_var = as.numeric(as.character(coefficient_var)),
    coefficient_095 = as.numeric(as.character(coefficient_095)),
    pred_005 = as.numeric(as.character(pred_005)),
    pred = as.numeric(as.character(pred)),
    pred_095 = as.numeric(as.character(pred_095)),
    b_clone_005 = as.numeric(as.character(b_clone_005)),
    b_clone = as.numeric(as.character(b_clone)),
    b_clone_095 = as.numeric(as.character(b_clone_095)),
    u_values_005 = as.numeric(as.character(u_values_005)),
    u_values = as.numeric(as.character(u_values)),
    u_values_095 = as.numeric(as.character(u_values_095))
  )

r_values_ <- r_values %>% 
  mutate(VAFpred = pred / coverage,
         VAFtrue = true / coverage) %>% 
  subset(coverage > 0) %>%
  group_by(individual,site) %>% 
  mutate(normalizedVAFpred = (VAFpred + 1) / (VAFpred[which(age == min(age))] + 1),
         normalizedVAFtrue = (VAFtrue + 1) / (VAFtrue[which(age == min(age))] + 1)) %>%
  ungroup() %>%
  mutate(individual = as.character(individual)) %>% 
  gather(key = 'key',value = 'value',VAFpred,VAFtrue) %>%
  mutate(
    pred_005 = ifelse(key == 'VAFpred',
                      pred_005 / coverage,
                      NA),
    pred_095 = ifelse(key == 'VAFpred',
                      pred_095 / coverage,
                      NA)
  ) 

r_values_list$model_f2 <- r_values

values_model$b_values[[1]] %>% 
  mutate(variable = rep(values_model$training_subset$unique_site[values_model$gene_idxs],each=9)) %>% 
  mutate(recurring = variable %in% values_model$training_subset$unique_site_multiple) %>% 
  write.csv(sprintf("site_coefficients_%s.csv",model_id),quote = F,row.names = F)
parameter_b_clone <- data.frame(
  mean = values_model$b_clone[[1]][values_model$b_clone[[1]]$labels == 'mean',][values_model$interference_idxs$ind_site,]$values,
  q0.05 = values_model$b_clone[[1]][values_model$b_clone[[1]]$labels == '0.05',][values_model$interference_idxs$ind_site,]$values,
  q0.95 = values_model$b_clone[[1]][values_model$b_clone[[1]]$labels == '0.95',][values_model$interference_idxs$ind_site,]$values,
  q0.50 = values_model$b_clone[[1]][values_model$b_clone[[1]]$labels == '0.50',][values_model$interference_idxs$ind_site,]$values,
  var = values_model$b_clone[[1]][values_model$b_clone[[1]]$labels == 'var',][values_model$interference_idxs$ind_site,]$values,
  HDPI_low = values_model$b_clone[[1]][values_model$b_clone[[1]]$labels == 'HDPI_low',][values_model$interference_idxs$ind_site,]$values,
  HDPI_high = values_model$b_clone[[1]][values_model$b_clone[[1]]$labels == 'HDPI_high',][values_model$interference_idxs$ind_site,]$values,
  individual = val_subset$unique_individual_true[interference_idxs$ind],  
  site = val_subset$unique_site[values_model$gene_idxs][interference_idxs$site]
) %>%
  mutate(recurring = site %in% values_model$training_subset$unique_site_multiple) %>% 
  distinct 
parameter_b_clone %>% 
  write.csv(sprintf("mutation_independent_coefficients_%s.csv",model_id),quote = F,row.names = F)

values_model$b_gene_values[[1]] %>% 
  mutate(variable = rep(values_model$training_subset$unique_gene,each=9)) %>% 
  write.csv(sprintf("gene_coefficients_%s.csv",model_id),quote = F,row.names = F)


explained_variance_total <- r_values %>%
  summarise(Genetic = rowSums(cov(data.frame(gene=coefficient*(age-values_model$min_age),
                                             clone=b_clone*(age-values_model$min_age)))) %>% 
              (function(x) x[1] / sum(x))) %>%
  unlist

variation_explained_plot <- r_values %>% 
  mutate(genes = str_match(genes,'[A-Z0-9]+')) %>% 
  group_by(genes) %>% 
  summarise(Genetic = rowSums(cov(data.frame(gene=coefficient*(age-values_model$min_age),
                                             clone=b_clone*(age-values_model$min_age)))) %>% (function(x) x[1]),
            Unknown = rowSums(cov(data.frame(gene=coefficient*(age-values_model$min_age),
                                             clone=b_clone*(age-values_model$min_age)))) %>% (function(x) x[2])
            ) %>% 
  mutate(Genetic = Genetic / (Genetic + Unknown),
         Unknown = Unknown / (Genetic + Unknown)) %>% 
  gather("key" = key,"value",-genes) %>% 
  mutate(model_name = paste0("Genetic-associated\ninter-individual\nvariation = ",100*round(explained_variance_total,3),'%')) %>%
  mutate(key = factor(key,levels = rev(unique(key)))) %>% 
  ggplot(aes(x = genes,y = value,fill = key)) + 
  geom_bar(stat = "identity",color = "black",position = "dodge") + 
  theme_gerstung(base_size = 15) + 
  scale_fill_manual(values = rev(pal_jama()(2)),name = NULL,breaks = c("Genetic","Unknown")) + 
  #rotate_x_text() + 
  xlab("Driver gene") + 
  ylab("Fraction") + 
  facet_grid(~ model_name) + 
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5),
        legend.margin=margin(0,0,0,0),
        strip.text = element_text(size = 15),
        axis.text.y = element_text(face = 'italic'),
        legend.box.margin=margin(0,0,0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0,0.01,0),
                     limits = c(0,1),
                     breaks = c(0,0.2,0.4,0.6,0.8,1.0)) +
  coord_flip() + 
  guides(fill=guide_legend(nrow=2,byrow=TRUE))

ggsave(useDingbats=FALSE,filename = sprintf("figures/%s/statistical_analysis/variation_explained.pdf",model_id),
       variation_explained_plot,height = 6,width = 3.2)
ggsave(filename = sprintf("figures/%s/statistical_analysis/variation_explained.svg",model_id),
       variation_explained_plot,height = 6,width = 3.2)
variation_explained_plot
```

A simple analysis of the coefficients reveals to us quite a hopeful result when we consider this information from a clinical perspective - 67.7% of the model's explained variation can be accounted by genetics. Looking at specific mutations we have quite a varied picture in these regards - more than 90% of the variation of CBL and U2AF1 trajectories can be explained by genetics, whereas more than 80% of the variation of IDH1, SRSF2 and TP53 trajectories can be explained by genetics.

### Assessing the goodness of fit with the residual effect

We here define the residual effect (RE) as $RE = 1 - s$, with $s=CDF_{\chi^2}(\sum_{i=1}^{n}Q_{N(0,1)}(CDF_{BBinom(coverage,\alpha_{inferred},\beta{inferred})}(counts))^2)$. We draw a cutoff at 0.7, considering trajectories with $RE < 0.7$ as good trajectories and those with $RE \get 0.7$ as bad trajectories. Bad trajectories are excluded from any downstream analysis. More visually, $RE > 0.7$ generally means that at least one of the datapoints for a trajectory is not contained within a 90% HPDI of the inferred trajectory. As such, we can also think of the $RE$ as a trajectory's "outlierness" and classify trajectories with $RE > 0.7$ as outliers.

```{r model_f2_summary_statistic, fig.height=5,fig.width=6}
prediction_bbinom <- function(effect,age,u,coverage,beta) {
  p <- inv.logit(effect * age + u) / 2
  alpha <- (beta*p) / (1-p)
  rbbinom(1000,
          size = coverage,
          alpha = alpha,
          beta = beta) %>%
    mean %>% 
    return
}

order_same_rank <- function(v) {
  u <- unique(v)
  o <- rep(0,length(v))
  for (i in 1:length(u)) {
    o[v == u[i]] <- i
  }
  return(o)
}

beta_value <- values_model$beta_values[[1]][1,1]

statistic_full <- r_values_ %>%
  subset(key == 'VAFpred') %>%
  mutate(
    mu_val_005 = inv.logit((coefficient_005) * (age-values_model$min_age) + u_values) / 2,
    mu_val_095 = inv.logit((coefficient_095) * (age-values_model$min_age) + u_values) / 2
  ) %>% 
  mutate(
    tail_prob = pbbinom(
      q = true,
      size = coverage,
      alpha = (mu_val * beta_value)/(1 - mu_val),
      beta = beta_value),
    tail_prob_005 = pbbinom(
      q = true,
      size = coverage,
      alpha = (mu_val_005 * beta_value)/(1 - mu_val_005),
      beta = beta_value),
    tail_prob_095 = pbbinom(
      q = true,
      size = coverage,
      alpha = (mu_val_095 * beta_value)/(1 - mu_val_095),
      beta = beta_value)
    ) %>%
  group_by(site,individual,genes) %>% 
  summarise(sum_stat = pchisq(sum(qnorm(tail_prob)^2),length(tail_prob)),
            sum_stat_005 = pchisq(sum(qnorm(tail_prob_005)^2),length(tail_prob)),
            sum_stat_095 = pchisq(sum(qnorm(tail_prob_095)^2),length(tail_prob))
            ) %>%
  mutate(split = 'full') %>%
  select(split,site,sum_stat,sum_stat_005,sum_stat_095,gene = genes,individual) %>%
  ungroup() %>%
  mutate(recurring = site %in% values_model$training_subset$unique_site_multiple)

total_variants_explained <- statistic_full %>%
  group_by(site) %>%
  summarise(TotalExplainedVariants = sum(sum_stat < 0.7),
            TotalVariants = length(sum_stat),
            TotalExplainedVariants005 = sum(sum_stat_005 < 0.7),
            TotalExplainedVariants095 = sum(sum_stat_095 < 0.7))

statistic_aggregated <- r_values_ %>%
  subset(key == 'VAFpred') %>%
  mutate(
    tail_prob = extraDistr::pbbinom(
    q = true,
    size = coverage,
    alpha = (mu_val * beta_value)/(1 - mu_val),
    beta = beta_value)) %>%
  group_by(site) %>% 
  summarise(sum_stat = pchisq(sum(qnorm(tail_prob)^2),length(tail_prob)),
            gene = genes[1],
            MinimumAge = min(age),
            MaximumAge = max(age),
            MaxValue = max(c(pred_095,value,true/coverage),na.rm = T)) %>%
  mutate(split = 'full') %>%
  ungroup() %>%
  merge(total_variants_explained,by = c("site"))

p <- statistic_full %>%
  ggplot(aes(x = gene,y = sum_stat)) +
  geom_point(aes(color = recurring),
             stat = 'identity',
             position = position_jitter(width = 0.1),
             alpha = 0.3) + 
  rotate_x_text() +
  theme_gerstung() +
  theme(legend.position = 'bottom',
        axis.text.x = element_text(face = 'italic')) + 
  xlab("") +
  ylab("Summary statistic") + 
  scale_y_continuous() + 
  geom_hline(yintercept = 0.7,alpha = 0.5) +
  scale_color_manual(name = NULL,
                     values = c("black","grey80"),
                     labels = c("Good fit","Poor fit")) +
  rotate_x_text()

p

statistic_full %>% 
  select(sum_stat) %>%
  mutate(sum_stat = sum_stat < .7) %>%
  table

statistics_lists$model_f2 <- list(
  individual_site = statistic_full,
  site = statistic_aggregated
)
```

### Visualising example trajectories

```{r model_f2_dynamic_visualization_all,fig.width = 30,fig.height = 30}
r_values_merged <- merge(statistic_full,
                         r_values_,
                         all_y = F,
                         all.x = F,
                         by.x = c("site","individual"),
                         by.y = c("site","individual")) %>%
  mutate(site_plot = sapply(site,function(x) str_split(x,';')[[1]][1])) %>%
  mutate(site_plot = gsub('-','\n',site_plot))

average_statistic <- r_values_merged %>% 
  group_by(site) %>%
  summarise(AverageStatistic = mean(sum_stat),
            TotalExplainedVariants = sum(sum_stat < 0.7),
            MinimumAge = min(age),
            MaximumAge = max(age),
            MaxValue = max(c(pred_095,value),na.rm = T)) 

r_values_merged %>% 
  ggplot(aes(x = age,y = value)) +
  geom_ribbon(alpha = 0.1,
              data = r_values_merged %>%
                subset(key == 'VAFpred'),
              aes(ymin = pred_005,
                  ymax = pred_095,
                  group = as.factor(paste0(as.character(individual),key))),
              na.rm = T) +
  geom_hline(aes(yintercept = (1/2e5) * (10/coefficient))) +
  geom_line(aes(colour = key,group = as.factor(paste0(as.character(individual),key)))) +
  facet_wrap(~ site_plot,scales = "free") + 
  scale_color_manual(name = NULL,
                     values = c("black","grey80"),
                     labels = c("Good fit","Poor fit")) +
  theme_gerstung(base_size = 10) + 
  theme(legend.position = 'bottom',
        axis.text.x = element_text(size=7)) +
  xlab("Age") + 
  ylab("VAF") + 
  scale_x_continuous(n.breaks = 4,
                     breaks = function(x) floor(x[1]) + seq(1,x[2],by = round((x[2] - x[1])/4))) +
  ggtitle("All inferred (blue) and observed (red) trajectories") +
  ggsave(useDingbats=FALSE,sprintf("figures/%s/inferred_trajectories/plot_all_trajectories.pdf",model_id),
         width = 30,height = 30)
```

```{r model_dynamic_visualization,warning=F,fig.width = 10.8,fig.height = 6}
ss <- statistic_aggregated %>%
  subset(sum_stat <= 0.05) %>%
  select(site) %>%
  unlist
#site_subset <- sort(val_subset$unique_site_multiple)[c(16,20,21,22,26,29,32,38)]
site_subset <- c("DNMT3Ant-R882H","IDH1-R132H","IDH2-R140Q","JAK2-V617F",
                 "SF3B1-K666N","SRSF2-P95H","TET2nt-I1873T","U2AF1-Q157P")

r_values_merged %>%
  pivot_wider(names_from = c("key"),values_from = c("pred_005","pred","pred_095")) %>%
  select(individual,site,sum_stat,true,age,coverage,genes,n,value,site_plot,
         pred_005 = pred_005_VAFpred,
         pred = pred_VAFpred,
         pred_095 = pred_095_VAFpred) %>%
  na.omit() %>% 
  subset(site %in% site_subset) %>%
  group_by(individual,site) %>%
  mutate(N = length(unique(age))) %>%
  #filter(N > 2) %>% 
  ungroup() %>% 
  #group_by(individual) %>% 
  mutate(PROJECTION = prcomp(data.frame(x = age,y = pred,z = sum_stat),center=T,scale=T)$x[,1]) %>% 
  group_by(individual,site) %>% 
  mutate(PROJECTION = max(PROJECTION)) %>% 
  group_by(site) %>%
  filter(PROJECTION %in% c(
    min(PROJECTION),max(PROJECTION),quantile(PROJECTION,0.5,type = 1)
    )) %>%
  mutate(site_plot = gsub(x = site_plot,'\n','-')) %>%
  mutate(site_plot = gsub('nt|t','',site_plot)) %>% 
  ggplot(aes(x = age,y = (pred)/coverage)) +
  geom_ribbon(alpha = 0.1,
              aes(ymin = pred_005,
                  ymax = pred_095,
                  group = as.factor(as.character(individual))),
              na.rm = T) +
  geom_line(aes(colour = sum_stat > 0.7,
                group = as.factor(as.character(individual))),
            size = 0.7,
            alpha = 0.8) +
  geom_point(aes(y = (true)/coverage),
             alpha=0.4) +
  geom_line(aes(colour = sum_stat > 0.7,
                y = (true)/coverage,
                group = as.factor(as.character(individual))),
            alpha=0.4,
            linetype=2,
            size = 0.7) +
  facet_wrap(~ site_plot,
             scales = "free",
             ncol = 4) + 
  # geom_label(
  #   label.r = unit(0,"cm"),
  #   label.size = unit(0,"cm"),
  #   label.padding = unit(0,"cm"),
  #   alpha = 0.7,
  #   data = subset(statistic_aggregated,site %in% site_subset),
  #   size = 3,
  #   aes(#group = site,
  #       x = MinimumAge + (MaximumAge - MinimumAge) * 0.35,
  #       y = MaxValue - MaxValue * 0.1,
  #       label = paste0('Explained\nvariants = ',format(100 * TotalExplainedVariants/TotalVariants,digits = 2),'%')),
  #   fontface = "italic") +
  theme_gerstung(base_size = 15) + 
  theme(legend.position = 'bottom',
        strip.text = element_text(size = 12),
        legend.key.height = unit(0,"cm")) +
  scale_y_continuous(trans = 'log10',
                     limits = c(5e-4,0.5)) + 
  xlab("Age") + 
  ylab("VAF") +
  scale_alpha(guide = F,range = c(0.4,1.0)) +
  ggsave(useDingbats=FALSE,sprintf("figures/%s/inferred_trajectories/Figure4_smaller.pdf",model_id),
         width = 10.8,height = 6) + 
  ggsave(useDingbats=FALSE,sprintf("figures/%s/inferred_trajectories/Figure4_smaller.pdf",model_id),
         width = 10.8,height = 6)
```

Observing the above trajectories gives a better insight into the effectiveness of our model in capturing trajectories with very distinct clonal dynamics.

```{r model_dynamic_visualization_genetic_only,fig.width=14,fig.height=8,message=F,warning=F}
pop_size <- 2e5
r_values %>% 
  select(coefficient,b_clone,site,genes,u_values) %>% 
  distinct %>% 
  subset(site %in% values_model$training_subset$unique_site_multiple) %>% 
  mutate(site = paste0(str_match(genes,'[0-9A-Z]+'),str_match(site,'-[A-Z0-9a-z]+'))) %>%
  mutate(full_effect = coefficient + b_clone) %>%
  apply(1,FUN = function(x) {
    data.frame(
      samples_genetic = inv.logit(seq(0,20) * as.numeric(x[1]) + as.numeric(x[5])),
      samples = inv.logit(seq(0,20) * as.numeric(x[6]) + as.numeric(x[5])),
      site = x[3],
      genes = x[4],
      x = seq(0,20),
      coefficient = x[1],
      full_effect = x[6]
    ) %>%
      return
  }) %>% 
  do.call(what = rbind) %>%
  group_by(site,full_effect) %>%
  mutate(samples = samples / samples[which.min(x)],
         samples_genetic = samples_genetic / samples_genetic[which.min(x)]) %>% 
  group_by(x,site) %>%
  mutate(samples_genetic = mean(samples_genetic)) %>% 
  ungroup() %>% 
  mutate(site = factor(site,levels = sort(as.character(unique(site))))) %>% 
  ggplot(aes(x = x,colour = genes)) +
  geom_line(aes(y = samples,group = full_effect),alpha = 0.7) + 
  geom_line(aes(y = samples_genetic,group = site),linetype = "longdash",size = 0.5,color = 'black') + 
  facet_wrap(~ site,scales = "free") +
  theme_gerstung(base_size = 15) +
  scale_y_continuous(trans = 'log10',breaks = 10^seq(0,10),labels = c("1x","10x","100x",paste0(1,'e',seq(3,10),'x')),
                     limits = c(1,1e4)) + 
  scale_colour_manual(values = gene_colours,guide = F) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks = element_line(size = 0.1)) +
  xlab("Time (years)") + 
  ylab("Fold increase in VAF")
```

### Inspecting coefficients

In this subsection we inspect coefficients for all genomic resolutions (gene, domain and site) individually and then observe the general picture of their combined effects. The most relevant aspects are the wide range of dynamics observable for all genes and the lack of evidence for a strong domain signal. Additionally, sites in the same gene do not appear to have in most cases a more or less pronounced signal - a clear exception of this is P95H in SRSF2, which grows much faster than mutations on the site site for different amino acids (P95A and P95L). The frameshift in T483fs in PPM1D appears to produce clones which grow faster than other mutations on PPM1d but its large credible interval prevents us from claiming anything with a higher degree o certainty. Mutations in U2AF1 and IDH1 are unusually fast growers when compared with other genes. Another striking aspect of this analysis is how the prevalence of mutations is quite a poor predictor of their clonal dynamics - JAK2 and SF3B1 mutations, while being quite prevalent, do not produce clones which grow particularly fast.

```{r model_f2_gene_coefficients, fig.width=14, fig.height=10}
values_b_gene <- values_model$b_gene_values[[1]]
values_b_gene$site_name <- rep(val_subset$unique_gene,
                               each = length(unique(values_model$b_gene_values[[1]]$labels)))
values_b_gene %>% spread(key = labels,value = values) %>%
  mutate(Truncating = grepl("[A-Z0-9]+t",site_name),
         site_name = str_match(site_name,"[A-Z0-9]+")) %>%
  ggplot(aes(y = mean,x = site_name,color = site_name,shape = Truncating)) +
  geom_point(size = 2,
             position = position_dodge(width = 1.0)) +
  geom_linerange(aes(ymin = HDPI_low,ymax = HDPI_high),
                 position = position_dodge(width = 1.0)) + 
  geom_hline(yintercept = 0,alpha = 0.2) +
  theme_gerstung(base_size = 15) +
  rotate_x_text() + 
  xlab("Driver gene") + 
  ylab("Effect size") + 
  facet_grid(~ site_name,scales = "free_x",space = "free_x") +
  scale_color_manual(values = gene_colours,
                     guide = F) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title = element_text(size = 15),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        strip.text = element_blank(),
        axis.text.x = element_text(face = 'italic'),
        legend.key.height = unit(0,"cm")
        ) + 
  scale_shape_discrete(breaks = c(TRUE,FALSE),
                       labels = c("Truncating","Non-truncating"),
                       name = NULL) +
  ggsave(useDingbats=FALSE,sprintf("figures/%s/dynamic_coefficients/FigureGene.pdf",model_id),
         height = 3.5,width = 8)
```

```{r model_f2_domain_coefficients, fig.width=14, fig.height=10}
if ("b_domain_values" %in% names(values_model)) {
  values_b_domain <- values_model$b_domain_values[[1]]
  values_b_domain$site_name <- rep(val_subset$unique_domain,
                                   each = length(unique(values_model$b_values[[1]]$labels)))
  values_b_domain$gene <-  sapply(values_b_domain$site_name,function(x) (strsplit(x,'-') %>% unlist)[1])
  values_b_domain %>% spread(key = labels,value = values) %>%
    ggplot(aes(y = `0.50`,x = substr(site_name,1,30),color = gene)) +
    geom_point() +
    geom_linerange(aes(ymin = `0.05`,ymax = `0.95`)) +
    geom_hline(yintercept = 0,alpha = 0.2) +
    theme_gerstung() +
    rotate_x_text() +
    xlab("") +
    ylab("Effect size") +
    facet_grid(~ gene,scales = "free_x",space = "free_x") +
    scale_color_manual(values = gene_colours,
                       guide = F) +
      theme(legend.position = "bottom",
          panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank(),
          strip.text = element_text(size = 5.7,face = "bold")) +
    ggsave(useDingbats=FALSE,sprintf("figures/%s/dynamic_coefficients/FigureDomain.pdf",model_id),
           height = 3.5,width = 8)
}
```

```{r model_f2_site_coefficients, fig.width=14, fig.height=10}
values_b_site <- values_model$b_site_values[[1]]
values_b_site$site_name <- rep(val_subset$unique_site_multiple,
                               each = length(unique(values_model$b_values[[1]]$labels)))
values_b_site$gene <-  sapply(values_b_site$site_name,function(x) (strsplit(x,'-') %>% unlist)[1])

values_b_site %>% spread(key = labels,value = values) %>%
  ggplot(aes(y = mean,x = substr(site_name,1,30),color = gene)) +
  geom_point() +
  geom_linerange(aes(ymin = HDPI_low,ymax = HDPI_high)) + 
  geom_hline(yintercept = 0,alpha = 0.2) +
  theme_gerstung() +
  rotate_x_text() + 
  xlab("") + 
  ylab("Effect size") + 
  facet_grid(~ gene,scales = "free_x",space = "free_x") +
  scale_color_manual(values = gene_colours,
                     guide = F) +
    theme(legend.position = "bottom",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        strip.text = element_text(size = 5.7,face = "bold")) +
  ggsave(useDingbats=FALSE,sprintf("figures/%s/dynamic_coefficients/FigureSite.pdf",model_id),
         height = 3.5,width = 8)
```

```{r model_f2_full_effects, fig.width=10, fig.height=6,warning=F}
min_max_plots <- c(min(c(r_values$coefficient_005,r_values$b_clone)),max(c(r_values$coefficient_095,r_values$b_clone)))
min_max_plots <- exp(min_max_plots) * 100 - 100
min_max_plots_growth <- sign(min_max_plots) * ceiling(abs(min_max_plots))
min_max_plots_round <- round(min_max_plots,-1)

gene_correspondence_truncation <- values_model$training_subset$full_data %>%
  group_by(Gene) %>%
  summarise(Truncating = unique(truncating)) %>%
  select(gene = Gene,Truncating)
  
values_b <- values_model$b_values[[1]]
values_b$site_name <- rep(val_subset$unique_site[values_model$gene_idxs],
                          each = length(unique(values_model$b_values[[1]]$labels)))
values_b$gene <- values_b$site_name %>% 
  sapply(function(x) unlist(str_split(x,pattern = '-'))[1])
values_b <- values_b

gene_values <- values_model$b_gene_values[[1]] %>%
  mutate(
    gene = rep(val_subset$unique_gene,
               each = length(unique(values_model$b_gene_values[[1]]$labels)))
  ) %>%
  subset(labels == "mean") 
gene_values[gene_values$gene == 'JAK2',] <- subset(values_b,gene == 'JAK2' & labels == '0.50')[c(1,2,3,5)]
gene_values[gene_values$gene == 'IDH2',] <- subset(values_b,gene == 'IDH2' & labels == '0.50')[c(1,2,3,5)]
gene_values <- gene_values %>% 
  mutate(gene_pure = str_match(gene,"[A-Z0-9]+")) %>%
  group_by(gene_pure) %>%  
  mutate(gene_average = mean(values)) %>%
  arrange(gene_average,values) %>%
  ungroup() %>% 
  select(gene_average = values,gene,gene_pure) 
gene_order <- gene_values %>% 
  select(gene_pure) %>%
  distinct %>% 
  unlist
sub_gene_order <- gene_order[gene_order %in% str_match(values_model$training_subset$unique_site_multiple,'[A-Z0-9a-z]+')]
gene_values[gene_values$gene == 'JAK2',] <- list(NA,'JAK2','JAK2')
gene_values[gene_values$gene == 'IDH2',] <- list(NA,'IDH2','IDH2')

gene_intervals <- values_model$b_gene_values[[1]] %>%
  mutate(
    gene = rep(val_subset$unique_gene,
               each = length(unique(values_model$b_gene_values[[1]]$labels)))
  ) %>%
  subset(labels %in% c("HDPI_high","HDPI_low")) %>% 
  spread(key = labels,value = values) %>%
  mutate(Truncating = grepl("[A-Z0-9]+t",gene)) %>%
  mutate(gene = str_match(gene,'[A-Z0-9]+')) %>% 
  select(-variable) %>%
  mutate(Truncating = ifelse(gene %in% c("ASXL1","PPM1D"),T,Truncating))

mutation_aa_order <- values_b %>%
  subset(site_name %in% val_subset$unique_site_multiple) %>%
  mutate(site_name = sapply(site_name,function(x) unlist(strsplit(x,'-'))[2])) %>%
  mutate(site_name = sapply(site_name,function(x) unlist(strsplit(x,';'))[1])) %>%
  mutate(site_name = sapply(site_name,function(x) {
    tmp <- unlist(strsplit(x,':'))
    return(tmp[length(tmp)])})) %>%
  subset(labels == "mean") %>% 
  arrange(values) %>%
  select(site_name) %>%
  unlist

Number_Of_Mutations <- r_values %>%
  select(individual,site) %>%
  distinct() %>%
  group_by(site) %>%
  summarise(n = length(site)) %>%
  ungroup() %>%
  mutate(site_name = site) %>%
  select(-site)

plot_growth_per_year_data <- values_b %>% 
  merge(Number_Of_Mutations,by = 'site_name') %>%
  subset(site_name %in% val_subset$unique_site_multiple) %>%
  mutate(Truncating = grepl("[A-Z0-9]+t",gene)) %>%
  spread(key = labels,value = values) %>%
  group_by(gene) %>%
  mutate(mean = exp(mean)*100 - 100,
         HDPI_low = exp(HDPI_low)*100 - 100,
         HDPI_high = exp(HDPI_high)*100 - 100) %>%
  merge(gene_values,by = "gene",all.y=T) %>% 
  group_by(gene) %>%
  mutate(gene_average = as.numeric(gene_average)) %>% 
  mutate(gene_average = ifelse(is.na(gene_average),NA,exp(gene_average)*100 - 100)) %>% 
  ungroup() %>% 
  mutate(site_name = sapply(site_name,function(x) unlist(strsplit(x,'-'))[2])) %>%
  mutate(site_name = sapply(site_name,function(x) unlist(strsplit(x,';'))[1])) %>%
  mutate(site_name = sapply(site_name,function(x) {
    tmp <- unlist(strsplit(x,':'))
    return(tmp[length(tmp)])})) %>%
  select(-Truncating) %>% 
  mutate(site_name = ifelse(is.na(site_name),' ',site_name)) %>% 
  merge(gene_correspondence_truncation,by = "gene") %>%
  mutate(gene = factor(as.character(gene_pure),levels = gene_order,
                       ordered = TRUE)) %>%
  mutate(Truncating_facet = ifelse(Truncating,'Trunc.','Non-Trunc.')) %>% 
  mutate(site_name = ifelse(site_name == " ",Truncating_facet,site_name)) %>% 
  group_by(gene,Truncating)

Segment_order <- plot_growth_per_year_data %>% 
  mutate(SiteNumeric = as.numeric(factor(site_name,levels = site_name[order(`0.50`)]))) %>%
  group_by(gene) %>%
  mutate(NT = sum(Truncating==F)) %>%
  group_by(gene,Truncating) %>%
  summarise(SiteNumericMax = max(SiteNumeric),
            SiteNumericMin = min(SiteNumeric),
            NT = NT[1],
            gene_average = gene_average[1]) %>%
  mutate(SiteNumericMax = ifelse(Truncating == T,SiteNumericMax + NT,SiteNumericMax),
         SiteNumericMin = ifelse(Truncating == T,SiteNumericMin + NT,SiteNumericMin)) %>%
  merge(gene_intervals,by = c("gene","Truncating"))

x_offset_pointrange <- -0.2
y_offset_pointrange <- 0
legend_elements <- list(
  pointrange = data.frame(
    x = 1.5 + x_offset_pointrange,
    ymin = 35 + y_offset_pointrange,
    ymax = 45 + y_offset_pointrange,
    y = 40  + y_offset_pointrange,
    gene = sort(plot_growth_per_year_data$gene)[1]
  ),
  text = data.frame(
    x = c(1.5,1.5,1.5) + 1 + x_offset_pointrange,
    y = c(35,40,45) + y_offset_pointrange,
    label = c(as.character('"HDPI"^"low"'),
              "Mean",
              as.character('"HDPI"^"high"')),
    gene = sort(plot_growth_per_year_data$gene)[1]
  ),
  rect = data.frame(
    xmin = 0.5 + x_offset_pointrange,
    xmax = 0.5 + table(plot_growth_per_year_data$gene)[1] + x_offset_pointrange,
    ymin = 30,
    ymax = 50 + y_offset_pointrange,
    gene = sort(plot_growth_per_year_data$gene)[1]
  )
)

plot_growth_per_year <- plot_growth_per_year_data %>% 
  ggplot(aes(color = gene)) +
  #geom_point(aes(colour = gene,y = gene_average),shape=95,size=9,alpha = 1.0) +
  geom_linerange(aes(ymin = HDPI_low,ymax = HDPI_high,
                     x = reorder(site_name,`0.50`+Truncating*1000))) + 
  geom_point(aes(size = n,
                 y = mean,
                 x = reorder(site_name,`0.50`+Truncating*1000))) +
  geom_hline(yintercept = 0,alpha = 0.2) +
  geom_segment(
    data = Segment_order,
    aes(x = SiteNumericMin-0.5,xend = SiteNumericMax+0.5,
        y = gene_average,yend = gene_average),
    alpha=0.7) +
  geom_rect(
    data = Segment_order,
    inherit.aes = F,
    aes(xmin = SiteNumericMin-0.5,xmax = SiteNumericMax+0.5,
        ymin = exp(HDPI_high)*100-100,ymax = exp(HDPI_low)*100-100,
        fill = gene),colour = NA,
    alpha = 0.2) +
  # begin legend for pointrange
  geom_rect(
    data = legend_elements$rect,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = ymin,
      ymax = ymax
    ),
    fill = "white",
    colour = NA
  ) +
  geom_pointrange(size = 0.5,
                  colour = "black",
                  data = legend_elements$pointrange,
                  aes(x = x,y = y,ymax = ymax,ymin = ymin)) +
  geom_text(
    data = legend_elements$text,
    hjust = 0,
    parse = T,
    colour = 'black',
    aes(x = x,y = y, label = label)
  ) + 
  # end legend for pointrange
  theme_gerstung(base_size = 15) +
  rotate_x_text() + 
  xlab(" ") + 
  ylab("Growth per year") + 
  facet_grid(~ gene,scales = "free_x",space = "free_x") +
  scale_color_manual(values = gene_colours,
                     guide = F) +
  scale_fill_manual(values = gene_colours,
                    guide = F) +
  theme(legend.position = c(0.09,0.84),
        legend.background = element_rect(colour = NA,fill = "white"),
        #panel.grid.major.x = element_blank(),
        #panel.grid.minor.x = element_blank(),
        panel.grid = element_blank(),
        strip.text = element_text(size = 11,face = "bold.italic"),
        legend.key.height = unit(0,"cm"),
        # strip.background = element_blank(),
        # plot.background = element_blank(),
        # panel.background = element_blank(),
        axis.text.x = element_text(face = "plain")) +
  #coord_cartesian(ylim = c(-10,60)) + 
  scale_y_continuous(breaks = seq(min_max_plots_round[1],min_max_plots_round[2],by = 20),
                     labels = paste0(seq(min_max_plots_round[1],min_max_plots_round[2],by = 20),'%'),
                     limits = min_max_plots_growth,
                     expand = c(0.05,0,0.0,0)) +
  scale_shape_manual(breaks = c(TRUE,FALSE),
                     values = c(16,17),
                     labels = c("Truncating","Non-truncating"),
                     name = NULL) +
  scale_size(range = c(2,5),
             breaks = c(2,10,25,50),
             name = "N") +
  guides(size = guide_legend(nrow = 1,direction = "horizontal")) 

library(grid)

g <- ggplot_gtable(ggplot_build(plot_growth_per_year))
strip_both <- which(grepl('strip-', g$layout$name))
axis_x <- which(grepl('axis-b', g$layout$name))
colors <- unique(gene_colours[gene_values$gene_pure])
gene_labels <- gene_values$gene_pure
k <- 1
m <- 1
for (i in strip_both) {
  j <- which(grepl("text", g$grobs[[i]]$grobs[[1]]$childrenOrder))
  L <- g$grobs[[i]]$grobs[[1]]$children[[j]]$children[[1]]$label
  if (L %in% gene_values$gene_pure) {
    g$grobs[[i]]$grobs[[1]]$children[[j]]$children[[1]]$gp$col <- colors[k]
    g$grobs[[i]]$layout$clip <- "off"
    #nt <- grepl('nt',g$grobs[[i]]$grobs[[1]]$children[[2]]$children[[1]]$label)
    if (k %% 2 == 0) {
      g$grobs[[i]]$heights <- unit.c(unit(6,"pt"),unit(-0.6,"cm"),unit(6,"pt"))
    } else {
      g$grobs[[i]]$heights <- unit.c(unit(6,"pt"),unit(0.2,"cm"),unit(6,"pt"))
    }
    m <- m+1
    k <- k+1
  } else {
    #g$grobs[[i]]$heights <- unit.c(unit(6,"pt"),unit(-,"cm"),unit(6,"pt"))
  }
}
for (i in axis_x) {
  L <- g$grobs[[i]]$children$axis$grobs[[2]]$children[[1]]$label
  gp <- g$grobs[[i]]$children$axis$grobs[[2]]$children[[1]]$gp
  ff <- ifelse(grepl('runc',L),'bold','plain')
  s <- ifelse(grepl('runc',L),gp$fontsize+1,gp$fontsize-1)
  gp <- gpar(
    fontsize = s,
    col = gp$col,
    fontfamily = gp$fontfamily,
    lineheight = gp$lineheight,
    fontface=ff)
  g$grobs[[i]]$children$axis$grobs[[2]]$children[[1]]$gp <- gp
}

grid.newpage()
grid.draw(g) 

ggsave(useDingbats=FALSE,plot = gridExtra::grid.arrange(g),
       sprintf("figures/%s/dynamic_coefficients/FigureFull.pdf",model_id),
       width = 10,height = 6) 
ggsave(plot = gridExtra::grid.arrange(g),
       sprintf("figures/%s/dynamic_coefficients/FigureFull.svg",model_id),
       width = 10,height = 6)
```

#### Inspecting mutation-independent effects

The advantage of having mutation-independent effects is that we have a coefficient which corresponds to effects that affect clone growth at an individual clone level, allowing us to more accurately quantify the genetic growth without the influences of deviations in growth dynamics at an individual clone and quantify deviations from the expected genetic growth. In doing so, we are 1) observe clear deviations from genetic effects and inspect the trajectories corresponding to them and 2) see if age has any significant association in these coefficients, which would entail that age has an effect on the growth dynamics of different clones. 

In this section, we first assess which individuals have 1) the highest median mutation-independent growth effect, 2) the lowest median mutation-independent growth effect and 3) the widest range of mutation-independent growth effects. Next, we assess whether there is any association between the age at study entry and mutation-independent growth effects.

```{r mutation_independent_effects,fig.width=12,fig.height=5}
single_occurring_aa <- full_data$amino_acid_change[!full_data$single_occurring] %>%
  unique
mutation_independent_individual_data <- r_values %>%
  group_by(individual) %>% 
  filter(any((site %in% single_occurring_aa))) %>% 
  select(individual,age,coefficient,b_clone) %>%
  group_by(individual,coefficient,b_clone) %>%
  summarise(age = max(age)) %>%
  distinct() %>%
  gather(key = 'key',value = 'value',-individual,-age) %>%
  group_by(key,individual) %>%
  summarise(m=min(value),med=median(value),M=max(value),N = length(value)) %>% 
  mutate(Range = M - m) %>% 
  subset(N > 1) %>% 
  subset(key != 'coefficient')
mutation_independent_individual_dist <- mutation_independent_individual_data %>% 
  ggplot(aes(x = reorder(as.factor(individual),med),y = med,ymin=m,ymax=M,
             color = med > 0)) +
  geom_hline(yintercept = 0) + 
  geom_pointrange(alpha=0.5) + 
  theme_gerstung(base_size = 15) + 
  rotate_x_text() + 
  scale_color_aaas(guide = F) + 
  ylab("Unknown cause growth effects") + 
  xlab("Individual")
```

```{r, fig.width=24,fig.height=20,warning=F}
top_5_mutation_independent_individual_data <- mutation_independent_individual_data %>%
  arrange(med) %>%
  tail(5) %>%
  ungroup() %>%
  select(individual) %>%
  unlist %>% 
  rev()

plot_list_top <- list()
for (ind in top_5_mutation_independent_individual_data) {
  tmp <- r_values %>% 
    subset(individual == ind) %>%
    mutate(site_num = 1+(as.numeric(as.factor(as.character(site)))+length(unique(age))) %% length(unique(age))) %>%
    mutate(plot = ifelse(age == age[site_num],b_clone,NA))
  tmp_2 <- tmp %>%
    subset(!is.na(plot))
  pos <- c(0,max(tmp$pred_095/tmp$coverage)*0.95)
  plot_list_top[[as.character(ind)]] <- tmp %>% 
    ggplot(aes(x = age,
               group = substr(site,1,15),
               colour = substr(site,1,15),
               y = pred/coverage,
               ymax = pred_095/coverage,
               ymin = pred_005/coverage,
               shape = site %in% single_occurring_aa,
               label = round(plot,3))) + 
    geom_line() + 
    geom_pointrange(size = 0.7,alpha = 0.6) +
    geom_label_repel(data = tmp_2,
                     aes(y = pred/coverage,x = age,label = round(plot,3),colour = substr(site,1,20)),
                     inherit.aes = F,
                     label.r = unit(0,'cm'),label.size = unit(0,'cm'),
                     alpha=0.7) + 
    ggtitle(ind) + 
    scale_shape_discrete(name = 'Present in more than\none individual') + 
    scale_color_aaas(name = NULL) + 
    theme_gerstung() + 
    xlab("Age") + 
    ylab("VAF") + 
    theme(legend.position = "bottom") +
    scale_y_continuous(limits = c(0,pos[2]/0.95)) +
    guides(colour=guide_legend(nrow=3,byrow=TRUE),
           shape=guide_legend(nrow=2,byrow=TRUE))
}

top_iqr_5_mutation_independent_individual_data <- mutation_independent_individual_data %>%
  arrange(Range) %>%
  tail(5) %>%
  ungroup() %>%
  select(individual) %>%
  unlist %>% 
  rev()
plot_list_top_iqr <- list()
for (ind in top_iqr_5_mutation_independent_individual_data) {
  tmp <- r_values %>% 
    subset(individual == ind) %>%
    mutate(site_num = 1+(as.numeric(as.factor(as.character(site)))+length(unique(age))) %% length(unique(age))) %>%
    mutate(plot = ifelse(age == age[site_num],b_clone,NA))
  tmp_2 <- tmp %>%
    subset(!is.na(plot))
  pos <- c(0,max(tmp$pred_095/tmp$coverage)*0.95)
  plot_list_top_iqr[[as.character(ind)]] <- tmp %>% 
    ggplot(aes(x = age,
               group = substr(site,1,15),
               colour = substr(site,1,15),
               y = pred/coverage,
               ymax = pred_095/coverage,
               ymin = pred_005/coverage,
               shape = site %in% single_occurring_aa,
               label = round(plot,3))) + 
    geom_line() + 
    geom_pointrange(size = 0.7,alpha = 0.6) +
    geom_label_repel(data = tmp_2,
                     aes(y = pred/coverage,x = age,label = round(plot,3),colour = substr(site,1,20)),
                     inherit.aes = F,
                     label.r = unit(0,'cm'),label.size = unit(0,'cm'),
                     alpha=0.7) + 
    ggtitle(ind) + 
    scale_shape_discrete(name = 'Present in more than\none individual') + 
    scale_color_aaas(name = NULL) + 
    theme_gerstung() + 
    xlab("Age") + 
    ylab("VAF") + 
    theme(legend.position = "bottom") +
    scale_y_continuous(limits = c(0,pos[2]/0.95)) +
    guides(colour=guide_legend(nrow=3,byrow=TRUE),
           shape=guide_legend(nrow=2,byrow=TRUE))
}

bottom_5_mutation_independent_individual_data <- mutation_independent_individual_data %>%
  arrange(-med) %>%
  tail(5) %>%
  ungroup() %>%
  select(individual) %>%
  unlist %>% 
  rev()
plot_list_bottom <- list()
for (ind in bottom_5_mutation_independent_individual_data) {
  tmp <- r_values %>% 
    subset(individual == ind) %>%
    mutate(site_num = 1+(as.numeric(as.factor(as.character(site)))+length(unique(age))) %% length(unique(age))) %>%
    mutate(plot = ifelse(age == age[site_num],b_clone,NA))
  tmp_2 <- tmp %>%
    subset(!is.na(plot))
  pos <- c(0,max(tmp$pred_095/tmp$coverage)*0.95)
  plot_list_bottom[[as.character(ind)]] <- tmp %>% 
    ggplot(aes(x = age,
               group = substr(site,1,15),
               colour = substr(site,1,15),
               y = pred/coverage,
               ymax = pred_095/coverage,
               ymin = pred_005/coverage,
               shape = site %in% single_occurring_aa)) + 
    geom_line() + 
    geom_pointrange(size = 0.7,alpha = 0.6) +
    geom_label_repel(data = tmp_2,
                     aes(y = pred/coverage,x = age,label = round(plot,3),colour = substr(site,1,20)),
                     inherit.aes = F,
                     label.r = unit(0,'cm'),label.size = unit(0,'cm'),
                     alpha=0.7) + 
    ggtitle(ind) + 
    scale_shape_discrete(name = 'Present in more than\none individual') + 
    scale_color_aaas(name = NULL) + 
    theme_gerstung() + 
    xlab("Age") + 
    ylab("VAF") + 
    theme(legend.position = "bottom") +
    scale_y_continuous(limits = c(0,pos[2]/0.95)) +
    guides(colour=guide_legend(nrow=3,byrow=TRUE),
           shape=guide_legend(nrow=2,byrow=TRUE))
}

plot_grid(
  plot_grid(
    plot_grid(plotlist = plot_list_top,ncol = 1),
    plot_grid(plotlist = plot_list_bottom,ncol = 1),
    plot_grid(plotlist = plot_list_top_iqr,ncol = 1),
    nrow = 1
  ),
  mutation_independent_individual_dist + coord_flip(),
  rel_widths = c(1.0,0.2),
  nrow = 1
) + 
  ggsave(useDingbats=FALSE,filename = sprintf("figures/%s/inferred_trajectories/divergent_trajectories.pdf",model_id),width = 24,height = 20)
```

The above figure is split into 4 columns, corresponding to 1) the 5 individuals with highest median mutation-independent growth effect, 2) the 5 individuals with the lowest median mutation-independent growth effect and 3) the 5 individuals with the widest range of mutation-independent growth effects and 4) the general distribution of mutation-independent growth effects *per* individual. In 1), a large part of high mutation-independent growth effects are attributable to sites which do not have a site effect (because they are present in no more than a single individual). However, a few cases where we observe what can be synergistic combinations of mutations, as is the case in 5247 and 158. In 2) we have present a few examples of clones which approach a level quite close to their maximum size ($VAF = 0.5$), with their growth slowing down before reaching the maximum VAF. This may be happening due to spatial constraints in the bone marrow where the clones develop. In 3) we have one fairly interesting example in 260, where two SRSF2-P95 clones are in clear competition, with the fastest growing one (SRSF2-P95H) suppresing the other clone.

```{r mutation_independent_age_association, fig.width=5,fig.height=5}
min_max_plots <- c(min(c(r_values$coefficient_005,r_values$b_clone)),max(c(r_values$coefficient_095,r_values$b_clone)))
min_max_plots <- exp(min_max_plots) * 100 - 100
min_max_plots_round <- round(min_max_plots,-1)
tmp <- r_values %>%
  group_by(individual,site,b_clone,genes) %>%
  summarise(age = min(age)) %>% 
  ungroup %>% 
  mutate(genes = str_match(genes,'[A-Z0-9]+')) 
cor_clone_age <- cor.test(tmp$b_clone,tmp$age)
TEXT <- c(
  sprintf('R^2 == %s',round(cor_clone_age$estimate^2,4)),
  sprintf('p-value == %s',format(cor_clone_age$p.value,digits=3,nsmall=3,scientific=T)))
r_values %>%
  group_by(individual,site,b_clone,genes) %>%
  summarise(age = min(age)) %>% 
  ungroup %>% 
  mutate(genes = str_match(genes,'[A-Z0-9]+')) %>% 
  ggplot(aes(x = age,y = exp(b_clone)*100-100,colour = genes)) + 
  geom_smooth(se = F,method = 'lm',formula = y ~ x,color = 'black',linetype=2) + 
  geom_point(alpha = 0.4) + 
  theme_gerstung(base_size = 15) + 
  xlab("Age at study entry") + 
  ylab("Unknown cause growth per year") + 
  scale_color_manual(values = gene_colours,name = "Gene") + 
  guides(colour=guide_legend(ncol=2,byrow=F)) +
  theme(legend.position = 'none') +
  scale_y_continuous(breaks = seq(min_max_plots_round[1],min_max_plots_round[2],by = 20),
                     labels = paste0(seq(min_max_plots_round[1],min_max_plots_round[2],by = 20),'%'),
                     limits = min_max_plots_growth,
                     expand = c(0.05,0,0.0,0)) +
  annotate('text',x = min(r_values$age),y = 40,
           label = TEXT[1],
           fontface = "italic",
           parse = T,
           size = 5,
           hjust = 0) + 
  annotate('text',x = min(r_values$age),y = 35,
           label = TEXT[2],
           fontface = "italic",
           size = 5,
           parse = T,
           hjust = 0) + 
  ggsave(useDingbats=FALSE,sprintf("figures/%s/dynamic_coefficients/MutationIndependentAge.pdf",model_id),
         height = 5,width = 5) +
  ggsave(sprintf("figures/%s/dynamic_coefficients/MutationIndependentAge.svg",model_id),
         height = 5,width = 5)
```

To assess whether there was a major effect played by age on accelerating clone growth, we calculated the correlation between both. This value was quite low ($R = -8*10^{-4}$) and not statistically significant ($p-value = 0.462$). As such, it is unlikely that age plays a role in accelerating clonal dynamics in a senior population.

```{r mutation_independent_site_association,fig.height=5,fig.width=14}
r_values %>%
  subset(site %in% values_model$training_subset$unique_site_multiple) %>% 
  group_by(individual,site,b_clone,genes) %>%
  summarise(age = min(age)) %>% 
  ungroup %>% 
  mutate(genes = str_match(genes,'[A-Z0-9]+')) %>% 
  mutate(site = as.character(site)) %>% 
  mutate(site = sapply(site,function(x) unlist(strsplit(x,'-'))[2])) %>%
  mutate(site = sapply(site,function(x) unlist(strsplit(x,';'))[1])) %>%
  mutate(site = sapply(site,function(x) {
    tmp <- unlist(strsplit(x,':'))
    return(tmp[length(tmp)])})) %>%
  ggplot(aes(x = site,y = b_clone,colour = genes)) + 
  geom_jitter(alpha = 0.4,width = 0.3) +
  geom_boxplot(outlier.colour = NA,fill = NA) +
  theme_gerstung(base_size = 15) + 
  xlab("") + 
  ylab("Unknown growth effect") + 
  scale_color_manual(values = gene_colours,name = "Gene") + 
  guides(colour=guide_legend(ncol=2,byrow=F)) +
  theme(legend.position = 'none',strip.text = element_text(angle = 90,face = 'italic')) +
  rotate_x_text() +
  facet_grid(~ genes,space="free",scales = "free_x") + 
  ggsave(useDingbats=FALSE,sprintf("figures/%s/dynamic_coefficients/MutationIndependentSiteBoxplot.pdf",model_id),
         height = 5,width = 5)
```

```{r mutation_independent_per_gene,fig.width=4,fig.height=5}
mutation_independent_coefficients <- r_values %>%
  group_by(individual,site,b_clone,genes) %>%
  summarise(age = min(age)) %>% 
  ungroup %>% 
  mutate(genes = str_match(genes,'[A-Z0-9]+'))
mutation_independent_coefficients %>% 
  ggplot(aes(x = genes,y = exp(b_clone)*100-100,colour = genes)) + 
  geom_jitter(alpha = 0.5,width = 0.3,height = 0,size = 0.8) +
  geom_boxplot(outlier.colour = NA,fill = NA,
               colour = "black") +
  theme_gerstung(base_size = 15) + 
  xlab("Driver gene") + 
  ylab("Unknown cause growth per year") + 
  scale_color_manual(values = gene_colours,name = "Gene") + 
  guides(colour=guide_legend(ncol=2,byrow=F)) +
  theme(legend.position = 'none',axis.text.x = element_text(face = 'italic')) +
  rotate_x_text() + 
  scale_y_continuous(breaks = seq(min_max_plots_round[1],min_max_plots_round[2],by = 20),
                     labels = paste0(seq(min_max_plots_round[1],min_max_plots_round[2],by = 20),'%'),
                     limits = min_max_plots_growth,
                     expand = c(0.05,0,0.0,0)) +
  ggsave(useDingbats=FALSE,sprintf("figures/%s/dynamic_coefficients/MutationIndependentGeneBoxplot.pdf",model_id),
         height = 6,width = 4)

var.test(mutation_independent_coefficients$b_clone[mutation_independent_coefficients$genes == 'TET2'],
         mutation_independent_coefficients$b_clone[mutation_independent_coefficients$genes == 'JAK2'],
         alternative = "greater")
```

These boxplots allow us to observe whether the mutation-independent effects are more variable between individuals. This allows us to assess which mutations lead to clones showing less variability in clonal dynamics. From this analysis we can also clearly see that the variability in TET2 is, for example, much greater than that of DNMT3A ($ratio\ of\ variances=3.10$, $p-value=2.442*10^{-15}$ for a standard one-sided F-test). This has clinical implications, as understanding what will happen to TET2 clones becomes much more complicated. 

```{r compare_gene_unknown_cause,fig.height=4,fig.width=4}
good_stat_sites <- statistic_full %>%
  subset(sum_stat < 0.7) %>% 
  select(site) %>% 
  unlist
Parameters <- data.frame(
  b_clone_mean = values_model$b_clone[[1]][values_model$b_clone[[1]]$labels == 'mean',][values_model$interference_idxs$ind_site,]$values,
  b_clone_var = values_model$b_clone[[1]][values_model$b_clone[[1]]$labels == 'var',][values_model$interference_idxs$ind_site,]$values,
  b_genetic_mean = values_model$b_values[[1]][values_model$b_values[[1]]$labels == 'mean',][values_model$interference_idxs$site,]$values,
  b_genetic_var = values_model$b_values[[1]][values_model$b_values[[1]]$labels == 'var',][values_model$interference_idxs$site,]$values,
  u_low = values_model$u_values[[1]][values_model$u_values[[1]]$labels == '0.05',][values_model$interference_idxs$ind_site,]$values,
  u_high = values_model$u_values[[1]][values_model$u_values[[1]]$labels == '0.95',][values_model$interference_idxs$ind_site,]$values,
  b_clone_005 = values_model$b_clone[[1]][values_model$b_clone[[1]]$labels == '0.05',][values_model$interference_idxs$ind_site,]$values,
  b_clone_095 = values_model$b_clone[[1]][values_model$b_clone[[1]]$labels == '0.95',][values_model$interference_idxs$ind_site,]$values,
  b_genetic_005 = values_model$b_values[[1]][values_model$b_values[[1]]$labels == '0.05',][values_model$interference_idxs$site,]$values,
  b_genetic_095 = values_model$b_values[[1]][values_model$b_values[[1]]$labels == '0.95',][values_model$interference_idxs$site,]$values,
  u_005 = values_model$u_values[[1]][values_model$u_values[[1]]$labels == '0.05',][values_model$interference_idxs$ind_site,]$values,
  u_095 = values_model$u_values[[1]][values_model$u_values[[1]]$labels == '0.95',][values_model$interference_idxs$ind_site,]$values,
  individual = val_subset$unique_individual_true[interference_idxs$ind],  
  site = val_subset$unique_site[values_model$gene_idxs][interference_idxs$site]
) %>%
  mutate(recurring = site %in% values_model$training_subset$unique_site_multiple) %>% 
  distinct %>%
  subset(site %in% good_stat_sites) %>%
  mutate(gene = str_match(site,'[0-9A-Z]+')) 

plot_min_max_effects <- c(
  min(Parameters$b_clone_005,Parameters$b_genetic_005),
  max(Parameters$b_clone_095,Parameters$b_genetic_095)
)
plot_min_max_effects <- exp(plot_min_max_effects) * 100 - 100
ggplot(Parameters,aes(x = exp(b_clone_mean) * 100 - 100,
                      xmax = exp(b_clone_005) * 100 - 100,
                      ymax = exp(b_clone_095) * 100 - 100,
                      y = exp(b_genetic_mean) * 100 - 100,
                      ymin = exp(b_genetic_005) * 100 - 100,
                      ymax = exp(b_genetic_095) * 100 - 100,
                      colour = gene)) + 
  geom_abline(slope = 1,linetype = "longdash") +
  geom_point() +
  theme_gerstung(base_size = 15) + 
  scale_y_continuous(limits = plot_min_max_effects,
                     breaks = seq(-20,100,by = 20),
                     labels = paste0(seq(-20,100,by = 20),'%')) + 
  scale_x_continuous(limits = plot_min_max_effects,
                     breaks = seq(-20,100,by = 20),
                     labels = paste0(seq(-20,100,by = 20),'%')) +
  xlab("Unknown cause\ngrowth per year") +
  ylab("Driver growth per year") +
  scale_colour_manual(values = gene_colours,
                      guide = F)
```

### Inferring ages at clone foundation

The ages at clone foundation are determined using the inferred median values for each parameter for each individual. To do this, we solve $t_0 = \frac{log(\frac{1}{n}) - \alpha + log(\frac{g}{\beta}) - 1}{\beta}$, with $n = 200,000$, $g = 13$, $\beta=growth\ coefficient$ and $alpha = individual\ offset$. We exclude from analysis trajectories with $RE \ge 0.7$ and negative growth coefficients (this leads to no real solution). We also remove impossible values for ages at clone foundation from further analysis ($>100$ and $<0$). 

```{r model_f2_age_at_clone_growth, fig.width=14,fig.height=5.5}
clone_ages <- r_values %>%
  merge(statistic_full,by = c("site","individual"),all = F) %>%
  filter(sum_stat < 0.7) %>% 
  filter(coefficient + b_clone > 0) %>%
  mutate(CloneAgeHenryLower_adj = t0_adjusted(u_values,coefficient + b_clone,13,5e4) + values_model$min_age,
         CloneAgeHenryUpper_adj = t0_adjusted(u_values,coefficient + b_clone,13,2e5) + values_model$min_age
         )
  
data_plot_clone_ages <- clone_ages %>%
  mutate(Truncating = grepl("[A-Z0-9]+t",genes),
         Gene = str_match(genes,"[A-Z0-9]+")[,1],
         #Gene = genes,
         amino_acid_change = as.character(site)) %>%
  subset(abs(CloneAgeHenryLower_adj) < 120 & CloneAgeHenryUpper_adj > -10) %>%
  mutate(aa_org = amino_acid_change) %>% 
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) unlist(strsplit(x,'-'))[2])) %>%
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) unlist(strsplit(x,';'))[1])) %>%
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) {
    tmp <- unlist(strsplit(x,':'))
    return(tmp[length(tmp)])})) %>% 
  mutate(Gene = factor(as.character(Gene),levels = gene_order)) %>%
  mutate(Truncating_numeric = as.numeric(Truncating)) %>% 
  select(individual,CloneAgeHenryLower_adj,CloneAgeHenryUpper_adj,Gene,amino_acid_change,Truncating,Truncating_numeric,aa_org) %>%
  distinct() %>%
  mutate(dummy_individual = reorder(1:length(individual), CloneAgeHenryLower_adj)) %>%
  group_by(amino_acid_change) %>%
  mutate(Order = mean((CloneAgeHenryLower_adj + CloneAgeHenryUpper_adj) / 2 + Truncating_numeric * 1000))

data_plot_clone_ages_summary <- data_plot_clone_ages %>%
  subset(aa_org %in% values_model$training_subset$unique_site_multiple) %>%
  group_by(amino_acid_change,Gene) %>%
  summarise(max_age = max(CloneAgeHenryLower_adj),
            min_age = min(CloneAgeHenryUpper_adj),
            n = length(CloneAgeHenryUpper_adj),
            Order = Order[1])%>%
  ungroup() %>%
  mutate(width = n / max(n) * 1)

plot_clone_ages <- data_plot_clone_ages %>%
  ungroup() %>%
  mutate(amino_acid_change = factor(amino_acid_change,levels = mutation_aa_order)) %>% 
  na.omit() %>% 
  ggplot(aes(group = dummy_individual,
             x = amino_acid_change)) +
  geom_hline(yintercept = 0) +
  geom_tile(
    data = data_plot_clone_ages_summary,
    inherit.aes = F,
    alpha = 0.2,
    aes(
      height = max_age - min_age,
      x = factor(amino_acid_change,levels = mutation_aa_order),
      y = (max_age + min_age)/2,
      width = 0.85,
      fill = Gene
    )
  ) + 
  geom_linerange(
    size = 0.3,
    alpha = 0.8,
    aes(ymax = CloneAgeHenryLower_adj,
        ymin = CloneAgeHenryUpper_adj,
        color = Gene),
    position = position_dodge(width=1.0)) +
  geom_point(
    size = 1.2,
    alpha = 0.9,
    position = position_dodge(width=1.0),
    aes(shape = Truncating,
        y = (CloneAgeHenryLower_adj + CloneAgeHenryUpper_adj) / 2,
        color = Gene)) +
  facet_grid(~ Gene,scales = "free",space = "free") + 
  theme_gerstung(base_size = 15) + 
  theme(strip.text = element_text(size = 10,face = 'bold.italic'),
        legend.position = 'bottom',
        legend.key.height = unit(0,"cm")) + 
  ylab("Age at clone foundation") +
  xlab("Site") +
  rotate_x_text() + 
  scale_color_manual(values = gene_colours,
                     guide = F) +
  scale_fill_manual(values = gene_colours,
                    guide = F) +
  scale_y_continuous(
    expand = c(0,0),
    limits = c(-10,100),
    oob = scales::squish,
    breaks = c(-500,-400,-300,-200,-100,0,10,20,30,40,50,60,70,80,90,100,1000,2000,3000,5000)
    ) + 
  scale_shape_discrete(breaks = c(TRUE,FALSE),
                       labels = c("Truncating","Non-truncating"),
                       name = NULL) 

g <- ggplot_gtable(ggplot_build(plot_clone_ages))
strip_both <- which(grepl('strip-', g$layout$name))
colors <- gene_colours[gene_order[gene_order %in% data_plot_clone_ages_summary$Gene]]
k <- 1
for (i in strip_both) {
  j <- which(grepl("text", g$grobs[[i]]$grobs[[1]]$childrenOrder))
  g$grobs[[i]]$grobs[[1]]$children[[j]]$children[[1]]$gp$col <- colors[k]
  g$grobs[[i]]$layout$clip <- "off"
  if (k %% 2 == 0) {
    g$grobs[[i]]$heights <- unit.c(unit(6,"pt"),unit(-0.5,"cm"),unit(6,"pt"))
  } else {
    g$grobs[[i]]$heights <- unit.c(unit(6,"pt"),unit(0.5,"cm"),unit(6,"pt"))
  }
  k <- k+1
}

grid.newpage()
grid.draw(g) 

ggsave(useDingbats=FALSE,gridExtra::grid.arrange(g),
       filename = sprintf("figures/%s/age_at_clone_foundation/Figure6.pdf",model_id),
       width = 14,height = 5.5)
ggsave(gridExtra::grid.arrange(g),
       filename = sprintf("figures/%s/age_at_clone_foundation/Figure6.svg",model_id),
       width = 14,height = 5.5)

total_clones <- r_values %>%
  merge(statistic_full,by = c("site","individual"),all = F) %>%
  select(individual,site,sum_stat,u_values,coefficient,b_clone) %>%
  distinct() %>%
  mutate(CloneAgeHenryLower_adj = t0_adjusted(u_values,coefficient + b_clone,13,5e4) + values_model$min_age,
         CloneAgeHenryUpper_adj = t0_adjusted(u_values,coefficient + b_clone,13,2e5) + values_model$min_age
         )

clones_considered <- list(
  total = nrow(total_clones),
  total_nn = nrow(total_clones %>% 
                    subset(b_clone + coefficient > 0)),
  post_qc = nrow(total_clones %>% 
                   subset(b_clone + coefficient > 0) %>%
                   subset(sum_stat < 0.7)),
  post_qc_age = nrow(total_clones %>% 
                       subset(b_clone + coefficient > 0) %>%
                       subset(sum_stat < 0.7) %>%
                       subset(abs(CloneAgeHenryLower_adj) < 120 & CloneAgeHenryUpper_adj > -10))
)
clones_considered$post_qc_age / clones_considered$total
```

```{r ages_density,fig.height=5,fig.width=4}
gene_order_age <- data_plot_clone_ages %>% 
  mutate(Gene = str_match(Gene,'[0-9A-Z]+')) %>% 
  mutate(AgeMidPoint = (CloneAgeHenryLower_adj + CloneAgeHenryUpper_adj)/2) %>%
  group_by(Gene) %>%
  summarise(m = median(AgeMidPoint)) %>%
  ungroup() %>%
  arrange(m) %>% 
  select(Gene) %>%
  unlist

data_plot_clone_ages %>%
  mutate(Gene = str_match(Gene,'[0-9A-Z]+')) %>% 
  mutate(Gene = factor(Gene,levels = gene_order_age)) %>% 
  mutate(AgeMidPoint = (CloneAgeHenryLower_adj + CloneAgeHenryUpper_adj)/2) %>%
  group_by(Gene) %>%
  mutate(MoreThanOne = length(Gene) >= 2) %>%
  ungroup() %>%
  subset(MoreThanOne) %>%
  ggplot(aes(y = AgeMidPoint,fill = Gene,
             x = Gene,group = Gene,
             color = Gene)) +
  geom_jitter(alpha = 0.5,width = 0.2,size=0.8) + 
  geom_boxplot(alpha = 1.0,fill=NA,outlier.colour = NA,
               colour = "black") +
  theme_gerstung(base_size = 15) + 
  theme(axis.text.y = element_text(face = 'italic'),
        legend.position = 'bottom',
        legend.key.height = unit(0,"cm")) + 
  #rotate_x_text() + 
  scale_fill_manual(values = gene_colours,
                    guide = F) +
  scale_colour_manual(values = gene_colours,
                      guide = F) +
  scale_shape_discrete(breaks = c(TRUE,FALSE),
                       labels = c("Truncating","Non-truncating"),
                       name = NULL) +
  scale_y_continuous(breaks = seq(0,100,10),expand = c(0.01,0.01),limits = c(0,90)) +
  ylab("Age at clone foundation") + 
  xlab("Driver gene") +
  coord_flip() + 
  ggsave(useDingbats=FALSE,sprintf("figures/%s/age_at_clone_foundation/Figure6_Boxplots.pdf",model_id),width = 4,height = 5) +
  ggsave(sprintf("figures/%s/age_at_clone_foundation/Figure6_Boxplots.svg",model_id),width = 4,height = 5) 
```

```{r age_at_clone_foundation_intervals}
sample_from_chains <- function(draws,
                               clone_idx,
                               gene_idx,
                               site_idx,
                               u_idx) {
  
  c_names <- colnames(values_model$draws$`11`)
}

u_c_names <- grep('^u',c_names,perl = T)
b_gene_c_names <- grep('^b_gene',c_names,perl = T)
b_site_c_names <- grep('^b_site',c_names,perl = T)
b_clone_c_names <- grep('^b_clone',c_names,perl = T)

sampling_idxs <- values_model$interference_idxs %>%
  mutate(site_label = values_model$training_subset$unique_site[gene_idxs][site]) %>%
  mutate(site = as.numeric(factor(site_label,values_model$training_subset$unique_site_multiple))) %>% 
  mutate(gene_label = str_match(site_label,'[A-Z0-9a-z]+')) %>%
  mutate(gene = as.numeric(factor(gene_label,values_model$training_subset$unique_gene))) %>%
  mutate(individual_label = values_model$training_subset$unique_individual_true[ind]) %>% 
  mutate(age = values_model$training_subset$ages[ind_age]) %>% 
  group_by(site,ind,ind_site,gene,site_label,gene_label,individual_label) %>%
  summarise(min_age = min(age)) %>% 
  ungroup() %>%
  distinct

age_estimates <- sampling_idxs %>%
  apply(1,function(x) {
    site_idx <- as.numeric(x[1])
    individual_idx <- as.numeric(x[2])
    individual_site_idx <- as.numeric(x[3])
    gene_idx <- as.numeric(x[4])
    
    ind <- values_model$training_subset$unique_individual_truex[2]
    t1 <- min(full_data$Age[full_data$SardID == values_model$training_subset$unique_individual_true[individual_idx]])

    b_gene <- sample(x = values_model$draws$`11`[,b_gene_c_names[gene_idx]],size = 1000,replace = F)
    if (!is.na(site_idx)) {
      b_site <- sample(x = values_model$draws$`11`[,b_site_c_names[site_idx]],size = 1000,replace = F)
    } else {
      b_site <- 0
    }
    b_clone <- sample(x = values_model$draws$`11`[,b_clone_c_names[individual_site_idx]],size = 1000,replace = F)
    u <- sample(x = values_model$draws$`11`[,u_c_names[individual_site_idx]],size = 1000,replace = F)
    
    age_distribution <- t0_adjusted(u,b_site+b_gene+b_clone,13,2e5) + values_model$min_age
    age_distribution <- ifelse(age_distribution < -1,-1,age_distribution)
    age_distribution <- ifelse(age_distribution > t1, t1, age_distribution)
    
    age_distribution <- age_distribution[!is.na(age_distribution)]
    return(c(mean(age_distribution),quantile(age_distribution,c(0.05,0.10,0.16,0.5,0.84,0.90,0.95))))
  }) %>% 
  t %>% 
  as_tibble()

colnames(age_estimates) <- c("Mean","Q05","Q10","Q16","Q50","Q84","Q90","Q95")
Parameters_Age <- cbind(
  sampling_idxs,age_estimates
) %>% 
  select(-gene) %>% 
  merge(Parameters,by.x = c("individual_label","site_label"),by.y = c("individual","site")) %>%
  mutate(
    b_clone_016 = b_clone_mean - sqrt(b_clone_var),
    b_clone_084 = b_clone_mean + sqrt(b_clone_var),
    b_genetic_016 = b_genetic_mean - sqrt(b_genetic_var),
    b_genetic_084 = b_genetic_mean + sqrt(b_genetic_var)
  ) 

M <- Parameters_Age %>%
  subset(b_genetic_mean + b_clone_mean > 0) %>% 
  subset(b_genetic_mean + b_clone_mean == max(b_genetic_mean + b_clone_mean) | 
           (b_genetic_mean + b_clone_mean) == min(b_genetic_mean + b_clone_mean,1)) %>%
  arrange(- b_genetic_mean - b_clone_mean)

sample_to_display <- sampling_idxs %>%
  subset(site %in% M$site & ind %in% M$ind & ind_site %in% M$ind_site) %>% 
  apply(1,function(x) {
    site_idx <- as.numeric(x[1])
    individual_idx <- as.numeric(x[2])
    individual_site_idx <- as.numeric(x[3])
    gene_idx <- as.numeric(x[4])
    
    ind <- values_model$training_subset$unique_individual_truex[2]
    t1 <- min(full_data$Age[full_data$SardID == values_model$training_subset$unique_individual_true[individual_idx]])

    b_gene <- sample(x = values_model$draws$`11`[,b_gene_c_names[gene_idx]],size = 1000,replace = F)
    if (!is.na(site_idx)) {
      b_site <- sample(x = values_model$draws$`11`[,b_site_c_names[site_idx]],size = 1000,replace = F)
    } else {
      b_site <- 0
    }
    b_clone <- sample(x = values_model$draws$`11`[,b_clone_c_names[individual_site_idx]],size = 1000,replace = F)
    u <- sample(x = values_model$draws$`11`[,u_c_names[individual_site_idx]],size = 1000,replace = F)
    
    age_distribution <- t0_adjusted(u,b_site+b_gene+b_clone,13,2e5) + values_model$min_age
    age_distribution <- ifelse(age_distribution < -1,-1,age_distribution)
    age_distribution <- ifelse(age_distribution > t1, t1, age_distribution)
    
    return(data.frame(
      age_distribution = age_distribution,
      effect_distribution = b_clone + b_site + b_gene
    ))
  }) 
sample_to_display_large <- sample_to_display[[1]]
sample_to_display_other <- sample_to_display[[2]] %>%
  na.omit()

kd <- ks::kde(sample_to_display_large, compute.cont=TRUE)
contour_95 <- with(kd, contourLines(x=eval.points[[1]], y=eval.points[[2]],
                                    z=estimate, levels=cont["5%"])[[1]])
contour_95_large <- data.frame(contour_95)


kd <- ks::kde(sample_to_display_other, compute.cont=TRUE)
contour_95 <- with(kd, contourLines(x=eval.points[[1]], y=eval.points[[2]],
                                    z=estimate, levels=cont["5%"])[[1]])
contour_95_other <- data.frame(contour_95) %>%
  mutate(x = ifelse(x < -1, -1 ,x))

Parameters_Age %>%
  subset(b_clone_mean + b_genetic_mean > 0) %>%
  ggplot(aes(x = Q50,xmin = Q10,xmax = Q90,
             y = exp(b_genetic_mean + b_clone_mean)*100-100,
             ymin = exp(b_genetic_005 + b_clone_005)*100-100,
             ymax = exp(b_genetic_095 + b_clone_095)*100-100,
             colour = gene)) +
  geom_linerange(size = 0.2,alpha = 0.3) + 
  geom_errorbarh(size = 0.2,alpha = 0.3) +
  geom_point() + 
  # geom_polygon(
  #   data = contour_95_other,
  #   colour = 'black',
  #   inherit.aes = F,
  #   fill = 'black',
  #   alpha = 0.1,
  #   aes(x = x,
  #       y = exp(y)*100-100)
  # ) + 
  # geom_polygon(
  #   data = contour_95_large,
  #   colour = 'black',
  #   inherit.aes = F,
  #   fill = 'black',
  #   alpha = 0.1,
  #   aes(x = x,
  #       y = exp(y)*100-100)
  # ) + 
  #scale_y_continuous(limits = c(-100,200)) + 
  scale_y_continuous(
    breaks = c(-20,0,20,40,60,80,100,120),
    labels = paste0(c(-20,0,20,40,60,80,100,120),'%')
  ) +
  #coord_cartesian(xlim = c(0,100)) +
  ylab("Growth per year") +
  xlab("Age at clone foundation") +
  theme_gerstung(base_size = 15) + 
  scale_color_manual(values = gene_colours,
                     guide = F)
```

While at an individual site level this analysis can be quite complicated, a few patterns become evident when looking at these values at the gene level - for instance, mutations such as those in JAK2, IDH1/2, PTPN11 and U2AF1 appear almost exclusively much later in life when compared with DNMT3A and TET2, which appear along an individual's life.

```{r growth_clonalAge_association,fig.height=5,fig.width=5}
data_growth_age_association <- clone_ages %>%
  mutate(Truncating = grepl("[A-Z0-9]+t",genes),
         Gene = str_match(genes,"[A-Z0-9]+"),
         amino_acid_change = as.character(site)) %>%
  subset(abs(CloneAgeHenryLower_adj) < 120 & CloneAgeHenryUpper_adj > -10) %>%
  subset(amino_acid_change %in% values_model$training_subset$unique_site_multiple) %>%
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) unlist(strsplit(x,'-'))[2])) %>%
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) unlist(strsplit(x,';'))[1])) %>%
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) {
    tmp <- unlist(strsplit(x,':'))
    return(tmp[length(tmp)])})) %>% 
  mutate(Gene = factor(as.character(Gene),levels = unique(str_match(gene_order,'[A-Z0-9]+')))) %>%
  mutate(Truncating_numeric = as.numeric(Truncating)) %>% 
  select(site,individual,
         coefficient,coefficient_005,coefficient_095,
         b_clone,b_clone_005,b_clone_095,
         CloneAgeHenryLower_adj,CloneAgeHenryUpper_adj,Gene) %>% 
  distinct() %>%
  mutate(AgeMidpoint = (CloneAgeHenryLower_adj + CloneAgeHenryUpper_adj)/2)

C <- cor.test(
  data_growth_age_association$coefficient,
  data_growth_age_association$AgeMidpoint)
TEXT <- c(
  sprintf('R^2 == %s',round(C$estimate^2,2)),
  sprintf('p-value == %s',format(C$p.value,digits=3,nsmall=3,scientific=T)))
square_growth_age_test <- data_growth_age_association %>%
  mutate(S = ifelse(coefficient < 0.2,"Slow","Fast"))
wilcox.test(square_growth_age_test$AgeMidpoint[square_growth_age_test$S == 'Slow'],
            square_growth_age_test$AgeMidpoint[square_growth_age_test$S == 'Fast'],
            alternative = "less")
square_growth_age <- data_growth_age_association %>%
  mutate(S = ifelse(coefficient < 0.2,"Slow","Fast")) %>% 
  group_by(S) %>%
  summarise(min_age = min(AgeMidpoint),
            max_age = max(AgeMidpoint),
            min_coe = min(coefficient),
            max_coe = max(coefficient),
            genes = paste(unique(Gene),collapse = ", "))

square_growth_age$genes <- c("list(italic('IDH1'),italic('U2AF1'),)",square_growth_age$genes[2])

data_growth_age_association %>% 
  ggplot(aes(y = exp(coefficient)*100-100,x = AgeMidpoint,
             ymax = exp(coefficient_095)*100-100,ymin = exp(coefficient_005)*100-100,
             color = Gene)) + 
  geom_rect(inherit.aes = F,
            data = square_growth_age,
            mapping = aes(xmin = min_age - 2,xmax = max_age + 2,
                          ymin = exp(min_coe - 0.02)*100-100,ymax = exp(max_coe + 0.02)*100-100),
            color = c("#747474","#cccccc"),
            fill = NA,
            size = 1,
            linetype = 2,
            alpha = 0.5) +
  annotate(geom = "text",
           label = square_growth_age$genes[[1]],
           y = (exp(square_growth_age$max_coe[1])*100-100+exp(square_growth_age$min_coe[1])*100-100)/2,
           x = square_growth_age$min_age[1] - 3,
           color = "#747474",
           fontface = 'bold',
           parse=T,
           size = 4,
           hjust = 1,
           vjust = 1) +
  annotate(geom = "text",
           label = "italic('SRSF2')~(P95H)",
           y = (exp(square_growth_age$max_coe[1])*100-100+exp(square_growth_age$min_coe[1])*100-100)/2-3,
           x = square_growth_age$min_age[1] - 3,
           color = "#747474",
           fontface = 'bold',
           parse=T,
           size = 4,
           hjust = 1,
           vjust = 1) +
  scale_color_manual(values = gene_colours) +
  geom_point(alpha = 0.8) + 
  geom_errorbar(alpha = 0.4) + 
  theme_gerstung(base_size = 15) + 
  theme(legend.position = "none") +
  scale_y_continuous(breaks = seq(min_max_plots_round[1],min_max_plots_round[2],by = 20),
                     labels = paste0(seq(min_max_plots_round[1],min_max_plots_round[2],by = 20),'%'),
                     limits = min_max_plots_growth,
                     expand = c(0.05,0,0.0,0)) +
  xlab("Age at clone foundation") + 
  ylab("Driver-associated growth per year") + 
  ggsave(useDingbats=FALSE,sprintf("figures/%s/age_at_clone_foundation/GeneticGrowthCoefficient_vs_AgeAtCloneFoundation.pdf",model_id),
         height = 5,width = 5) +
  ggsave(sprintf("figures/%s/age_at_clone_foundation/GeneticGrowthCoefficient_vs_AgeAtCloneFoundation.svg",model_id),
         height = 5,width = 5)
```

Particularly we see here an enrichment for faster growing clones at later stages in life. This, however, might be due to our detection power as well - mutations that grow slower may very well appear much later in life, but they would require some time before they become identifiable with current our methodology.

### Assessing the predictive power of this model

Upon colony sequencing, 13 individuals were sequenced at an additional timepoint. We use this data and compare it with what we have so far to understand how well our model performs when predicting for individuals which have been studied for some time. Prediction in new individuals is assessed in the `vaf_modelling_paper_clinical_application_analysis.Rmd` notebook.

```{r predictive_power_6th_tp,fig.width=14,fig.height=6.5,warning=F,message=F}
beta_value <- values_model$beta_values[[1]][1,1]
r_values_for_prediction <- r_values %>%
  select(site,individual,
         coefficient,coefficient_005,coefficient_095,
         b_clone,b_clone_005,b_clone_095,
         u_values,u_values_005,u_values_095) %>%
  distinct %>%
  mutate(site = paste(str_match(site,'[A-Z0-9]+'),str_match(site,'-.*'),sep = ''))
predict_timepoint <- function(Individual,Site,Age,coverage=1000) {
  Site <- as.character(Site)
  Individual <- as.numeric(Individual)
  Age <- as.numeric(Age) - values_model$min_age
  tmp <- r_values_for_prediction %>% 
    subset((individual == Individual) & (site == Site)) %>% 
    mutate(
      p_005 = inv.logit((coefficient_005+b_clone_005) * Age + u_values_005) * 0.5,
      p = inv.logit((coefficient+b_clone) * Age + u_values) * 0.5,
      p_095 = inv.logit((coefficient_095+b_clone_095) * Age + u_values_095) * 0.5) %>% 
    transmute(
      coefficient = coefficient,
      b_clone = b_clone,
      u_values = u_values,
      p = p
    ) 
  S <- rbbinom(1000,coverage,(tmp$p*beta_value)/(1-tmp$p),beta_value)/coverage
  S <- ifelse(S > 0.5, 0.5,S)
  tmp$pred_005 <- quantile(S,0.05,names=F,na.rm = T)
  tmp$pred = mean(S)
  tmp$pred_095 = quantile(S,0.95,names=F,na.rm = T)
  return(tmp)
}
data_6th_tp <- load_data_6th_tp()

predictions_6th_tp <- data_6th_tp %>%
  apply(1,FUN=function(x) {c(x[1],x[12],x[4],predict_timepoint(x[1],x[12],x[4],2000),VAF=as.numeric(x[10]))}) %>%
  lapply(unlist) %>% 
  do.call(what=rbind) %>%
  as.data.frame() %>%
  mutate(coefficient = as.numeric(as.character(coefficient)),
         b_clone = as.numeric(as.character(b_clone)),
         u_values = as.numeric(as.character(u_values)),
         p = as.numeric(as.character(p)),
         pred_005 = as.numeric(as.character(pred_005)),
         pred = as.numeric(as.character(pred)),
         pred_095 = as.numeric(as.character(pred_095)),
         VAF = as.numeric(as.character(VAF)))

last_tp <- r_values %>%
  mutate(site = paste(str_match(site,'[A-Z0-9]+'),str_match(site,'-.*'),sep = '')) %>% 
  subset(individual %in% data_6th_tp$SardID & site %in% data_6th_tp$amino_acid_change) %>% 
  transmute(SardID=individual,amino_acid_change=site,Age=age,coefficient,b_clone,u_values,
            p=mu_val,pred_005=pred_005/coverage,pred=pred/coverage,pred_095=pred_095/coverage,VAF=true/coverage) %>%
  group_by(SardID,amino_acid_change) %>%
  filter(Age == max(Age)) %>% 
  ungroup() %>%
  rbind(predictions_6th_tp) %>%
  mutate(gene = sapply(amino_acid_change,function(x) unlist(strsplit(x,'-'))[1])) %>%
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) unlist(strsplit(x,'-'))[2])) %>%
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) unlist(strsplit(x,';'))[1])) %>%
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) {
    tmp <- unlist(strsplit(x,':'))
    return(tmp[length(tmp)])})) %>% 
  mutate(amino_acid_change = paste(gene,amino_acid_change,sep = '-')) %>% 
  na.omit() %>% 
  mutate(
    SardID = gsub(" ","",as.character(SardID)),
    amino_acid_change = as.character(amino_acid_change)) %>% 
  group_by(SardID,amino_acid_change) %>%
  mutate(last_timepoint = Age == max(Age)) %>%
  ungroup() %>%
  mutate(Age = as.numeric(Age)) %>% 
  filter(amino_acid_change %in% c("DNMT3A-R882H","GNB1-K57E",
                                  "JAK2-V617F","SF3B1-K666N",
                                  "SF3B1-K700E","SRSF2-P95H",
                                  "SRSF2-P95L","TET2-Q1542X",
                                  "U2AF1-Q157P","U2AF1-Q157R"))
last_tp_only <- last_tp %>%
  group_by(SardID, amino_acid_change) %>% 
  filter(Age == max(Age))

prediction_data <- r_values %>%
  mutate(site = paste(str_match(site,'[A-Z0-9]+'),str_match(site,'-.*'),sep = '')) %>% 
  subset(individual %in% data_6th_tp$SardID & site %in% data_6th_tp$amino_acid_change) %>% 
  transmute(SardID=individual,amino_acid_change=site,Age=age,coefficient,b_clone,u_values,
            p=mu_val,pred_005=pred_005/coverage,pred=pred/coverage,pred_095=pred_095/coverage,VAF=true/coverage) %>%
  mutate(gene = sapply(amino_acid_change,function(x) unlist(strsplit(x,'-'))[1])) %>%
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) unlist(strsplit(x,'-'))[2])) %>%
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) unlist(strsplit(x,';'))[1])) %>%
  mutate(amino_acid_change = sapply(amino_acid_change,function(x) {
    tmp <- unlist(strsplit(x,':'))
    return(tmp[length(tmp)])})) %>% 
  mutate(amino_acid_change = paste(gene,amino_acid_change,sep = '-')) %>% 
  na.omit() %>% 
  mutate(
    SardID = gsub(" ","",as.character(SardID)),
    amino_acid_change = as.character(amino_acid_change)) %>% 
  filter(amino_acid_change %in% c("DNMT3A-R882H","GNB1-K57E",
                                  "JAK2-V617F","SF3B1-K666N",
                                  "SF3B1-K700E","SRSF2-P95H",
                                  "SRSF2-P95L","TET2-Q1542X",
                                  "U2AF1-Q157P","U2AF1-Q157R"))

prediction_data %>%
  #filter(any(last_timepoint == T)) %>% 
  ggplot(aes(x = Age,
             y = pred,
             ymin = pred_005, 
             ymax = pred_095,
             group = SardID,
             fill = SardID,
             color = SardID)) + 
  geom_line(aes(y = VAF),linetype = 'dotted',color = 'black',alpha = 0.2) + 
  geom_ribbon(alpha = 0.2,color = NA) + 
  geom_point(aes(y = VAF),alpha = 0.2,color = 'black') + 
  geom_ribbon(data = last_tp,alpha = 0.2,color = NA) + 
  geom_point(data = last_tp_only,aes(y = VAF),shape = 1) + 
  geom_point(data = last_tp_only,aes(y = VAF),color = 'black',shape = 3) + 
  geom_line(data = last_tp,aes(y = VAF),linetype = 'dotted',color = 'black',alpha = 0.2) + 
  geom_line(alpha = 0.6) + 
  geom_line(data = last_tp,linetype = 'longdash',alpha = 0.6) +
  facet_wrap(~ paste(amino_acid_change,sep='-'),nrow = 2,scales = 'free') +
  #scale_y_continuous(trans = 'log10') + 
  theme_gerstung(base_size = 15) + 
  scale_color_discrete(guide=F) + 
  scale_fill_discrete(guide=F) + 
  scale_shape_discrete(guide=F) + 
  scale_y_continuous(trans = 'log10',limits = c(5e-4,0.5)) +
  scale_x_continuous(breaks = seq(0,100,by = 10),limits = c(58,92),expand=c(0,0)) + 
  ylab("VAF") + 
  theme(strip.text = element_text(size = 15),
        plot.title = element_text(face = "italic")) +
  ggtitle("Predictions for the last timepoint conditioned on the previous timepoints") +
  ggsave(useDingbats=FALSE,sprintf("figures/%s/inferred_trajectories/predicted_trajectories_6th_tp.pdf",model_id),
         width = 14,height = 6.5)
```

```{r compare_truth_and_predictions,fig.height=4,fig.width=7}
plot_min_max <- c(
  min(last_tp_only$VAF,last_tp_only$pred),
  max(last_tp_only$VAF,last_tp_only$pred)
)
performance_scatter_plot <- last_tp_only %>%
  #filter(any(last_timepoint == T)) %>% 
  group_by(gene) %>%
  mutate(
    LABEL = ifelse(0.03-VAF == min(0.03-VAF),
                   gene,
                   NA)
  ) %>% 
  ggplot(aes(x = VAF,
             y = pred,
             color = gene)) + 
  geom_abline(slope = 1,linetype = 3) + 
  geom_point(size=3) + 
  # geom_text_repel(aes(label = LABEL,
  #                     fill = gene),
  #                 direction = "y",
  #                 force = 10,
  #                 fontface = "italic",
  #                 color = "black",
  #                 alpha = 1.0,
  #                 label.r = unit(0,"cm"),
  #                 label.size = unit(0,"cm")) + 
  xlab("Observed VAF") + 
  ylab("Predicted VAF")  + 
  annotate(
    geom = "text",
    x = 0.015,
    y = 0.25,
    label = sprintf("MAE = %.3f",mean(abs(last_tp_only$VAF-last_tp_only$pred))),
    facefont = 'bold'
  ) + 
  theme_gerstung(base_size = 15) +
  scale_x_continuous(trans = 'log10',limits = plot_min_max) + 
  scale_y_continuous(trans = 'log10',limits = plot_min_max) +
  scale_color_manual(values=gene_colours,name=NULL) + 
  scale_fill_manual(values=gene_colours,name=NULL) + 
  theme(legend.text = element_text(face = "italic")) 

interesting_cases_6th_tp_plot <- prediction_data %>%
  #filter(any(last_timepoint == T)) %>% 
  subset(amino_acid_change %in% c("SF3B1-K666N","SRSF2-P95H")) %>% 
  ggplot(aes(x = Age,
             y = pred,
             ymin = pred_005, 
             ymax = pred_095,
             group = SardID,
             fill = SardID,
             color = SardID)) + 
  geom_line(aes(y = VAF),linetype = 'dotted',color = 'black',alpha = 0.2) + 
  geom_ribbon(alpha = 0.2,color = NA) + 
  geom_point(aes(y = VAF),alpha = 0.2,color = 'black') + 
  geom_ribbon(data = subset(last_tp,amino_acid_change %in% c("SF3B1-K666N","SRSF2-P95H")),alpha = 0.2,color = NA) + 
  geom_point(data = subset(last_tp_only,amino_acid_change %in% c("SF3B1-K666N","SRSF2-P95H")),
             aes(y = VAF),shape = 1) + 
  geom_point(data = subset(last_tp_only,amino_acid_change %in% c("SF3B1-K666N","SRSF2-P95H")),
             aes(y = VAF),color = 'black',shape = 3) + 
  geom_line(data = subset(last_tp,amino_acid_change %in% c("SF3B1-K666N","SRSF2-P95H")),
            aes(y = VAF),linetype = 'dotted',color = 'black',alpha = 0.2) + 
  geom_line(alpha = 0.6) + 
  geom_line(data = subset(last_tp,amino_acid_change %in% c("SF3B1-K666N","SRSF2-P95H")),
            linetype = 'longdash',alpha = 0.6) +
  facet_wrap(~ paste(amino_acid_change,sep='-'),nrow = 2,scales = 'free_x') +
  #scale_y_continuous(trans = 'log10') + 
  theme_gerstung(base_size = 15) + 
  scale_color_discrete(guide=F) + 
  scale_fill_discrete(guide=F) + 
  scale_shape_discrete(guide=F) + 
  scale_y_continuous(trans = 'log10') +
  scale_x_continuous(breaks = seq(0,100,by = 10),limits = c(58,92),expand=c(0,0)) + 
  ylab("VAF") + 
  theme(strip.text = element_text(size = 15))
  ggsave(useDingbats=FALSE,sprintf("figures/%s/inferred_trajectories/performance_and_predicted_6thpoint_trajectories.pdf",model_id),
         width = 14,height = 6.5)

plot_grid(
  plot_grid(performance_scatter_plot + theme(legend.position = "none",
                                             panel.grid = element_blank()),
            interesting_cases_6th_tp_plot,nrow=1,rel_widths = c(4.5/7,2.5/7)),
  plot_grid(get_legend(performance_scatter_plot + theme(legend.position = "bottom")),
            ggplot() + theme_void(),
            rel_widths = c(4.5/7,2.5/7)),
  rel_heights = c(0.9,0.1),
  ncol = 1
)
```

```{r predictive_power_boxplot,fig.height=4,fig.width=8}
plot_grid(
  data.frame(
    values = c(abs(prediction_data$VAF-prediction_data$pred),abs(last_tp_only$VAF-last_tp_only$pred)),
    dataset = c(rep('Train',length(prediction_data$VAF)),rep('Test',length(last_tp_only$VAF)))
  ) %>% 
    ggplot(aes(x = dataset,y = values,group = dataset)) +
    geom_boxplot() +
    theme_gerstung(base_size = 15) + 
    xlab("") + 
    ylab("Absolute error"),
  data.frame(
    values = c(abs(prediction_data$VAF-prediction_data$pred),abs(last_tp_only$VAF-last_tp_only$pred)),
    truth = c(prediction_data$VAF,last_tp_only$VAF),
    dataset = c(rep('Train',length(prediction_data$VAF)),rep('Test',length(last_tp_only$VAF)))
  ) %>% 
    ggplot(aes(x = truth,y = values,colour = dataset)) +
    geom_point() +
    theme_gerstung(base_size = 15) + 
    theme(legend.position = "bottom") + 
    xlab("log(VAF)") + 
    ylab("Absolute error"),
  nrow=1
)

cat(sprintf('\nMAE(train) = %.4f\n',mean(prediction_data$VAF-prediction_data$pred)))
cat(sprintf('MAE(test) = %.4f\n\n',mean(abs(last_tp_only$VAF-last_tp_only$pred))))
```

Here we can observe that $MAE(test)=3.37\%$, which is 6.74 times larger than our clone detection limit ($0.5\%$). As such, it can be risky to make predictions at low detection values.

## Global picture

Using what we have so far we can replicate the general picture derived by [McKerrel et al.](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4542313/).

```{r global_picture,fig.width=9,fig.height=6}
average_age <- data_plot_clone_ages %>%
  mutate(MidPoint = (CloneAgeHenryLower_adj + CloneAgeHenryUpper_adj)/2) %>%
  group_by(amino_acid_change) %>%
  summarise(MedianAge = mean(MidPoint),
            Gene = Gene[1]) %>% 
  transmute(site = paste(str_match(Gene,"[A-Z0-9]+"),
                         amino_acid_change,sep = '-'),
            Gene = Gene,
            MedianAge = MedianAge)

most_recurring_coefficients_gene <- r_values_list$model_f2 %>% 
  mutate(Gene = str_match(genes,'[A-Z0-9]+')) %>%
  group_by(Gene) %>% 
  filter(site == names(sort(table(site),decreasing = T)[1]) | grepl('SRSF2-P95[AHL]|U2AF1-Q',site)) %>%
  select(Gene,site,coefficient) %>%
  distinct() %>%
  mutate(site = paste(Gene,str_match(site,'[0-9A-Za-z]+$'),sep = '-'))

global_picture_parameters <- merge(average_age,most_recurring_coefficients_gene,
                                   by = c("Gene","site"),
                                   all = F) %>%
  arrange(Gene)

TOT_READS <- 1e3
global_picture_out <- list()
for (S in unique(global_picture_parameters$site)) {
  subset_df <- global_picture_parameters[global_picture_parameters$site == S,]
  u_one_cell <- logit(1/TOT_READS) - subset_df$MedianAge * subset_df$coefficient
  time_draws <- inv.logit(u_one_cell + seq(floor(subset_df$MedianAge),85) * subset_df$coefficient) / 2
  global_picture_out[[S]] <- data.frame(
    time = seq(floor(subset_df$MedianAge),85),
    draws = time_draws,
    site = S,
    Gene = subset_df$Gene
  )
}

text_global_picture_out <- global_picture_out %>% 
  do.call(what = rbind) %>%
  subset(Gene %in% c("DNMT3A","U2AF1","SRSF2","JAK2","IDH1","SF3B1")) %>%
  subset(!(site %in% c('SRSF2-P95A','U2AF1-Q157P'))) %>% 
  group_by(site) %>% 
  summarise(x = max(time),
            y = max(draws),
            label = site[1])

global_picture_out %>% 
  do.call(what = rbind) %>%
  subset(Gene %in% c("DNMT3A","U2AF1","SRSF2","JAK2","IDH1","SF3B1")) %>% 
  subset(!(site %in% c('SRSF2-P95A','U2AF1-Q157P'))) %>% 
  ggplot(aes(x = time,y = draws,colour = Gene,group = site)) + 
  geom_line(size = 2,
            alpha = 0.5) + 
  # geom_text(
  #   inherit.aes = F,
  #   size = 4,
  #   hjust = 0,
  #   fontface="bold",
  #   data = text_global_picture_out,
  #   aes(x = x,y = y,label = label)
  # ) +
  geom_label_repel(inherit.aes = F,
            size = 4,
            alpha = 0.5,
            data = text_global_picture_out,
            hjust = 0,
            vjust = 1.0,
            fontface = "bold",min.segment.length = 0,force = 0.001,
            direction = "y",
            label.r = unit(0,"cm"),
            label.size = unit(0,"cm"),
            aes(x = x,y = y,label = label,fill = str_match(label,'[0-9A-Z]+'))) +
  scale_y_continuous(trans = 'log10',breaks = c(#0,5e-6,1e-5,1e-4,
                                                0.001,0.010,0.10,0.20,0.3,0.5),
                     minor_breaks = NULL,
                     expand = c(0,0.1)) + 
  scale_x_continuous(limits = c(30,100),
                     breaks = c(40,60,80),
                     minor_breaks = NULL,
                     expand = c(0,0.05)) +
  scale_color_manual(guide=F,values = gene_colours) + 
  scale_fill_manual(guide=F,values = gene_colours) +
  theme_gerstung(base_size = 20) +
  theme(legend.position = 'bottom',
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.line = element_line()) +
  xlab("Age") + 
  ylab("logVAF") +
  ggsave(useDingbats=FALSE,sprintf("figures/%s/inferred_trajectories/GlobalPicture.pdf",model_id),
         height = 6,width = 7) +
  ggsave(sprintf("figures/%s/inferred_trajectories/GlobalPicture.svg",model_id),
         height = 6,width = 7)

```

## Statistical analysis 

```{r comparing_statistics,fig.height=5,fig.width=18}
statistics_data <- statistics_lists$model_f2$individual_site %>% 
  select(site,sum_stat,gene,individual) %>% 
  mutate(model = "") %>%
  mutate(recurrent = ifelse(site %in% values_model$training_subset$unique_site_multiple,TRUE,FALSE)) %>%
  group_by(model) %>%
  mutate(gene = str_match(gene,"[A-Z0-9]+"),
         truncating = str_match(gene,"[nt]+")) %>%
  mutate(model_label = sprintf("%s\np(s-value < 0.7) = %s",model,round(sum(sum_stat < 0.7)/length(sum_stat),3))) %>%
  mutate(gene_numeric = as.numeric(factor(gene,levels = rev(unique(gene))))) 

gene_numeric_correspondence <- statistics_data %>%
  ungroup() %>%
  select(gene,gene_numeric) %>%
  distinct()

individuals_with_no_interactions <- interference_idxs %>% 
  select(ind,inter_site) %>%
  distinct() %>%
  transmute(individual = val_subset$unique_individual_true[ind],
            has_interaction = !is.na(inter_site))
```

```{r statistics_model_d,fig.height=7,fig.width=7}
statistics_data %>%
  ungroup() %>%
  ggplot(aes(x = gene_numeric,y = sum_stat,colour = recurrent)) + 
  geom_hline(yintercept = 0.7,alpha = 0.5) + 
  geom_density2d(size = 0.5,
                 alpha = 0.2) +
  geom_point(position = position_jitter(width = 0.3),
             size = 1,
             alpha = 0.4) + 
  theme_gerstung(base_size = 15) +
  facet_grid(~ model_label) + 
  rotate_x_text() + 
  stat_summary(
    aes(x = gene_numeric),
    geom = 'label',
    color = 'black',
    label.r = unit(0, "lines"),
    label.size = 0,
    size = 4,
    alpha = 0.85,
    label.padding = unit(0.05,'cm'),
    hjust = 1,
    fun.data = function(x) {
      return(data.frame(y = 0.699,label = sprintf("%s",round(sum(x < 0.7)/length(x),2))))
    }) +
  scale_x_continuous(
    breaks = gene_numeric_correspondence$gene_numeric,
    labels = gene_numeric_correspondence$gene,
    sec.axis = dup_axis(name = NULL),
    expand = c(0.02,0.02)) +
  scale_y_continuous(
    breaks = c(0.00,0.25,0.5,0.7,0.75,1.00),
    expand = c(0.01,0.01)
  ) +
  scale_color_lancet(name = "Present in more\nthan 1 individiual") +
  theme(legend.position = "bottom",
        panel.background = element_rect(fill = NULL,colour = 'black'),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour = 'grey',size = 0.2)) + 
  coord_flip() + 
  xlab("Driver gene") + 
  ylab("s-value") + 
  ggsave(useDingbats=FALSE,sprintf("figures/%s/statistical_analysis/StatisticD2.pdf",model_id),height = 6.5,width = 6)

statistics_data %>%
  ungroup() %>%
  mutate(model_label = paste("Proportion of explained trajectories =",round(sum(sum_stat < 0.7)/length(sum_stat),3))) %>% 
  mutate(good = sum_stat < 0.7) %>% 
  mutate(recurrent = as.factor(ifelse(recurrent,"Yes","No"))) %>%
  ggplot(aes(x = gene_numeric,
             y = sum_stat)) + 
  geom_boxplot(aes(group = paste(gene)),
               #position = position_dodge(0.8),
               alpha=0.8) +
  stat_summary(
    aes(x = gene_numeric),
    geom = 'label',
    color = 'black',
    label.r = unit(0, "lines"),
    label.size = 0,
    angle = 90,
    size = 4,
    alpha = 0.85,
    label.padding = unit(0.05,'cm'),
    vjust = 0.5,
    hjust = 0,
    fun.data = function(x) {
      return(data.frame(y = 0.699,label = sprintf("%s",format(sum(x < 0.7)/length(x),digits=2,nsmall=2))))
    }) +
  geom_hline(yintercept = 0.7,alpha = 0.5,linetype=5) + 
  theme_gerstung(base_size = 15) +
  facet_grid(~ model_label) + 
  rotate_x_text() + 
  scale_x_continuous(
    breaks = gene_numeric_correspondence$gene_numeric,
    labels = gene_numeric_correspondence$gene,
    sec.axis = dup_axis(name = NULL),
    expand = c(0.02,0.02)) +
  scale_y_continuous(
    breaks = c(0.00,0.25,0.5,0.7,0.75,1.00),
    expand = c(0.01,0.01)
  ) +
  scale_color_manual(
    values = c("#00468BFF","#ED0000FF"),
    label = c("Yes","No"),
    name = "Present in more\nthan 1 individiual") +
  theme(legend.position = "bottom",
        panel.background = element_rect(fill = NULL,colour = 'black'),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid = element_blank()) + 
  xlab("Driver gene") + 
  ylab("Residual effect") + 
  coord_flip() + 
  ggsave(sprintf("figures/%s/statistical_analysis/StatisticD2Boxplot.svg",model_id),
         height = 6.5,width = 6) +
  ggsave(useDingbats=FALSE,sprintf("figures/%s/statistical_analysis/StatisticD2Boxplot.pdf",model_id),
         height = 6.5,width = 6) 

statistics_data_bars <- statistics_data %>%
  ungroup() %>%
  mutate(model_label = sprintf("Clonal trajectories\ncompatible with\nmodel = %s%%",100*round(sum(sum_stat < 0.7)/length(sum_stat),3))) %>% 
  mutate(good = sum_stat < 0.7,
         not_good = sum_stat > 0.7) %>% 
  gather(key = "key",value = "value",good,not_good) %>%
  mutate(key = factor(key,levels = c("not_good","good"))) %>%
  filter(value == T) %>%
  select(site,sum_stat,gene,individual,gene_numeric,key,value,model_label) %>%
  group_by(gene,model_label) %>%
  mutate(total = length(key)) %>%
  group_by(gene,key,model_label) %>%
  summarise(N = length(site),
            Total = total[1]) %>% 
  mutate(Proportion = N / Total,
         Proportion005 = ifelse(
           key == "good",
           qbeta(p = 0.05,N+1,Total+1-N),
           NA),
          Proportion095 = ifelse(
           key == "good",
           qbeta(p = 0.95,N+1,Total+1-N),
           NA)) %>% 
  mutate(
    Proportion095 = ifelse(Proportion095 < Proportion,
                           Proportion,
                           Proportion095),
    Proportion005 = ifelse(Proportion005 > Proportion,
                           Proportion,
                           Proportion005)
  ) 

gene_order_explained <- statistics_data_bars$gene[statistics_data_bars$key == 'good'][
  order(statistics_data_bars$Proportion[statistics_data_bars$key == 'good'])]

statistics_data_bars <- statistics_data_bars %>%
  mutate(
    gene = factor(gene,rev(gene_order_explained),ordered = T)
  )

statistics_data_bars %>% 
  ggplot(aes(x = gene,y = Proportion,fill = key)) + 
  geom_bar(stat = 'identity',color='black') +
  geom_linerange(aes(ymin = Proportion005-0.0005,ymax = Proportion095+0.0005),
                 colour = "black",size = 2) +
  geom_linerange(aes(ymin = Proportion,ymax = Proportion095),
                 colour = "black",
                 size = 1) + 
  geom_linerange(aes(ymin = Proportion,ymax = Proportion005),
                 colour = "grey80",
                 size = 1) + 
  theme_gerstung(base_size = 15) +
  facet_grid(~ model_label) + 
  #rotate_x_text() + 
  scale_x_discrete(
    expand = c(0.02,0.02)) +
  scale_y_continuous(
    breaks = c(0.00,0.5,1.00),
    expand = c(0,0)
  ) +
  scale_fill_manual(name = NULL,
                    values = c("black","grey80"),
                    breaks = c("good","not_good"),
                    labels = c("Good fit","Poor fit")) +
  theme(legend.position = "bottom",
        #panel.background = element_rect(fill = NULL,colour = 'black'),
        strip.text = element_text(size = 15),
        axis.text.y = element_text(face = 'italic'),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,0,0)) + 
  guides(fill=guide_legend(nrow=2,byrow=TRUE)) +
  xlab("Driver gene") + 
  ylab("Trajectories compatible\nwith model") + 
  coord_flip() +
  ggsave(sprintf("figures/%s/statistical_analysis/StatisticD2barplot.svg",model_id),
         height = 6,width = 3.2) +
  ggsave(useDingbats=FALSE,sprintf("figures/%s/statistical_analysis/StatisticD2barplot.pdf",model_id),
         height = 6,width = 3.2) 
```

```{r save_r_values_bad_statistics,fig.width=18,fig.height=10,echo=F}
r_values %>% 
  write.csv(sprintf("r_values_%s.csv",model_id))
r_values %>%
  merge(statistics_data,by = c("site","individual")) %>% 
  group_by(individual) %>%
  filter(any(sum_stat >= 0.7)) %>%
  mutate(bad = sum_stat >= 0.7) %>%
  ggplot(aes(x = age,
             y = pred / coverage,
             ymax = pred_095 / coverage,
             ymin = pred_005 / coverage,
             group = site,
             color = gene)) + 
  geom_line(aes(linetype = bad)) + 
  geom_line(aes(y = true/coverage),linetype = 3,
            alpha = 0.8) + 
  geom_point(aes(y = true/coverage),
             alpha = 0.8) + 
  geom_linerange() + 
  facet_wrap(~ individual,scales="free") + 
  theme_gerstung() + 
  scale_color_manual(values = gene_colours,name=NULL) + 
  theme(legend.position = "bottom") + 
  scale_linetype_manual(values = c(1,5)) + 
  scale_y_continuous(trans = 'log10') + 
  ggsave(useDingbats=FALSE,sprintf("figures/%s/inferred_trajectories/bad_trajectories.pdf",model_id),
         height = 10,width = 18)
```

```{r bad_statistics_examples,fig.width=8,fig.height=8}
data_bad_statistics <- r_values %>%
  merge(statistics_data,by = c("site","individual")) %>% 
  group_by(individual) %>%
  filter(any(sum_stat >= 0.7)) %>% 
  mutate(bad = sum_stat >= 0.7) %>%
  mutate(site_ = site) %>% 
  mutate(site = as.character(site)) %>% 
  mutate(site = sapply(site,function(x){
    tmp <- unlist(strsplit(x,'-'))
    paste(tmp[2:length(tmp)],collapse = '-')
  })) %>%
  mutate(site = sapply(site,function(x) unlist(strsplit(x,';'))[1])) %>%
  mutate(site = sapply(site,function(x) {
    tmp <- unlist(strsplit(x,':'))
    return(tmp[length(tmp)])})) %>%
  mutate(site = paste(gene,site,sep='-'))

competition_example <- data_bad_statistics %>% 
  subset(individual == 22109) %>%
  mutate(Label = "Competition")

sequencing_artefact_example <- data_bad_statistics %>%
  subset(individual == 1554) %>%
  mutate(Label = "Sequencing artefact")

size_dependent_deceleration_example <- data_bad_statistics %>%
  subset(individual == 11959) %>%
  mutate(Label = "Size-dependent deceleration")

sequencing_artefact_stochastic_growth_example <- data_bad_statistics %>%
  subset(individual == 29622) %>%
  mutate(Label = "Sequencing artefact/Stochastic growth regime")

plot_df <- list(
  competition_example,
  sequencing_artefact_example,
  size_dependent_deceleration_example,
  sequencing_artefact_stochastic_growth_example
)

p_list <- lapply(plot_df,function(x) {
  x %>% 
    ggplot(aes(x = age,group = site,colour = site)) + 
    geom_line(aes(y = pred / coverage,linetype = bad),
              position = position_dodge(width = 0.5)) + 
    geom_line(aes(y = true/coverage),linetype = 3,
              position = position_dodge(width = 0.5),
              alpha = 0.8) + 
    geom_point(aes(y = true/coverage),
               position = position_dodge(width = 0.5),
               alpha = 0.8,size=2) + 
    geom_linerange(aes(ymin = pred_005/coverage,ymax = pred_095/coverage),
                   position = position_dodge(width = 0.5)) + 
    scale_y_continuous(trans = 'log10',breaks = c(0.001,0.005,0.01,0.03,0.1,0.3,0.5)) + 
    theme_gerstung(base_size = 15) + 
    scale_colour_lancet(name = NULL) + 
    xlab("Age") + 
    ylab("VAF") + 
    scale_linetype_manual(values = c(5,1),
                          name=NULL,
                          breaks = c(T,F),
                          drop=F,
                          labels=c("Poor fit","Good fit"),
                          guide = F) + 
    theme(legend.position = "bottom") + 
    guides(colour = guide_legend(nrow = 2,byrow = T))
}) 

plot_grid(plotlist = p_list) + 
  ggsave(useDingbats=F,sprintf("figures/%s/inferred_trajectories/bad_trajectories_nice.pdf",model_id),
         height = 8,
         width = 8)
```

```{r saving_everything_exporting_as_script}
r_values <- r_values_list$model_f2

save(r_values,statistics_data,clone_ages,file = sprintf('vaf_modelling_coefficients_%s.Rdata',model_id))

knitr::purl("vaf_modelling_paper_1.Rmd")
```

# Session details

```{r session_details}
cat(system("git log -n1", intern=TRUE), sep="\n")
```

```{r session_objects}
l <- ls()
data.frame(variable=l, Reduce("rbind",lapply(l, function(x) data.frame(classes=paste(class(get(x)),collapse = ','), size=format(object.size(get(x)), units="auto")))))
```

```{r session_packages}
sessionInfo()
```